### 2022-05-18 diagram_base.tex
all at once

\documentclass[tikz,border=2mm]{standalone}

\usepackage{grffile}
\usetikzlibrary{arrows,positioning}

\begin{document}

\begin{tikzpicture}[node distance=3.5cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
    \node [int, ultra thick] (S)             {$S$};
    \node [int, right=of S, ultra thick] (E)             {$E$};
    \node [int, above right=of E, ultra thick] (Ia)             {$I_a$};
    \node [int, below right=of E, ultra thick] (Is)             {$I_s$};
    \node [int, right=of Ia, ultra thick] (R)             {$R$};
    \node [int, right=of Is, ultra thick] (D)             {$D$};
    \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
    \path[->, ultra thick] (E) edge node [sloped, above] {$p \mu$} (Ia);
    \path[->, ultra thick] (E) edge node [sloped, above] {$(1-p) \mu$} (Is);
    \path[->, ultra thick] (Ia) edge node [above] {$\gamma$} (R);
    \path[->, ultra thick] (Is) edge node [sloped, above] {$(1-f) \gamma$} (R);
    \path[->, ultra thick] (Is) edge node [above] {$f \gamma$} (D);
    \node [right=1.5cm of S, ultra thick] (Ts)             {};
    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta$} ([yshift=5mm]Ts.north);
    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta$} (Ts.south);
    \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};
\end{tikzpicture}
\end{document}

### 2022-05-18 diagram_immune.tex
all at once

--- 2022-05-18 diagram_base.tex
@@
     \node [int, right=of Is, ultra thick] (D)             {$D$};
-    \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$p \mu$} (Ia);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$(1-p) \mu$} (Is);
+    \path[->, ultra thick] (S) edge node [above] {${\color{red}(1-  \epsilon_i)} \lambda(t)$} (E);
+    \path[->, ultra thick] (E) edge node [sloped, above] {$\big[1-{\color{red}(1-\epsilon_s)} (1-p)\big] \mu$} (Ia);
+    \path[->, ultra thick] (E) edge node [sloped, above] {${\color{red}(1-\epsilon_s)} (1-p) \mu$} (Is);
     \path[->, ultra thick] (Ia) edge node [above] {$\gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [sloped, above] {$(1-f) \gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [above] {$f \gamma$} (D);
+    \path[->, ultra thick] (Is) edge node [sloped, above] {$\big[1-{\color{red}(1 - \epsilon_d)} f\big] \gamma$} (R);
+    \path[->, ultra thick] (Is) edge node [above] {${\color{red}(1 - \epsilon_d)} f \gamma$} (D);
     \node [right=1.5cm of S, ultra thick] (Ts)             {};
@@
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta$} (Ts.south);
-    \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};
+    \node [below=8.5cm of S, anchor=west] (pic) {\includegraphics{figure_immune.ggp.pdf}};
 \end{tikzpicture}

### 2022-05-18 figure.R

@@ -1,46 +1,0 @@
-library(ggplot2); theme_set(theme_bw(base_family="Times"))
-library(egg)
-library(shellpipes)
-
-startGraphics()
-
-loadEnvironments()
-
-pop <- 1e7
-
-g1 <- ggplot(allout_i_summarize) +
-  geom_line(aes(p, (D_p+D_n) * pop, col=factor(epsilon_i))) +
-  scale_x_continuous("Proportion asymptomatic") +
-  scale_y_continuous("Total deaths") +
-  scale_color_viridis_d("VE") +
-  ggtitle("A. Protection against infection") +
-  theme(
-    panel.grid = element_blank(),
-    legend.position = c(0.85, 0.7)
-  )
-
-g2 <- ggplot(allout_s_summarize) +
-  geom_line(aes(p, (D_p+D_n) * pop, col=factor(epsilon_s))) +
-  scale_x_continuous("Proportion asymptomatic") +
-  scale_y_continuous("Total deaths") +
-  scale_color_viridis_d("VE") +
-  ggtitle("B. Protection against symptoms") +
-  theme(
-    panel.grid = element_blank(),
-    legend.position = "none"
-  )
-
-g3 <- ggplot(allout_d_summarize) +
-  geom_line(aes(p, (D_p+D_n) * pop, col=factor(epsilon_d))) +
-  scale_x_continuous("Proportion asymptomatic") +
-  scale_y_continuous("Total deaths") +
-  scale_color_viridis_d("VE") +
-  ggtitle("C. Protection against deaths") +
-  theme(
-    panel.grid = element_blank(),
-    legend.position = "none"
-  )
-
-gfinal <- ggarrange(g1, g2, g3, nrow=1, draw=FALSE)
-
-saveGG(gfinal, width=12, height=4)

### 2022-05-18 figure_base.R

@@ -1,0 +1,48 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 16))
+library(egg)
+library(shellpipes)
+
+startGraphics()
+
+simulate_base <- rdsRead()
+
+simulate_base_max <- simulate_base %>%
+  group_by(sim) %>%
+  filter(
+    time==max(time)
+  )
+
+g1 <- ggplot(simulate_base_max) +
+  geom_line(aes(p, 1-S_n, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion infected",
+                     limits=c(0, 1),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = c(0.8, 0.45)
+  )
+
+g2 <- ggplot(simulate_base_max) +
+  geom_line(aes(p, D_n, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, 0.01),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
+
+gfinal <- ggarrange(g1, g2, nrow=1, draw=FALSE)
+
+saveGG(gfinal, width=8, height=3)

### 2022-05-18 figure_immune.R

@@ -1,0 +1,61 @@
+library(ggplot2); theme_set(theme_bw(base_family="Times"))
+library(egg)
+library(shellpipes)
+
+startGraphics()
+
+loadEnvironments()
+
+ymax <- 0.0052
+
+g1 <- ggplot(simulate_immune_i_summarize) +
+  geom_line(aes(p, (D_p+D_n), col=factor(epsilon_i))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, ymax),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d("Protection", option="B", end=0.9) +
+  ggtitle("A. Protection against\ninfection") +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
+
+g2 <- ggplot(simulate_immune_s_summarize) +
+  geom_line(aes(p, (D_p+D_n), col=factor(epsilon_s))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, ymax),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d("Protection", option="B", end=0.9) +
+  ggtitle("B. Protection against\nsymptoms") +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
+
+g3 <- ggplot(simulate_immune_d_summarize) +
+  geom_line(aes(p, (D_p+D_n), col=factor(epsilon_d))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, ymax),
+                     expand=c(0, 0)) +
[…]

### 2022-05-18 figure_variant.R

@@ -1,0 +1,138 @@
+library(ggplot2); theme_set(theme_bw(base_family="Times"))
+library(gridExtra)
+library(egg)
+library(shellpipes)
+
+startGraphics()
+
+loadEnvironments()
+
+ymax <- 0.0052
+ymax2 <- 0.19
+ymax3 <- 0.000155
+
+simulate_orig_profile_summarize <- simulate_orig_profile %>%
+  group_by(sim) %>%
+  filter(time==max(time)) %>%
+  mutate(
+    type="Resident"
+  )
+
+simulate_var_profile_summarize <- simulate_var_profile %>%
+  group_by(sim) %>%
+  filter(time==max(time)) %>%
+  mutate(
+    type="Invading"
+  )
+
+simulate_var_cross_profile_summarize <- simulate_var_cross_profile %>%
+  group_by(sim) %>%
+  filter(time==max(time)) %>%
+  mutate(
+    type="Invading"
+  )
+
+g1 <- ggplot(simulate_orig_profile_summarize, aes(p, (D_p+D_n), lty=type)) +
+  #annotate("segment", x=p_orig, y=tail(simulate_orig$D_p+simulate_orig$D_n, 1),
+  #         xend=p_var1, yend=tail(simulate_var1$D_p+simulate_var1$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1)-0.00015,
+  #         arrow = arrow(length=unit(0.1, "inches")), col="orange") +
+  #annotate("segment", x=p_orig, y=tail(simulate_orig$D_p+simulate_orig$D_n, 1),
+  #         xend=p_var2-0.03, yend=tail(simulate_var2$D_p+simulate_var2$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1),
+  #         arrow = arrow(length=unit(0.1, "inches")), col="purple") +
+  geom_line() +
+  geom_line(data=simulate_var_profile_summarize) +
+  geom_point(x=p_orig, y=tail(simulate_orig$D_p+simulate_orig$D_n, 1), size=3) +
+  geom_point(x=p_var1, y=tail(simulate_var1$D_p+simulate_var1$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1), col="orange", size=3, shape="triangle") +
[…]

### 2022-05-18 funs.R

@@ -3,7 +3,7 @@ model <- function(t, y, parms) {
     foi <- (beta * (Ia_n + Ia_p) + (1 - delta) * beta * (Is_n + Is_p))
     
     dS_n <- - foi * S_n ## naive
-    dS_p <- - (1-epsilon_i) * foi * S_n ## protected
+    dS_p <- - (1-epsilon_i) * foi * S_p ## protected
     
     dE_n <- foi * S_n - mu * E_n
     dIa_n <- p * mu * E_n - gamma * Ia_n
@@ -13,7 +13,7 @@ model <- function(t, y, parms) {
     
     dD_n <- f * gamma * Is_n
     
-    dE_p <- (1-epsilon_i) * foi * S_n - mu * E_p
+    dE_p <- (1-epsilon_i) * foi * S_p - mu * E_p
     dIa_p <- (1-(1-epsilon_s) * (1-p)) * mu * E_p - gamma * Ia_p
     dIs_p <- (1-epsilon_s) * (1-p) * mu * E_p - gamma * Is_p
     
@@ -21,24 +21,26 @@ model <- function(t, y, parms) {
     
     dD_p <- (1-epsilon_d) * f * gamma * Is_p
     
-    list(c(dS_n, dS_p, dE_n, dIa_n, dIs_n, dR_n, dD_n, dE_p, dIa_p, dIs_p, dR_p, dD_p))
+    list(c(dS_n, dS_p, dE_n, dIa_n, dIs_n, dR_n, dD_n, dE_p, dIa_p, dIs_p, dR_p, dD_p),
+         dD_n=dD_n, dD_p=dD_p)
   })
 }
 
 simulate <- function(N=1,
                      S0=0.5,
-                     beta=6/7,
-                     mu=2/7,
-                     gamma=1/7,
-                     delta=0.9,
+                     D0=0,
+                     beta=4/5,
+                     mu=1/2,
+                     gamma=1/5,
+                     delta=0.8,
                      f=0.01,
                      p=0.5,
                      epsilon_i=0,
                      epsilon_s=0,
                      epsilon_d=0,
-                     tvec=seq(0, 800, by=1),
-                     I0=1e-3) {
[…]

### 2022-05-18 param_variant.R

@@ -1,0 +1,7 @@
+p_orig <- p_var1 <- 0.2
+p_var2 <- 0.5
+delta <- 0.8
+epsilon_s <- 0.4
+epsilon_i <- 0.4
+tvec_orig <- seq(0, 365, by=1)
+tvec_var <- seq(365, 365*2, by=1)

### 2022-05-18 simulate_base.R

@@ -1,0 +1,20 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+p <- seq(0, 1, length.out=101)
+delta <- 0:5 * 0.2
+
+simparam <- expand.grid(p, delta)
+names(simparam) <- c("p", "delta")
+
+simulation_base <- apply(simparam, 1, function(x) {
+  out <- do.call(simulate, c(as.list(x), S=1, D0=0))
+  
+  as.data.frame(append(out, x))
+}) %>% bind_rows(.id="sim")
+
+rdsSave(simulation_base)

### 2022-05-18 simulate_immune.R

@@ -5,7 +5,9 @@ library(shellpipes)
 
 loadEnvironments()
 
-p <- seq(0, 1, length.out=21)
+delta <- 0.8
+
+p <- seq(0, 1, length.out=101)
 epsilon_i <- epsilon_s <- epsilon_d <- c(0, 0.2, 0.4, 0.6, 0.8)
 
 simparam <- expand.grid(p, epsilon_i, epsilon_s, epsilon_d)
@@ -16,11 +18,11 @@ simparam2 <- simparam %>%
     (epsilon_i==0) + (epsilon_s==0) + (epsilon_d==0) >= 2
   )
 
-allout <- apply(simparam2, 1, function(x) {
-  out <- do.call(simulate, as.list(x))
+simulate_immune <- apply(simparam2, 1, function(x) {
+  out <- do.call(simulate, c(as.list(x), delta=delta))
   
   as.data.frame(append(out, x))
 }) %>% bind_rows(.id="sim")
 
-rdsSave(allout)
+rdsSave(simulate_immune)
 

### 2022-05-18 simulate_variant.R

@@ -1,0 +1,46 @@
+library(deSolve)
+library(shellpipes)
+
+loadEnvironments()
+
+simulate_orig <- simulate(N=1,
+                          S0=1,
+                          delta=delta,
+                          p=p_orig,
+                          tvec=tvec_orig)
+
+simulate_var1 <- simulate(N=1,
+                          S0=tail(simulate_orig, 1)$S_n,
+                          D0=tail(simulate_orig, 1)$D_n,
+                          delta=delta,
+                          p=p_var1,
+                          tvec=tvec_var,
+                          epsilon_s=epsilon_s)
+
+simulate_var1_cross <- simulate(N=1,
+                          S0=tail(simulate_orig, 1)$S_n,
+                          D0=tail(simulate_orig, 1)$D_n,
+                          delta=delta,
+                          p=p_var1,
+                          tvec=tvec_var,
+                          epsilon_i=epsilon_i,
+                          epsilon_s=epsilon_s)
+
+simulate_var2 <- simulate(N=1,
+                          S0=tail(simulate_orig, 1)$S_n,
+                          D0=tail(simulate_orig, 1)$D_n,
+                          delta=delta,
+                          p=p_var2,
+                          tvec=tvec_var,
+                          epsilon_s=epsilon_s)
+
+simulate_var2_cross <- simulate(N=1,
+                                S0=tail(simulate_orig, 1)$S_n,
+                                D0=tail(simulate_orig, 1)$D_n,
+                                delta=delta,
+                                p=p_var2,
+                                tvec=tvec_var,
+                                epsilon_i=epsilon_i,
+                                epsilon_s=epsilon_s)
+
+saveEnvironment()

### 2022-05-18 simulate_variant_profile.R

@@ -1,0 +1,45 @@
+library(dplyr)
+library(deSolve)
+library(shellpipes)
+
+loadEnvironments()
+
+p <- seq(0, 1, length.out=101)
+
+simulate_orig_profile <- lapply(p, function(p) {
+  out <- simulate(S=1, D0=0, p=p)
+  
+  out$p <- p
+  
+  out
+}) %>% bind_rows(.id="sim")
+
+simulate_var_profile  <- lapply(p, function(p) {
+  out <- simulate(S0=tail(simulate_orig, 1)$S_n,
+                  D0=tail(simulate_orig, 1)$D_n,
+                  p=p,
+                  epsilon_s=epsilon_s)
+  
+  out$D_n <- out$D_n - tail(simulate_orig, 1)$D_n
+  
+  out$p <- p
+  
+  out
+}) %>% bind_rows(.id="sim")
+
+
+simulate_var_cross_profile  <- lapply(p, function(p) {
+  out <- simulate(S0=tail(simulate_orig, 1)$S_n,
+                  D0=tail(simulate_orig, 1)$D_n,
+                  p=p,
+                  epsilon_s=epsilon_s,
+                  epsilon_i=epsilon_i)
+  
+  out$D_n <- out$D_n - tail(simulate_orig, 1)$D_n
+  
+  out$p <- p
+  
+  out
+}) %>% bind_rows(.id="sim")
+
+saveEnvironment()

### 2022-05-18 summarize.R

@@ -1,30 +1,0 @@
-library(dplyr)
-
-library(shellpipes)
-
-allout <- rdsRead()
-
-outsumm <- . %>%
-  group_by(sim) %>%
-  filter(time==max(time)) %>%
-  ungroup
-
-allout_i <- allout %>%
-  filter(epsilon_s==0, epsilon_d==0)
-
-allout_s <- allout %>%
-  filter(epsilon_i==0, epsilon_d==0)
-
-allout_d <- allout %>%
-  filter(epsilon_i==0, epsilon_s==0)
-
-allout_i_summarize <- allout_i %>%
-  outsumm
-
-allout_s_summarize <- allout_s %>%
-  outsumm
-
-allout_d_summarize <- allout_d %>%
-  outsumm
-
-saveEnvironment()

### 2022-05-18 summarize_immune.R

@@ -1,0 +1,30 @@
+library(dplyr)
+
+library(shellpipes)
+
+simulate_immune <- rdsRead()
+
+outsumm <- . %>%
+  group_by(sim) %>%
+  filter(time==max(time)) %>%
+  ungroup
+
+simulate_immune_i <- simulate_immune %>%
+  filter(epsilon_s==0, epsilon_d==0)
+
+simulate_immune_s <- simulate_immune %>%
+  filter(epsilon_i==0, epsilon_d==0)
+
+simulate_immune_d <- simulate_immune %>%
+  filter(epsilon_i==0, epsilon_s==0)
+
+simulate_immune_i_summarize <- simulate_immune_i %>%
+  outsumm
+
+simulate_immune_s_summarize <- simulate_immune_s %>%
+  outsumm
+
+simulate_immune_d_summarize <- simulate_immune_d %>%
+  outsumm
+
+saveEnvironment()

### 2022-05-21 diagram_base.tex
a bit more

@@
     \path[->, ultra thick] (E) edge node [sloped, above] {$(1-p) \mu$} (Is);
-    \path[->, ultra thick] (Ia) edge node [above] {$\gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [sloped, above] {$(1-f) \gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [above] {$f \gamma$} (D);
+    \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);
+    \path[->, ultra thick] (Is) edge node [sloped, above] {$(1-f) \gamma_s$} (R);
+    \path[->, ultra thick] (Is) edge node [above] {$f \gamma_s$} (D);
     \node [right=1.5cm of S, ultra thick] (Ts)             {};
-    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta$} ([yshift=5mm]Ts.north);
-    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta$} (Ts.south);
+    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta_a$} ([yshift=5mm]Ts.north);
+    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
     \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};

### 2022-05-21 diagram_immune.tex
a bit more

@@
 \begin{tikzpicture}[node distance=3.5cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
-    \node [int, ultra thick] (S)             {$S$};
-    \node [int, right=of S, ultra thick] (E)             {$E$};
-    \node [int, above right=of E, ultra thick] (Ia)             {$I_a$};
-    \node [int, below right=of E, ultra thick] (Is)             {$I_s$};
-    \node [int, right=of Ia, ultra thick] (R)             {$R$};
-    \node [int, right=of Is, ultra thick] (D)             {$D$};
+    \node [int, ultra thick] (S)             {$S_p$};
+    \node [int, right=of S, ultra thick] (E)             {$E_p$};
+    \node [int, above right=of E, ultra thick] (Ia)             {$I_{p,a}$};
+    \node [int, below right=of E, ultra thick] (Is)             {$I_{p,s}$};
+    \node [int, right=of Ia, ultra thick] (R)             {$R_p$};
+    \node [int, right=of Is, ultra thick] (D)             {$D_p$};
     \path[->, ultra thick] (S) edge node [above] {${\color{red}(1-  \epsilon_i)} \lambda(t)$} (E);
@@
     \path[->, ultra thick] (E) edge node [sloped, above] {${\color{red}(1-\epsilon_s)} (1-p) \mu$} (Is);
-    \path[->, ultra thick] (Ia) edge node [above] {$\gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [sloped, above] {$\big[1-{\color{red}(1 - \epsilon_d)} f\big] \gamma$} (R);
-    \path[->, ultra thick] (Is) edge node [above] {${\color{red}(1 - \epsilon_d)} f \gamma$} (D);
+    \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);
+    \path[->, ultra thick] (Is) edge node [sloped, above] {$\big[1-{\color{red}(1 - \epsilon_d)} f\big] \gamma_s$} (R);
+    \path[->, ultra thick] (Is) edge node [above] {${\color{red}(1 - \epsilon_d)} f \gamma_s$} (D);
     \node [right=1.5cm of S, ultra thick] (Ts)             {};
-    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta$} ([yshift=5mm]Ts.north);
-    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta$} (Ts.south);
+    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta_a$} ([yshift=5mm]Ts.north);
+    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
     \node [below=8.5cm of S, anchor=west] (pic) {\includegraphics{figure_immune.ggp.pdf}};

### 2022-05-21 main.tex
a bit more

\documentclass[12pt]{article}
% \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
\usepackage[top=1in,footskip=1in]{geometry}

\usepackage{graphicx}
\usepackage{xspace}
%\usepackage{adjustbox}

\usepackage{grffile}

\newcommand{\comment}{\showcomment}
%% \newcommand{\comment}{\nocomment}

\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\bmb}[1]{\comment{blue}{BMB}{#1}}
\newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc} % make sure fancy dashes etc. don't get dropped

\usepackage{lineno}
\linenumbers

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{unsrt}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rres}{\Rx{\mathrm{res}}}
\newcommand{\Rinv}{\Rx{\mathrm{inv}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rt}{\ensuremath{{\mathcal R}(t)}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\rres}{\rx{\mathrm{res}}}
\newcommand{\rinv}{\rx{\mathrm{inv}}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation-period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\usepackage{lettrine}

\newcommand{\dropcapfont}{\fontfamily{lmss}\bfseries\fontsize{26pt}{28pt}\selectfont}
\newcommand{\dropcap}[1]{\lettrine[lines=2,lraise=0.05,findent=0.1em, nindent=0em]{{\dropcapfont{#1}}}{}}

\begin{document}

\begin{flushleft}{
	\Large
	\textbf\newline{
		Intermediate levels of symptomaticity can lead to the worst population-level outcomes
	}
}
\newline
\\
Joshua Weitz\textsuperscript{1,2,3},
Sang Woo Park\textsuperscript{4},
Jonathan Dushoff\textsuperscript{5,6,7}
\\
\bigskip
\textbf{1} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{2} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{3} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
\\
\textbf{4} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{5} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{7} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\bigskip

\bigskip
\end{flushleft}

\section{Introduction}

SARS-CoV-2 has had devastating effects at the population level even as individuals have experienced highly disparate infection outcomes.
Notably, many individuals have mild or even asymptomatic cases making it harder to estimate the magnitude of spread and fatality rate. 
The documented case-fatality rate (CFR) at the population level varies with population, testing infrastructure, and case definitions, typically between 1\%--4\% \citep{rajgor2020many,VERITY2020669,yang2020early}.
But, many infections are never documented. 
The infection fatality rate (IFR), which takes undocumented infections into account, has been estimated to be closer to 0.5\%--1\% for demographics similar to that of the United States \citep{levin2020assessing}. 
This means that more than 99\% of individuals infected with COVID-19 will survive. 
Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical and even asymptomatic. 
Yet, asymptomatically infected individuals can still transmit to others. 
Hence, asymptomatic infections can be a double-edged sword. 
On the one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
On the other hand, asymptomatic infections are less likely to be detected, and are may therefore be more likely to infect others, since precautions are less likely to be taken.
This means that the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
\swp{I feel like ``severe'' is not the right word here. Severe could mean higher f or lower p. But we strictly mean lower p here. Maybe: ``more likely to result in symptomatic infection at the individual level.''}.
\jd{This seems OK as long as we disambiguate later. Symptomaticity is prerequisite to any other kind of severity, and it's the route that we're talking about. I did change the word in a way that may help a little.}

To explore this idea, we propose a simple epidemic model,
in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively.  
Asymptomatically infected individuals always recover, whereas a fraction $f$ of symptomatically infected individuals die.
Our key assumption is symptomatically infected individuals  take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$. 
The full model equations are shown in Section.~\ref{sec.methods}. We start by making the parsimonious assumption that 
transmission rates and recovery rates are equivalent for the two routes. 
We assume that asymptomatic individuals do not die, and evaluate the effects of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).

\jd{How long do you run the ODE? Probably it's better to use the final-size formula.}

Figure~\ref{fig.reduce} evaluates epidemic outcomes using epidemiological parameters similar to that of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their effective transmission rate by $\delta$. 
When asymptomatic proportion is very high, fatalities are very low, as would be expected.
When asymptomatic proportion is low, however, the disease spreads poorly or not at all.
Thus, the peak fatality level is at intermediate levels of asymptomatic spread.

\jd{I added some simpler explanations above. I'm not sure why we need any of the stuff in the rest of the \P\ (now below).}
At high values of $p$, then the disease spreads unabated, reaching a herd immunity level at $1-\frac{1}{{\cal{R}}_0}$ before leading to epidemic overshoot and termination of the epidemic due to susceptible
depletion. 
Hence, differences in $\delta$ are immaterial: nearly everyone gets infected, but almost no one dies because nearly every case is asymptomatic.  
In contrast, the dynamics of the system are markedly different as $p\rightarrow 0$, i.e., all individuals are symptomatic and there are very few asymptomatically infected individuals. 
In this limit, then the total number of cases decreases, potentially to 0 (in a deterministic limit, see bottom panel). 
Although more individuals die per infection (\swp{meaning that IFR becomes higher}), the decrease in total infections also leads to a decrease in total fatalities. 
As a result, the severity of the impact as evaluated at the population level peaks at intermediate levels of asympmtomatic infection at the individual scale. Notably, the model also predicts that an outbreak will not initiate
when asymptomatic infections are rare and symptomatic reduction in transmission is high.
\jd{It would be good to explain clearly (given the nature of the paper) that the only assumption is that we have is that $\RR_a = \delta R_i$.}
The finding that fatalities peak at intermediate levels of asymptomatic infection $p$ can be
anticipated through analysis of the epidemic takeoff thresholds and the herd immunity threshold. The epidemic strength in this model is:
\begin{equation}
    {\cal{R}}_p = {\cal{R}}_0\left(p+(1-p)(1-\delta)\right)
\end{equation}
where ${\cal{R}}_0=\beta/\gamma$ is assumed to be equivalent for both the asymptomatic and symptomatic transmission routes. Keeping ${\cal{R}}_0$ fixed, then there is a critical level of asymptomatic transmission, $p_c$
\begin{equation}
    p_c = \frac{1}{{\cal{R}}_0}\delta+\frac{1-\delta}{\delta}
\end{equation}
such that $p>p_c$ is required for an outbreak. In the limit that $\delta=1$, then $p_c=1/{\cal{R}}_0$ (here equivalent
to $p_c=1/4$.  The critical outbreak requirement decreases with increasing $\delta$ such that an outbreak will happen
for all values of $p$ as long as $\delta<1-1/{\cal{R}}_0$ -- notice the difference between the cases 
of $\delta=0.7$ and $\delta=0.8$ which bridge this critical point.  Next, the total number of fatalities should be
$f$ multiplied by the fraction of symptomatic cases.  At the herd immunity threshold there should already
have been $f(1-p)\left(1-1/{\cal{R}}_0\right)$ fatalities.  Although cases go up with $p$ the likelihood of a fatal outcome goes down with $p$. The maximum can be found by finding the value $\hat{p}$ associated with the herd immunity threshold. In the case of $\delta=1$ (symptomatic individuals can have fatal outcomes but are sinks for transmission), then $\hat{p}=\frac{1}{{\cal{R}}_0^{1/2}}$, i.e., $\hat{p}=0.5$ for ${\cal{R}}=4$ as in this example.  The same
principle holds for sufficiently high values of $\delta$.

We then extend the model to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
The dynamics of immunologically naive individuals are equivalent to our original model.
The dynamics of protected individuals include three additional parameters, which characterize the degree of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously; we do not consider the effect of immunity on transmission.

The impact of protection against infection $\epsilon_i$ is analogous to changing ${\cal{R}}_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immunity}A).
We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between ${\cal{R}}_0$ and the final size.
The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immunity}B).
Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can ready transmit infections to other individuals. 
Finally, protection against deaths $\epsilon_d$ directly modulates the CFR and therefore linearly scales the fatality curves (\fref{immunity}C).

In summary, using a simplified model we have shown that asymptomatic infections can represent
a double-edge sword insofar as the represent a better outcome for some individuals but a mechanism for onwards transmission that leads to a worse outcome for the population as a whole. SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms.  Although mitigation efforts have often prioritized responding to symptoms -- including symptom-based testing, fever checks, mask-wearing for infectious individuals -- a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and in the event that individuals are at risk for severe outcomes.

\section*{Methods}
\label{sec.methods}
The model dynamics are as follows
\begin{eqnarray}
\dot{S}=\beta_a S I_a -(1-\delta) \beta_s S I_s \\
\dot{E} = \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
\dot{I}_a = p \mu E - \gamma_a I_a\\
\dot{I}_s = (1-p) \mu E -\gamma_s I_s\\
\dot{R} = \gamma_a I_a + (1-f) \gamma_s I_s \\
\dot{D} = f \gamma_s I_s
\end{eqnarray}
where the transmission rate $\beta$ and recover rate $\gamma$ can be potentially differ
between asymptomatic and symptomatically infected individuals.  Here, $\delta$ denotes the reduction in transmissability due to responsive measures taken by symptomatically infected indivivduals. Note that the SEIR model framework does not impact final size outcomes, so for convenience, all final size simulations ignore the $E$ class.

\bibliography{main}

\end{document}

### 2022-05-23 main.tex
commit before meeting

@@
 
-\section{Introduction}
-
 SARS-CoV-2 has had devastating effects at the population level even as individuals have experienced highly disparate infection outcomes.
@@
 On the one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
-On the other hand, asymptomatic infections are less likely to be detected, and are may therefore be more likely to infect others, since precautions are less likely to be taken.
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, and are may therefore be more likely to infect others, since precautions are less likely to be taken.
 This means that the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
-\swp{I feel like ``severe'' is not the right word here.
-Severe could mean higher f or lower p.
-But we strictly mean lower p here.
-Maybe: ``more likely to result in symptomatic infection at the individual level.''}.
-\jd{This seems OK as long as we disambiguate later.
-Symptomaticity is prerequisite to any other kind of severity, and it's the route that we're talking about.
-I did change the word in a way that may help a little.}
 
 To explore this idea, we propose a simple epidemic model,
-in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively.  
+in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively (\fref{base}).  
 Asymptomatically infected individuals always recover, whereas a fraction $f$ of symptomatically infected individuals die.
-Our key assumption is symptomatically infected individuals  take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$. 
-The full model equations are shown in Section.~\ref{sec.methods}. We start by making the parsimonious assumption that 
-transmission rates and recovery rates are equivalent for the two routes. 
+Asymptomatically and symptomatically infected individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
+Our key assumption is symptomatically infected individuals take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
+we note that the parameter $\delta$ may also capture intervention measures that target symptomatically infected individuals, such as symptom-based isolation. 
+As our nominal parameters, we assume lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
 We assume that asymptomatic individuals do not die, and evaluate the effects of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
 
-\jd{How long do you run the ODE? Probably it's better to use the final-size formula.}
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{diagram_base.pdf}
+\caption{
+\textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
+(Left) $S$ represents suceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+See Methods for model details.
+(Middle) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+(Right) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, and $f=0.01$.
+We assume that $10^{-4}$ proportion of individuals are initially infected.
+}
+\label{fig:base}
+\end{figure}
 
-Figure~\ref{fig.reduce} evaluates epidemic outcomes using epidemiological parameters similar to that of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their effective transmission rate by $\delta$.
-When asymptomatic proportion is very high, fatalities are very low, as would be expected.
-When asymptomatic proportion is low, however, the disease spreads poorly or not at all.
-Thus, the peak fatality level is at intermediate levels of asymptomatic spread.
+\fref{base} evaluates epidemic outcomes using epidemiological parameters similar to that of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
+In the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproductin number:
+\begin{equation}
+\RR_0 = (1-p) (1-\delta) \RR_s + p \RR_a,
+\end{equation}
+where $\RR_s = \beta_s/\gamma_s$ and $\RR_a = \beta_a/\gamma_a$ represent the reproduction numbers of asymptomatic and symptomatic individuals.
+This relationship changes as $\delta$ increases.
+In particular, when $\delta > 1-\RR_a/\RR_s$ (in our case, $\delta > 0.25)$, the basic reproduction number increases with $p$ because the behavioral effect causes the effective symptomatic transmission rate to be lower than that that of asymptomatic individuals;
+therefore, the number of infections increases with $p$.
+For even higher values of $\delta$, we find that there is a critical level of asymptomatic propotion, $p_c$:
+\begin{equation}
+    p_c = \frac{1 - (1-\delta) \RR_s}{\RR_a - (1-\delta) \RR_s}
+\end{equation}
+such that $p>p_c$ is required for an outbreak. 
+In the limit that $\delta=1$ (i.e., when symptomatically individuals does not transmit at all), then $p_c=1/\RR_a$.
 
-\jd{I added some simpler explanations above. I'm not sure why we need any of the stuff in the rest of the \P\ (now below).}
-At high values of $p$, then the disease spreads unabated, reaching a herd immunity level at $1-\frac{1}{{\cal{R}}_0}$ before leading to epidemic overshoot and termination of the epidemic due to susceptible
-depletion. 
-Hence, differences in $\delta$ are immaterial: nearly everyone gets infected, but almost no one dies because nearly every case is asymptomatic.  
-In contrast, the dynamics of the system are markedly different as $p\rightarrow 0$, i.e., all individuals are symptomatic and there are very few asymptomatically infected individuals. 
-In this limit, then the total number of cases decreases, potentially to 0 (in a deterministic limit, see bottom panel). 
-Although more individuals die per infection (\swp{meaning that IFR becomes higher}), the decrease in total infections also leads to a decrease in total fatalities. 
-As a result, the severity of the impact as evaluated at the population level peaks at intermediate levels of asympmtomatic infection at the individual scale.
-Notably, the model also predicts that an outbreak will not initiate
-when asymptomatic infections are rare and symptomatic reduction in transmission is high.
-\jd{It would be good to explain clearly (given the nature of the paper) that the only assumption is that we have is that $\RR_a = \delta R_i$.}
-The finding that fatalities peak at intermediate levels of asymptomatic infection $p$ can be
-anticipated through analysis of the epidemic takeoff thresholds and the herd immunity threshold. The epidemic strength in this model is:
-\begin{equation}
-    {\cal{R}}_p = {\cal{R}}_0\left(p+(1-p)(1-\delta)\right)
-\end{equation}
-where ${\cal{R}}_0=\beta/\gamma$ is assumed to be equivalent for both the asymptomatic and symptomatic transmission routes.
-Keeping ${\cal{R}}_0$ fixed, then there is a critical level of asymptomatic transmission, $p_c$
-\begin{equation}
-    p_c = \frac{1}{{\cal{R}}_0}\delta+\frac{1-\delta}{\delta}
-\end{equation}
-such that $p>p_c$ is required for an outbreak. In the limit that $\delta=1$, then $p_c=1/{\cal{R}}_0$ (here equivalent
-to $p_c=1/4$.  The critical outbreak requirement decreases with increasing $\delta$ such that an outbreak will happen
-for all values of $p$ as long as $\delta<1-1/{\cal{R}}_0$ -- notice the difference between the cases 
-of $\delta=0.7$ and $\delta=0.8$ which bridge this critical point.  Next, the total number of fatalities should be
-$f$ multiplied by the fraction of symptomatic cases.  At the herd immunity threshold there should already
-have been $f(1-p)\left(1-1/{\cal{R}}_0\right)$ fatalities.
-Although cases go up with $p$ the likelihood of a fatal outcome goes down with $p$.
-The maximum can be found by finding the value $\hat{p}$ associated with the herd immunity threshold.
-In the case of $\delta=1$ (symptomatic individuals can have fatal outcomes but are sinks for transmission), then $\hat{p}=\frac{1}{{\cal{R}}_0^{1/2}}$, i.e., $\hat{p}=0.5$ for ${\cal{R}}=4$ as in this example.
-The same
-principle holds for sufficiently high values of $\delta$.
+More intricate patterns emerge for fatalities.
+When $\delta = 0$, total fataities decrease with $p$ because IFR ($(1-p)f$) decreases.
+In contrast, for high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
+although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
 
 We then extend the model to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
-The dynamics of immunologically naive individuals are equivalent to our original model.
-The dynamics of protected individuals include three additional parameters, which characterize the degree of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
+The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
+The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$ (\fref{immune}).
 For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously; we do not consider the effect of immunity on transmission.
+We assume a relatively strong behaviroal effect $\delta=0.8$ for illustration.
 
-The impact of protection against infection $\epsilon_i$ is analogous to changing ${\cal{R}}_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immunity}A).
-We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between ${\cal{R}}_0$ and the final size.
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{diagram_immune.pdf}
+\caption{
+\textbf{Schematic diagram and simulations of a model with symptom-responsive transmission reduction and immunity.}
+(Top) The subscript $p$ represents protected individuals. 
+Immunity may provide protection against infection, symptoms, or deaths.
+The dynamics of immunologically naive individuals are described in (\fref{base}).
+(Bottom) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
+We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
+We assume that $10^{-4}$ proportion of individuals are initially infected.
+}
+\label{fig:immune}
+\end{figure}
+
+The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immunity}A).
+We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
 The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
 the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immunity}B).
-Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can ready transmit infections to other individuals.
-Finally, protection against deaths $\epsilon_d$ directly modulates the CFR and therefore linearly scales the fatality curves (\fref{immunity}C).
+Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
+This also means that the critical level of asymptomatic propotion decreases, allowing more dangeous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immunity}C).
 
-In summary, using a simplified model we have shown that asymptomatic infections can represent
-a double-edge sword insofar as the represent a better outcome for some individuals but a mechanism for onwards transmission that leads to a worse outcome for the population as a whole.
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figure_variant.ggp.pdf}
+\caption{
+\textbf{Dynamics of invading variants under symptom-responsive transmission reduction and immunity.}
+(Top) The subscript $p$ represents protected individuals. 
+Immunity may provide protection against infection, symptoms, or deaths.
+The dynamics of immunologically naive individuals are described in (\fref{base}).
+(Bottom) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
+We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$ for illustration.
+We assume that $10^{-4}$ proportion of individuals are initially infected.
+}
+\label{fig:variant}
+\end{figure}
+
+Finally, we use our framework to understand the impact of behavioral effects on invading variants (\fref{variant}).
+In doing so, we first simulate the dynamics of a resident variant for 1 year using our base model (\fref{base}).
+We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the resident variant.
+We consider two types of variants (which are simulated separately): one with the same severity $p$ (orange) and a milder one with higher $p$ (purple).
+
+First, we consider a scenario in which immunity only provides protection against symptoms, $\epsilon_s = 0.4$ (\fref{variant}A--C).
+In this case, protection against symptoms allows new variants to spread faster, resulting in larger outbreaks (\fref{variant}B).
+Although, a milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
+The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, resident variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
+In general, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms, given that asymptomatically infected individuals can transmit better than symptomatically infected individuals (due to either behavioral and intervention effects).
+
+Next, we consider a more realistic scenario in which immunity provides protection against symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
+In this case, cross protection against infection has a disproportionately larger effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, resident variant.
+Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level.
+
+In summary, using a simplified model we have shown that asymptomatic infections can represent a double-edge sword insofar as they represent a better outcome for some individuals but a mechanism for onwards transmission that leads to a worse outcome for the population as a whole.
+Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variants:
+while protection against symptoms prevent diseases at the individual level, it can inadvertently lead to more infections, and potentially deaths, at the population level.
+For example, milder variants can result in even faster epidemic growth rates and larger outbreaks, even though it can lead to less fatalities if the variant is sufficiently mild.
+
+Our simulatons of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
+Despite moderate levels of vaccine effectivenss against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity has provided limited protection against infections \citep{pearson2021omicron}.
+For example, its immune evasive property allowed the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
+therefore, even though the Omicron variant is thought to be milder \citep{MENNI20221618}, the number of deaths caused by the Omicron variant has been comparable to
+
+We hypothesize that the behavioral effect also played a role in allowing the Omicron to spread better: milder infections caused by the Omicron variant likely
+
+Even though the Omicron variant has been reported to be milder than previous variants  it has
+
+Despite high levels of vaccination
+
+There are several limitations to our analysis.
+First of all, behavioral and intervention effects must
+
 SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms. 
-Although mitigation efforts have often prioritized responding to symptoms -- including symptom-based testing, fever checks, mask-wearing for infectious individuals -- a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and in the event that individuals are at risk for severe outcomes.
+Although mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and in the event that individuals are at risk for severe outcomes.
+As more variants continue to emerge, updating vaccines to prevent infections, and not just diseases, is
 

### 2022-05-23 figure_base.R

@@ -1,9 +1,7 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 16))
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
 library(egg)
 library(shellpipes)
-
-startGraphics()
 
 simulate_base <- rdsRead()
 
@@ -25,7 +23,8 @@ g1 <- ggplot(simulate_base_max) +
   scale_color_viridis_d(expression(delta)) +
   theme(
     panel.grid = element_blank(),
-    legend.position = c(0.8, 0.45)
+    legend.position = c(0.8, 0.4),
+    legend.background = element_rect(fill=NA)
   )
 
 g2 <- ggplot(simulate_base_max) +

### 2022-05-23 figure_immune.R

@@ -2,8 +2,6 @@ library(ggplot2); theme_set(theme_bw(base_family="Times"))
 library(egg)
 library(shellpipes)
 
-startGraphics()
-
 loadEnvironments()
 
 ymax <- 0.0052
@@ -18,7 +16,7 @@ g1 <- ggplot(simulate_immune_i_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d("Protection", option="B", end=0.9) +
-  ggtitle("A. Protection against\ninfection") +
+  ggtitle("Protection against\ninfection") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -34,7 +32,7 @@ g2 <- ggplot(simulate_immune_s_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d("Protection", option="B", end=0.9) +
-  ggtitle("B. Protection against\nsymptoms") +
+  ggtitle("Protection against\nsymptoms") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -50,7 +48,7 @@ g3 <- ggplot(simulate_immune_d_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(epsilon), option="B", end=0.9) +
-  ggtitle("C. Protection against\ndeaths") +
+  ggtitle("Protection against\ndeaths") +
   theme(
     panel.grid = element_blank(),
     legend.position = "right"

### 2022-05-23 figure_variant.R

@@ -1,15 +1,14 @@
+library(dplyr)
 library(ggplot2); theme_set(theme_bw(base_family="Times"))
 library(gridExtra)
 library(egg)
 library(shellpipes)
 
-startGraphics()
-
 loadEnvironments()
 
 ymax <- 0.0052
 ymax2 <- 0.19
-ymax3 <- 0.000155
+ymax3 <- 0.00007
 
 simulate_orig_profile_summarize <- simulate_orig_profile %>%
   group_by(sim) %>%
@@ -133,6 +132,6 @@ g6 <- ggplot(simulate_orig, aes(time, dD_n + dD_p)) +
 
 gcomb2 <- ggarrange(g5, g6, nrow=2, draw=FALSE)
 
-gfinal <- arrangeGrob(g1, gcomb1, g4, gcomb2, widths=c(1, 2))
+gfinal <- arrangeGrob(g1, gcomb1, g4, gcomb2, widths=c(1, 1))
 
-saveGG(gfinal, target="figure_variant.Rout", width=10, height=6)
+saveGG(gfinal, target="figure_variant.Rout", width=8, height=6)

### 2022-05-23 funs.R

@@ -1,6 +1,6 @@
 model <- function(t, y, parms) {
   with(as.list(c(parms, y)), {
-    foi <- (beta * (Ia_n + Ia_p) + (1 - delta) * beta * (Is_n + Is_p))
+    foi <- (beta_a * (Ia_n + Ia_p) + (1 - delta) * beta_s * (Is_n + Is_p))
     
     dS_n <- - foi * S_n ## naive
     dS_p <- - (1-epsilon_i) * foi * S_p ## protected
@@ -29,7 +29,8 @@ model <- function(t, y, parms) {
 simulate <- function(N=1,
                      S0=0.5,
                      D0=0,
-                     beta=4/5,
+                     beta_a=4/5 * 0.75,
+                     beta_s=4/5,
                      mu=1/2,
                      gamma=1/5,
                      delta=0.8,
@@ -42,7 +43,7 @@ simulate <- function(N=1,
                      I0=1e-4) {
   y0 <- c(S_n=S0-I0, S_p=1-S0-D0, E_n=0, Ia_n=p * I0, Is_n=(1-p) * I0,  R_n=0, D_n=D0, E_p=0, Ia_p=0, Is_p=0, R_p=0, D_p=0)
   
-  parms <- c(beta=beta, mu=mu, gamma=gamma, delta=delta, f=f, p=p, epsilon_i=epsilon_i, epsilon_s=epsilon_s, epsilon_d=epsilon_d)
+  parms <- c(beta_a=beta_a, beta_s=beta_s, mu=mu, gamma=gamma, delta=delta, f=f, p=p, epsilon_i=epsilon_i, epsilon_s=epsilon_s, epsilon_d=epsilon_d)
   
   out <- as.data.frame(ode(y0, tvec, model, parms))
   

### 2022-05-25 diagram_pre.tex
working

--- 2022-05-21 diagram_base.tex
@@
     \node [int, right=of S, ultra thick] (E)             {$E$};
-    \node [int, above right=of E, ultra thick] (Ia)             {$I_a$};
-    \node [int, below right=of E, ultra thick] (Is)             {$I_s$};
-    \node [int, right=of Ia, ultra thick] (R)             {$R$};
-    \node [int, right=of Is, ultra thick] (D)             {$D$};
+    \node [int, right=of E, ultra thick] (Ip)             {$I_p$};
+    \node [int, right=of Ip, ultra thick] (Is)             {$I_s$};
+    \node [int, above right=of Is, ultra thick] (R)             {$R$};
+    \node [int, below right=of Is, ultra thick] (D)             {$D$};
     \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$p \mu$} (Ia);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$(1-p) \mu$} (Is);
-    \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);
+    \path[->, ultra thick] (E) edge node [sloped, above] {$\mu$} (Ip);
+    \path[->, ultra thick] (Ip) edge node [sloped, above] {$\sigma$} (Is);
     \path[->, ultra thick] (Is) edge node [sloped, above] {$(1-f) \gamma_s$} (R);
-    \path[->, ultra thick] (Is) edge node [above] {$f \gamma_s$} (D);
+    \path[->, ultra thick] (Is) edge node [sloped, above] {$f \gamma_s$} (D);
     \node [right=1.5cm of S, ultra thick] (Ts)             {};
-    \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta_a$} ([yshift=5mm]Ts.north);
-    \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};
+    \path[->, ultra thick, dashed] (Ip.north) edge [out=90, in=90] node [sloped, above] {$\beta_p$} ([yshift=5mm]Ts.north);
+    \path[->, ultra thick, dashed] (Is.south) edge [out=-90, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
+    \node [below=12cm of S, anchor=west] (pic) {\includegraphics{figure_pre.ggp.pdf}};
 \end{tikzpicture}

### 2022-05-25 main.tex
working

@@
 
+High values of $\delta$ may seem unrealistic---in practice, $\delta$ cannot be greater than the amouont of post-symptomatic transmission.
+While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to
+
+mitigation steps took place and before awareness of the
+epidemic became widespread
+
+\citep{he2020temporal}.
+
+\cite{sender2021unmitigated} recently showed the proportion of pre-symptomatic transmission is likely lower than previously estimated
+
+can be as low as 20\% (95\%CI: 6\%--32\%) in the absence of intervention measures, providing indirect support for the plausibility of high $\delta$ values.
+In Supplementary Figure S1, we therefore consider a more realistic model which includes a presymptomatic stage before asymptomatic and symptomatic stages; and we assume that the behavioral effect
+
 We then extend the model to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
@@
 Despite moderate levels of vaccine effectivenss against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity has provided limited protection against infections \citep{pearson2021omicron}.
-For example, its immune evasive property allowed the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
-therefore, even though the Omicron variant is thought to be milder \citep{MENNI20221618}, the number of deaths caused by the Omicron variant has been comparable to
-
-We hypothesize that the behavioral effect also played a role in allowing the Omicron to spread better: milder infections caused by the Omicron variant likely
-
-Even though the Omicron variant has been reported to be milder than previous variants  it has
-
-Despite high levels of vaccination
+Its immune evasive property allowed the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
+therefore, even though the Omicron variant has been estimated to be milder \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant have remained high \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
 

### 2022-05-25 figure_pre.R

@@ -1,0 +1,37 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
+library(egg)
+library(shellpipes)
+
+simulate_pre <- rdsRead()
+
+simulate_pre_max <- simulate_pre %>%
+  group_by(sim) %>%
+  filter(
+    time==max(time)
+  ) %>%
+  mutate(
+    power=paste0("f(theta)==f[0](1-theta^", power, ")"),
+    power=factor(power,
+                 levels=paste0("f(theta)==f[0](1-theta^", c(1, 2, 5, 10), ")"))
+  )
+
+g1 <- ggplot(simulate_pre_max) +
+  geom_line(aes(p_pre, D, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous(expression(Proportion~presymptomatic~"transmission,"~theta), 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, 0.01),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  facet_wrap(~power, labeller = label_parsed) +
+  theme(
+    panel.grid = element_blank(),
+    strip.background = element_blank(),
+    legend.position = "right"
+  )
+
+saveGG(g1, width=8, height=6)
+

### 2022-05-25 funs_pre.R

@@ -1,0 +1,43 @@
+model_pre <- function(t, y, parms) {
+  with(as.list(c(parms, y)), {
+    foi <- (beta_p * Ip + beta_a * Ia + (1 - delta) * beta_s * Is)
+    
+    dS <- - foi * S
+    
+    dE <- foi * S - mu * E
+    dIp <- mu * E - sigma * Ip
+    dIa <- p * sigma * Ip - gamma * Ia
+    dIs <- (1-p) * sigma * Ip - gamma * Is
+    dR <- gamma * Ia + (1-f) * gamma * Is
+    dD <- f * gamma * Is
+    
+    list(c(dS, dE, dIp, dIa, dIs, dR, dD),
+         dD=dD)
+  })
+}
+
+simulate_pre <- function(N=1,
+                         beta_a,
+                         beta_s,
+                         beta_p,
+                         mu=1/2,
+                         sigma=1/2,
+                         gamma=1/3,
+                         delta=0.8,
+                         f=0.01,
+                         p=0.5,
+                         tvec=seq(0, 365, by=1),
+                         I0=1e-4) {
+  y0 <- c(S=N-I0, E=0, Ip=I0, Ia=0, Is=0, R=0, D=0)
+  
+  parms <- c(beta_a=beta_a, beta_s=beta_s, beta_p=beta_p, mu=mu, sigma=sigma, gamma=gamma, delta=delta, f=f, p=p)
+  
+  out <- as.data.frame(ode(y0, tvec, model_pre, parms))
+  
+  out
+}
+
+p_pre_to_f <- function(x, f=0.01, 
+                       power=1) {
+  f * (1 - x^power)
+}

### 2022-05-25 simulate_pre.R

@@ -1,0 +1,39 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+## all presymptomatic transmission
+R0 <- 4
+D_s <- 3
+D_p <- 2
+p <- 0
+
+p_pre <- seq(0, 1, length.out=101)
+delta <- 0:5 * 0.2
+power <- c(1, 2, 5, 10)
+
+simparam <- expand.grid(p_pre, delta, power)
+names(simparam) <- c("p_pre", "delta", "power")
+
+simulation_pre <- apply(simparam, 1, function(x) {
+  with(as.list(x), {
+    Rpre <- R0 * p_pre
+    Rsymp <- R0 * (1-p_pre)
+    
+    beta_s <- Rsymp/D_s
+    beta_p <- Rpre/D_p
+    
+    f <- p_pre_to_f(p_pre, power=power)
+    
+    
+    out <- do.call(simulate_pre, c(as.list(x[-c(1, 3)]), f=f, beta_s=beta_s, beta_p=beta_p, beta_a=0, p=p))
+    
+    as.data.frame(append(out, x))
+  })
+  
+}) %>% bind_rows(.id="sim")
+
+rdsSave(simulation_pre)

### 2022-05-25 diagram_sub.tex
lots to commit

--- 2022-05-21 diagram_base.tex
@@
     \node [int, right=of S, ultra thick] (E)             {$E$};
-    \node [int, above right=of E, ultra thick] (Ia)             {$I_a$};
-    \node [int, below right=of E, ultra thick] (Is)             {$I_s$};
+    \node [int, right=of E, ultra thick] (Ip)             {$Ip$};
+    \node [int, above right=of Ip, ultra thick] (Ia)             {$I_a$};
+    \node [int, below right=of Ip, ultra thick] (Is)             {$I_s$};
     \node [int, right=of Ia, ultra thick] (R)             {$R$};
@@
     \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$p \mu$} (Ia);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$(1-p) \mu$} (Is);
+    \path[->, ultra thick] (E) edge node [above] {$\sigma$} (Ip);
+    \path[->, ultra thick] (Ip) edge node [sloped, above] {$p \mu$} (Ia);
+    \path[->, ultra thick] (Ip) edge node [sloped, above] {$(1-p) \mu$} (Is);
     \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);
@@
     \path[->, ultra thick, dashed] (Ia.west) edge [out=180, in=90] node [sloped, above] {$\beta_a$} ([yshift=5mm]Ts.north);
+    \path[->, ultra thick, dashed] (Ip.north) edge [out=90, in=90] node [sloped, above] {$\beta_p$} ([yshift=5mm]Ts.north);
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};
+    \node [below=15cm of S, anchor=west] (pic) {\includegraphics{figure_sub.ggp.pdf}};
 \end{tikzpicture}

### 2022-05-25 main.tex
lots to commit

@@
 
-High values of $\delta$ may seem unrealistic---in practice, $\delta$ cannot be greater than the amouont of post-symptomatic transmission.
-While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to
+High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic---in practice, $\delta$ cannot be greater than the amouont of post-symptomatic transmission.
+While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to intervention and behavioral effects already as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
+Instead, \cite{sender2021unmitigated} recently showed that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) in the absence of intervention measures;
+while this study provides indirect support for the feasibility of high $\delta$ values, it is still uncertain whether similar effects can be found in more realistic settings.
 
-mitigation steps took place and before awareness of the
-epidemic became widespread
+We therefore consider a series of mathematical models with increasing complexities to answer a more general question:
+does intermediate amount of subclinical transmission lead to peak fatalities?
 
-\citep{he2020temporal}.
-
-\cite{sender2021unmitigated} recently showed the proportion of pre-symptomatic transmission is likely lower than previously estimated
-
-can be as low as 20\% (95\%CI: 6\%--32\%) in the absence of intervention measures, providing indirect support for the plausibility of high $\delta$ values.
 In Supplementary Figure S1, we therefore consider a more realistic model which includes a presymptomatic stage before asymptomatic and symptomatic stages; and we assume that the behavioral effect

### 2022-05-25 figure_sub.R

@@ -3,9 +3,9 @@ library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
 library(egg)
 library(shellpipes)
 
-simulate_pre <- rdsRead()
+simulate_sub <- rdsRead()
 
-simulate_pre_max <- simulate_pre %>%
+simulate_sub_max <- simulate_sub %>%
   group_by(sim) %>%
   filter(
     time==max(time)
@@ -14,25 +14,27 @@ simulate_pre_max <- simulate_pre %>%
     power=paste0("f(theta)==f[0](1-theta^", power, ")"),
     power=factor(power,
                  levels=paste0("f(theta)==f[0](1-theta^", c(-1, 1, 2, 5), ")"),
-                 labels=c("No~tradeoff", paste0("f(theta)==f[0](1-theta^", c(1, 2, 5), ")")))
+                 labels=c("No~tradeoff", paste0("f(theta)==f[0](1-theta^", c(1, 2, 5), ")"))),
+    p_pre_ratio=factor(paste0(p_pre_ratio*100, "*\'%\'"),
+                       levels=paste0(0:5*0.2*100, "*\'%\'"))
   )
 
-g1 <- ggplot(simulate_pre_max) +
-  geom_line(aes(p_pre, D, col=factor(delta), group=factor(delta))) +
-  scale_x_continuous(expression(Proportion~presymptomatic~"transmission,"~theta), 
+g1 <- ggplot(simulate_sub_max) +
+  geom_line(aes(p_sub, D, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous(expression(Proportion~subclinical~"transmission,"~phi), 
                      breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
                      labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
                      expand=c(0, 0)) +
   scale_y_continuous("Proportion deceased",
-                     limits=c(0, 0.01),
+                     limits=c(0, 0.0103),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(delta)) +
-  facet_wrap(~power, labeller = label_parsed) +
+  facet_grid(p_pre_ratio~power, labeller = label_parsed) +
   theme(
     panel.grid = element_blank(),
     strip.background = element_blank(),
     legend.position = "right"
   )
 
-saveGG(g1, width=8, height=6)
+saveGG(g1, width=8, height=8)
 

### 2022-05-25 figure_sub2.R

@@ -1,0 +1,40 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
+library(egg)
+library(shellpipes)
+
+simulate_sub <- rdsRead()
+
+simulate_sub_max <- simulate_sub %>%
+  group_by(sim) %>%
+  filter(
+    time==max(time)
+  ) %>%
+  mutate(
+    power=paste0("f(phi)==f[0](1-phi^", power, ")"),
+    power=factor(power,
+                 levels=paste0("f(phi)==f[0](1-phi^", c(-1, 1, 2, 5), ")"),
+                 labels=c("No~tradeoff", paste0("f(phi)==f[0](1-phi^", c(1, 2, 5), ")"))),
+    p_pre_ratio=factor(paste0(p_pre_ratio*100, "*\'%\'"),
+                       levels=paste0(0:5*0.2*100, "*\'%\'"))
+  )
+
+g1 <- ggplot(simulate_sub_max) +
+  geom_line(aes(p_sub, D, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous(expression(Proportion~subclinical~"transmission,"~phi), 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, 0.0103),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  facet_grid(p_pre_ratio~power, labeller = label_parsed) +
+  theme(
+    panel.grid = element_blank(),
+    strip.background = element_blank(),
+    legend.position = "right"
+  )
+
+saveGG(g1, width=8, height=8)
+

### 2022-05-25 funs_sirs.R

@@ -1,0 +1,56 @@
+model_sirs <- function(t, y, parms) {
+  with(as.list(c(parms, y)), {
+    beta <- beta * (1 + phi * cos(2 * pi * t/365))
+    
+    foi <- (beta * (Ia_n + Ia_p) + (1 - delta) * beta * (Is_n + Is_p))
+    
+    dS_n <- b - foi * S_n - b * S_n ## naive
+    dS_p <- sigma * R_n + sigma * R_p - (1-epsilon_i) * foi * S_p - b * S_p ## protected
+    
+    dE_n <- foi * S_n - mu * E_n- b * E_n
+    dIa_n <- p * mu * E_n - gamma * Ia_n - b * Ia_n
+    dIs_n <- (1-p) * mu * E_n - gamma * Is_n - b * Is_n
+    
+    dR_n <- gamma * Ia_n + (1-f) * gamma * Is_n - b * R_n - sigma * R_n
+    
+    dD_n <- f * gamma * Is_n
+    
+    dE_p <- (1-epsilon_i) * foi * S_p - mu * E_p - b * E_p
+    dIa_p <- (1-(1-epsilon_s) * (1-p)) * mu * E_p - gamma * Ia_p - b * Ia_p
+    dIs_p <- (1-epsilon_s) * (1-p) * mu * E_p - gamma * Is_p - b * Ia_p
+    
+    dR_p <- gamma * Ia_p + (1-(1-epsilon_d) * f) * gamma * Is_p - b * R_p - sigma * R_p
+    
+    dD_p <- (1-epsilon_d) * f * gamma * Is_p
+    
+    list(c(dS_n, dS_p, dE_n, dIa_n, dIs_n, dR_n, dD_n, dE_p, dIa_p, dIs_p, dR_p, dD_p),
+         dD_n=dD_n, dD_p=dD_p)
+  })
+}
+
+simulate_sirs <- function(N=1,
+                     S0=1,
+                     D0=0,
+                     beta=4/5,
+                     mu=1/2,
+                     gamma=1/5,
+                     delta=0.9,
+                     sigma=1/365,
+                     b=1/365/80,
+                     f=0.01,
+                     p=0.25,
+                     phi=0.1,
+                     epsilon_i=0,
+                     epsilon_s=0,
+                     epsilon_d=0,
[…]

### 2022-05-25 simulate_pre.R

@@ -13,7 +13,7 @@ p <- 0
 
 p_pre <- seq(0, 1, length.out=101)
 delta <- 0:5 * 0.2
-power <- c(1, 2, 5, 10)
+power <- c(-1, 1, 2, 5)
 
 simparam <- expand.grid(p_pre, delta, power)
 names(simparam) <- c("p_pre", "delta", "power")
@@ -26,8 +26,11 @@ simulation_pre <- apply(simparam, 1, function(x) {
     beta_s <- Rsymp/D_s
     beta_p <- Rpre/D_p
     
-    f <- p_pre_to_f(p_pre, power=power)
-    
+    if (power==-1) {
+      f <- p_pre_to_f(0)
+    } else {
+      f <- p_pre_to_f(p_pre, power=power)
+    }
     
     out <- do.call(simulate_pre, c(as.list(x[-c(1, 3)]), f=f, beta_s=beta_s, beta_p=beta_p, beta_a=0, p=p))
     

### 2022-05-25 simulate_sub.R

@@ -1,0 +1,90 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+R <- 4
+D_p <- 2
+D_s <- D_a <- 3
+ratio <- 0.75
+
+p_sub <- seq(0, 1, length.out=101) # proportion of subclinical transmission
+p_pre_ratio <- 0:5 * 0.2 # proportion of subclinical transmission that is caused by presymptomatic transmission
+delta <- 0:5 * 0.2
+power <- c(-1, 1, 2, 5)
+
+simparam <- expand.grid(p_sub, p_pre_ratio, delta, power)
+names(simparam) <- c("p_sub", "p_pre_ratio", "delta", "power")
+
+simulation_sub <- apply(simparam, 1, function(x) {
+  with(as.list(x), {
+    # what we're given
+    # Rtotal = p * Rasymp + (1-p) * Rsymp + Rpre
+    # p_sub = (p * Rasymp + Rpre)/Rtotal
+    # p_pre_ratio = Rpre/(p * Rasymp + Rpre)
+    # Rasymp/Rsymp = ratio
+    # R = Rsymp + Rpre
+    
+    # Rtotal = p * Rasymp + (1-p) * Rsymp + Rpre
+    # p_sub * (p * Rasymp + (1-p) * Rsymp + Rpre) = p * Rasymp + Rpre
+    # p * Rasymp + Rpre = Rpre/p_pre_ratio
+    # p * Rasymp = (1/p_pre_ratio-1) * Rpre
+    # p = (1/p_pre_ratio-1) * Rpre/Rasymp 
+    # p= (1/p_pre_ratio-1) * Rpre * 1/ratio/Rsymp
+    
+    # p_sub * ((1/p_pre_ratio) * Rpre + (1-p) * Rsymp) = (1/p_pre_ratio) * Rpre
+    
+    # (1-p) * Rsymp= Rsymp-(1/p_pre_ratio-1) * Rpre/ratio
+    
+    # (1/p_pre_ratio) * Rpre + Rsymp -(1/p_pre_ratio-1) * Rpre/ratio = Rpre/p_pre_ratio/p_sub
+    # (1/p_pre_ratio) + Rsymp/Rpre - (1/p_pre_ratio-1)/ratio = 1/p_pre_ratio/p_sub
[…]

### 2022-05-25 simulate_sub2.R

@@ -1,0 +1,88 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+R <- 4
+D_p <- 2
+D_s <- D_a <- 3
+ratio <- 0.75
+
+p_sub <- seq(0, 1, length.out=101) # proportion of subclinical transmission
+p_pre_ratio <- 0:5 * 0.2 # proportion of subclinical transmission that is caused by presymptomatic transmission
+delta <- 0:5 * 0.2
+power <- c(-1, 1, 2, 5)
+
+simparam <- expand.grid(p_sub, p_pre_ratio, delta, power)
+names(simparam) <- c("p_sub", "p_pre_ratio", "delta", "power")
+
+simulation_sub <- apply(simparam, 1, function(x) {
+  with(as.list(x), {
+    # what we're given
+    # Rtotal = p * Rasymp + (1-p) * Rsymp + Rpre
+    # p_sub = (p * Rasymp + Rpre)/Rtotal
+    # p_pre_ratio = Rpre/(p * Rasymp + Rpre)
+    # Rasymp/Rsymp = ratio
+    # R = Rsymp + Rpre
+    
+    # Rtotal = p * Rasymp + (1-p) * Rsymp + Rpre
+    # p_sub * (p * Rasymp + (1-p) * Rsymp + Rpre) = p * Rasymp + Rpre
+    # p * Rasymp + Rpre = Rpre/p_pre_ratio
+    # p * Rasymp = (1/p_pre_ratio-1) * Rpre
+    # p = (1/p_pre_ratio-1) * Rpre/Rasymp 
+    # p= (1/p_pre_ratio-1) * Rpre * 1/ratio/Rsymp
+    
+    # p_sub * ((1/p_pre_ratio) * Rpre + (1-p) * Rsymp) = (1/p_pre_ratio) * Rpre
+    
+    # (1-p) * Rsymp= Rsymp-(1/p_pre_ratio-1) * Rpre/ratio
+    
+    # (1/p_pre_ratio) * Rpre + Rsymp -(1/p_pre_ratio-1) * Rpre/ratio = Rpre/p_pre_ratio/p_sub
+    # (1/p_pre_ratio) + Rsymp/Rpre - (1/p_pre_ratio-1)/ratio = 1/p_pre_ratio/p_sub
[…]

### 2022-05-26 main.tex
close to finishing...

@@
 \textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
-(Left) $S$ represents suceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(Left) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
@@
 \fref{base} evaluates epidemic outcomes using epidemiological parameters similar to that of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
-In the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproductin number:
+In the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number:
 \begin{equation}
@@
 therefore, the number of infections increases with $p$.
-For even higher values of $\delta$, we find that there is a critical level of asymptomatic propotion, $p_c$:
+For even higher values of $\delta$, we find that there is a critical level of asymptomatic proportion, $p_c$:
 \begin{equation}
@@
 More intricate patterns emerge for fatalities.
-When $\delta = 0$, total fataities decrease with $p$ because IFR ($(1-p)f$) decreases.
+When $\delta = 0$, total fatalities decrease with $p$ because IFR ($(1-p)f$) decreases.
 In contrast, for high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
@@
 
-High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic---in practice, $\delta$ cannot be greater than the amouont of post-symptomatic transmission.
+High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic---in practice, $\delta$ cannot be greater than the amount of post-symptomatic transmission.
 While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to intervention and behavioral effects already as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
@@
 
-We therefore consider a series of mathematical models with increasing complexities to answer a more general question:
-does intermediate amount of subclinical transmission lead to peak fatalities?
+We therefore consider two additional of mathematical models with increasing complexities to answer a more general question:
+does intermediate amount of subclinical (including both asymptomatic and presymptomatic) transmission lead to peak fatalities?
+First, we analyze a model with presymptomatic transmission alone (Supplementary Figure S1).
+When the reproduction number is fixed, increasing the proportion of presymptomatic transmission $\theta$ causes larger outbreaks, and therefore more deaths, in the presence of symptom-responsive transmission reduction ($\delta > 0$).
+However, standard compartmental models implicitly assume that the fatality rate $f$ is independent of the amount of presymptomatic transmission---in the extreme case where all transmission happens presymptomatically, infected individuals no longer shed infectious virus after symptom onset, meaning that they should be less likely to die from infection.
+We therefore assume a tradeoff between the amount of presymptomatic transmission $\theta$ and the fatality rate $f$ that follows a power law: $f(\theta) = f_0 (1-\theta^a)$, which monotonically decreases from $f_0$ to 0.
+In the simple case where the fatality rate decreases linearly with $\theta$, the relationship between the amount of presymptomatic transmission $\theta$ and total fatalities is equivalent to that between the amount of asymptomatic transmission and total fatalities in the original model (\fref{base}).
+As we increase the exponent $a$, we obtain a more realistic tradeoff curve such that fatality rate $f$ remains roughly constant for low to intermediate values of $\theta$ and suddenly decreases to $0$ as $\theta$ approaches 1;
+in these cases, peak fatalities occur at intermediate levels of presymptomatic transmission for even lower values of $\delta=0.6$, which now represents reduction in transmission after symptom onset.
 
-In Supplementary Figure S1, we therefore consider a more realistic model which includes a presymptomatic stage before asymptomatic and symptomatic stages; and we assume that the behavioral effect
+We then extend our model to consider the effects of generalized \textit{subclinical} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S2).
+In particular, we fix the reproduction number of symptomatic individuals and calculate the proportion of deceased individuals for given values of proportion of total subclinical transmission and proportion of subclinical transmission that is caused by presymptomatic transmission.
+This model generalizes two earlier models:
+when all subclinical transmission is caused by asymptomatic transmission, we have the first model (\fref{base});
+and when all subclinical transmission is caused by presymptomatic transmission, we have the second model (Supplementary Figure S1).
+Across almost all ranges of scenarios we consider, we find that peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effect, $\delta > 0.6$ (Supplementary Figure S2);
+in particular, we find that this pattern is robust even in the absence of the tradeoff between the amount of presymptomatic transmission and the fatality rate.
+One exception is the case in which all subclinical transmission is caused by presymptomatic transmission as shown earlier.
+Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions are expected to hold for more general cases with respect to subclinical transmission.
 
-We then extend the model to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
+We now apply our framework to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
 The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
@@
 Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
-This also means that the critical level of asymptomatic propotion decreases, allowing more dangeous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
 Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immunity}C).
@@
 
-Our simulatons of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
-Despite moderate levels of vaccine effectivenss against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity has provided limited protection against infections \citep{pearson2021omicron}.
+Our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
+Despite moderate levels of vaccine effectiveness against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity has provided limited protection against infections \citep{pearson2021omicron}.
 Its immune evasive property allowed the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
@@
 between asymptomatic and symptomatically infected individuals.
-Here, $\delta$ denotes the reduction in transmissability due to responsive measures taken by symptomatically infected indivivduals.
+Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
 Note that the SEIR model framework does not impact final size outcomes, so for convenience, all final size simulations ignore the $E$ class.

### 2022-05-27 main.tex
done for now

@@
 (Right) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, and $f=0.01$.
+We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$.
 We assume that $10^{-4}$ proportion of individuals are initially infected.
@@
 
-The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immunity}A).
+The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}A).
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
 The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
-the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immunity}B).
+the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
 Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
 This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
-Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immunity}C).
+Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}C).
 
@@
 \textbf{Dynamics of invading variants under symptom-responsive transmission reduction and immunity.}
-(Top) The subscript $p$ represents protected individuals. 
-Immunity may provide protection against infection, symptoms, or deaths.
-The dynamics of immunologically naive individuals are described in (\fref{base}).
-(Bottom) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
-We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$ for illustration.
-We assume that $10^{-4}$ proportion of individuals are initially infected.
+(A, D) Asymptomaticity--fatality curves for the first (solid lines) and second waves (dashed lines).
+Points represent specific scenarios we assume for the first and second waves.
+Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using same parameters as \fref{base} and $\delta=0.8$.
+Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ (wildtype) for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune}.
+(B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
+(C, F) Dynamics of daily deaths for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
 }
@@
 Finally, we use our framework to understand the impact of behavioral effects on invading variants (\fref{variant}).
-In doing so, we first simulate the dynamics of a resident variant for 1 year using our base model (\fref{base}).
-We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the resident variant.
-We consider two types of variants (which are simulated separately): one with the same severity $p$ (orange) and a milder one with higher $p$ (purple).
+In doing so, we first simulate the dynamics of a wildtype variant for 1 year using our base model (\fref{base}).
+We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the wildtype variant.
+We consider two types of variants (which are simulated separately): one with the same severity $p$ (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
 
@@
 Although, a milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
-The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, resident variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
+The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
 In general, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms, given that asymptomatically infected individuals can transmit better than symptomatically infected individuals (due to either behavioral and intervention effects).
@@
 Next, we consider a more realistic scenario in which immunity provides protection against symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
-In this case, cross protection against infection has a disproportionately larger effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, resident variant.
+In this case, cross protection against infection has a disproportionately larger effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
 Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level.
@@
 There are several limitations to our analysis.
-First of all, behavioral and intervention effects must
+First of all, behavioral and intervention effects must sufficiently high in order for the fatality to peak at intermediate levels of asymptomaticity.
+While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of subclinical transmission on population-level outcomes.
+\swp{what are some other limitations we should address?}
 
@@
 Although mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and in the event that individuals are at risk for severe outcomes.
-As more variants continue to emerge, updating vaccines to prevent infections, and not just diseases, is
+As more variants continue to emerge, updating vaccines to prevent infections, and not just diseases, will be critical to controlling the course of the pandemic.
 
-\section*{Methods}
-\label{sec.methods}
-The model dynamics are as follows
-\begin{eqnarray}
-\dot{S}=\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} = \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
-\dot{I}_a = p \mu E - \gamma_a I_a\\
-\dot{I}_s = (1-p) \mu E -\gamma_s I_s\\
-\dot{R} = \gamma_a I_a + (1-f) \gamma_s I_s \\
-\dot{D} = f \gamma_s I_s
-\end{eqnarray}
-where the transmission rate $\beta$ and recover rate $\gamma$ can be potentially differ
-between asymptomatic and symptomatically infected individuals.
+\pagebreak
+
+\section*{Supplementary Materials}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\subsection*{Methods}
+
+\subsubsection*{Models without immunity}
+
+First, we consider a simple, compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
+The model dynamics are as follows:
+\begin{align}
+\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
+\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
+\dot{I}_a &= p \mu E - \gamma_a I_a\\
+\dot{I}_s &= (1-p) \mu E -\gamma_s I_s\\
+\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
+\dot{D} &= f \gamma_s I_s
+\end{align}
+where the transmission rate $\beta$ and recovery rate $\gamma$ can be potentially differ between asymptomatic and symptomatically infected individuals.  
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
-Note that the SEIR model framework does not impact final size outcomes, so for convenience, all final size simulations ignore the $E$ class.
+Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\mu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$.
+
+We then consider a model with presymptomatic transmission (and without asymptomatic transmission), where all infected individuals eventually develop symptoms:
+\begin{align}
+\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
+\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
+\dot{I}_p &= \mu E - \sigma I_p\\
+\dot{I}_s &= \sigma I_p -\gamma_s I_s\\
+\dot{R} &= (1-f) \gamma_s I_s \\
+\dot{D} &= f \gamma_s I_s
+\end{align}
+For this model, the presymptomatic $\RR_p$ and symptomatic $\RR_s$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$ and $\RR_s=\beta_s/\gamma_s$ in the absence of the behavioral effect; and the basic reproduction number is equal to the sum of the two.
+Then, the intrinsic proportion of presymptomatic transmission is given by: 
+\begin{equation}
+\theta = \frac{\RR_p}{\RR_p + \RR_s}.
+\end{equation}
+We assume there is a tradeoff between the proportion of presymptomatic transmission and fatality rate: 
+\begin{equation}
+f(\theta) = f_0 (1-\theta^a),
+\end{equation}
+where $f_0=0.01$ represents the baseline fatality rate and the exponent $a$ is varied between 1 and 5.
+Throughout simulations, we assume $\mathcal R_0 = 4$, $1/\sigma=2\,\mathrm{days}$, and $1/\gamma_s=3\,\mathrm{days}$.
+All other parameters are same as before.
+
+Finally, we combine both models to include both presymptomatic and asymptomatic transmission:
+\begin{align}
+\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
+\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
+\dot{I}_p &= \mu E - \sigma I_p\\
+\dot{I}_a &= p \sigma I_p - \gamma_a I_a\\
+\dot{I}_s &= (1-p) \sigma I_p -\gamma_s I_s\\
+\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
+\dot{D} &= f \gamma_s I_s
+\end{align}
+For this model, the reproduction number of individuals who will eventually develop symptoms is equal to: $\RR_p + \RR_s$;
+similarly, the reproduction number of individuals who remain asymptomatic is equal to: $\RR_p + \RR_a$.
+Since proportion $p$ of all infections is asymptomatic, the basic reproduction number is given by the weighted average of these two reproduction numbers: 
+\begin{equation}
+\RR_0 = p(\RR_p + \RR_a) + (1-p) (\RR_p + \RR_s) = \RR_p + p \RR_a + (1-p) \RR_s.
+\end{equation}
+Then, the proportion of subclinical transmission $\phi$ is given by:
+\begin{equation}
+\phi = \frac{\RR_p + p \RR_a}{\RR_0}.
+\end{equation}
+
+For simulations of the combined model, we start by fixing the reproduction number of individuals who will eventually develop symptoms: $\RR_{\textrm{symp}} = \RR_p + \RR_s = 4$. 
+Consistent with previous assumptions, we also assume that asymptomatic reproduction number is lower than that of the symptomatic reproduction number: $\RR_a = \rho \RR_s$ where $\rho = 0.75$.
+Then, for a given value of the proportion of subclinical transmission $\phi$ and proportion of subclinical transmission caused by the presymptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
+More specifically:
+\begin{align}
+\RR_p &= \frac{\RR_{\textrm{symp}}}{1 + y}\\
+\RR_s &= \RR_{\textrm{symp}} - \RR_p\\
+\RR_a &=  \rho \RR_s\\
+p &= \left(\frac{1}{\eta} - 1 \right) \frac{\RR_p}{\RR_a},
+\end{align}
+where $y = (1/\phi - 1)/\eta + (1/\eta - 1)/\rho$.
+All other parameters are same as before.
+
+\subsubsection*{Model with immunity}
+
+We then model the spread of infection in a partially immune population.
+The model dynamics are as follows:
+\begin{align}
+\dot{S} &= -\lambda (t) S \\
+\dot{E} &= \lambda (t) S - \mu E\\
+\dot{I}_a &= p \mu E - \gamma_a I_a\\
+\dot{I}_s &= (1-p) \mu E -\gamma_s I_s\\
+\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
+\dot{D} &= f \gamma_s I_s\\
+\dot{S}_p &= - (1-\epsilon_i) \lambda (t) S_p \\
+\dot{E}_p &= (1-\epsilon_i) \lambda (t) S_p - \mu E_p\\
+\dot{I}_{p, a} &= (1 - (1-\epsilon_s) (1-p)) \mu E_p - \gamma_a I_{p,a}\\
+\dot{I}_{p, s} &= (1-\epsilon_s) (1-p) \mu E_p -\gamma_s I_{p,s}\\
+\dot{R}_p &= \gamma_a I_{p,a} + (1-(1-\epsilon_d) f) \gamma_s I_{p,s} \\
+\dot{D}_p &= (1-\epsilon_d) f \gamma_s I_{p,s}
+\end{align}
+where $\epsilon$ represents the degree of protection against infection, symptoms and death. 
+The force of infection $\lambda(t)$ is given by:
+\begin{equation}
+\lambda(t) = \beta_a (I_a + I_{p,a}) + (1-\delta) \beta_s (I_s + I_{p,s}).
+\end{equation}
+Here, subscripts $p$ denote individuals who are immune and therefore are protected.
+
+\pagebreak
+
+\section*{Supplementary Figures}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{diagram_pre.pdf}
+\caption{
+\textbf{Schematic diagram and simulations of a model with presymptomatic transmission and symptom-responsive transmission reduction.}
+(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+See Methods for model details.
+(Bottom) Total deaths as a function of the proportion of presymptomatic transmission $\theta$ across a wide range scenarios for $\delta$ and a tradeoff between the proportion of presymptomatic infections $\theta$ and fatality rates $f$.
+}
+\end{center}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{diagram_sub.pdf}
+\caption{
+\textbf{Schematic diagram and simulations of a model with presymptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
+(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+See Methods for model details.
+(Bottom) Total deaths as a function of the proportion of subclinical transmission $\phi$ across a wide range scenarios for $\delta$ and a tradeoff between the proportion of presymptomatic infections $\theta$ and fatality rates $f$.
+}
+\end{center}
+\end{figure}
+
+\pagebreak
 

### 2022-05-27 figure_variant.R

@@ -14,21 +14,21 @@ simulate_orig_profile_summarize <- simulate_orig_profile %>%
   group_by(sim) %>%
   filter(time==max(time)) %>%
   mutate(
-    type="Resident"
+    type="First wave"
   )
 
 simulate_var_profile_summarize <- simulate_var_profile %>%
   group_by(sim) %>%
   filter(time==max(time)) %>%
   mutate(
-    type="Invading"
+    type="Second wave"
   )
 
 simulate_var_cross_profile_summarize <- simulate_var_cross_profile %>%
   group_by(sim) %>%
   filter(time==max(time)) %>%
   mutate(
-    type="Invading"
+    type="Second wave"
   )
 
 g1 <- ggplot(simulate_orig_profile_summarize, aes(p, (D_p+D_n), lty=type)) +
@@ -40,9 +40,9 @@ g1 <- ggplot(simulate_orig_profile_summarize, aes(p, (D_p+D_n), lty=type)) +
   #         arrow = arrow(length=unit(0.1, "inches")), col="purple") +
   geom_line() +
   geom_line(data=simulate_var_profile_summarize) +
-  geom_point(x=p_orig, y=tail(simulate_orig$D_p+simulate_orig$D_n, 1), size=3) +
-  geom_point(x=p_var1, y=tail(simulate_var1$D_p+simulate_var1$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1), col="orange", size=3, shape="triangle") +
-  geom_point(x=p_var2, y=tail(simulate_var2$D_p+simulate_var2$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1), col="purple", size=3, shape="square") +
+  geom_point(x=p_orig, y=tail(simulate_orig$D_p+simulate_orig$D_n, 1), size=3, aes(col="Wildtype", shape="Wildtype")) +
+  geom_point(x=p_var1, y=tail(simulate_var1$D_p+simulate_var1$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1), size=3, aes(col="Variant 1", shape="Variant 1")) +
+  geom_point(x=p_var2, y=tail(simulate_var2$D_p+simulate_var2$D_n, 1)-tail(simulate_orig$D_p+simulate_orig$D_n, 1), size=3, aes(col="Variant 2", shape="Variant 2")) +
[…]

### 2022-05-30 main.tex
some thoughts and comments

@@
 
-SARS-CoV-2 has had devastating effects at the population level even as individuals have experienced highly disparate infection outcomes.
-Notably, many individuals have mild or even asymptomatic cases making it harder to estimate the magnitude of spread and fatality rate. 
-The documented case-fatality rate (CFR) at the population level varies with population, testing infrastructure, and case definitions, typically between 1\%--4\% \citep{rajgor2020many,VERITY2020669,yang2020early}.
-But, many infections are never documented. 
-The infection fatality rate (IFR), which takes undocumented infections into account, has been estimated to be closer to 0.5\%--1\% for demographics similar to that of the United States \citep{levin2020assessing}.
+SARS-CoV-2 has had devastating effects at the population level.
+However, many individuals have mild cases, making it harder to estimate the magnitude of spread and fatality rate. 
+The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \%--4\% \citep{rajgor2020many,VERITY2020669,yang2020early}.
+But many infections are never documented;
+the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for demographics similar to that of the United States \citep{levin2020assessing}.
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
-Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical and even asymptomatic. 
-Yet, asymptomatically infected individuals can still transmit to others. 
-Hence, asymptomatic infections can be a double-edged sword.
-On the one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, and are may therefore be more likely to infect others, since precautions are less likely to be taken.
+Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical or even asymptomatic.
+
+Individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
+This means that the presence of asymptomatic infections has countervailing effects at the population level. 
+On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, and may therefore be more likely to infect others, since precautions are less likely to be taken.
 This means that the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
@@
 Asymptomatically and symptomatically infected individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
-Our key assumption is symptomatically infected individuals take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
+Our key assumption is that symptomatically infected individuals take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
 we note that the parameter $\delta$ may also capture intervention measures that target symptomatically infected individuals, such as symptom-based isolation. 
-As our nominal parameters, we assume lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
-We assume that asymptomatic individuals do not die, and evaluate the effects of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
+For our main simulations, we assume lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
+We assume that asymptomatic individuals do not die, and evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
+\jd{It's probably friendlier to talk mostly about \RR, we can then mention $\gamma$ only briefly.}
 
@@
 
-\fref{base} evaluates epidemic outcomes using epidemiological parameters similar to that of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
+\fref{base} shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
 In the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number:
@@
 This relationship changes as $\delta$ increases.
-In particular, when $\delta > 1-\RR_a/\RR_s$ (in our case, $\delta > 0.25)$, the basic reproduction number increases with $p$ because the behavioral effect causes the effective symptomatic transmission rate to be lower than that that of asymptomatic individuals;
-therefore, the number of infections increases with $p$.
-For even higher values of $\delta$, we find that there is a critical level of asymptomatic proportion, $p_c$:
+In particular, when $\delta > 1-\RR_a/\RR_s$ (in our case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic transmission rate (including behavioral response) is less than that the asymptomatic rate.
+For $\delta$ in this range, we can find a critical level of asymptomatic proportion, $p_c$:
 \begin{equation}
@@
 such that $p>p_c$ is required for an outbreak. 
-In the limit that $\delta=1$ (i.e., when symptomatically individuals does not transmit at all), then $p_c=1/\RR_a$.
+\jd{ff.~s.~seems unnecessary:} In the limit that $\delta=1$ (i.e., when symptomatically individuals does not transmit at all), then $p_c=1/\RR_a$.
 
-More intricate patterns emerge for fatalities.
-When $\delta = 0$, total fatalities decrease with $p$ because IFR ($(1-p)f$) decreases.
-In contrast, for high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
+When behavioral protection is high, the effect of asymptomatic proportion on fatalities shows countervailing effects of individual-level protection and population-level risk.
+For high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
 although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
+In contrast, when $\delta$ is small enough that $\RR_s\geq\RR_a$, then total fatalities decrease with $p$ because the IFR ($(1-p)f$) decreases.
 
@@
 while this study provides indirect support for the feasibility of high $\delta$ values, it is still uncertain whether similar effects can be found in more realistic settings.
+\jd{We need to be a bit more nuanced here: we get the low value by focusing on just a couple of weeks, so we're sort of suggesting a counter-factual where nobody notices COVID for a long time.}
 
@@
 However, standard compartmental models implicitly assume that the fatality rate $f$ is independent of the amount of presymptomatic transmission---in the extreme case where all transmission happens presymptomatically, infected individuals no longer shed infectious virus after symptom onset, meaning that they should be less likely to die from infection.
+\jd{We need a bit more detail here.
+I have no idea what proportion of COVID deaths are post-viral.
+Maybe I'm COVID-ignorant, but I want more info: many flu deaths are post-viral, and it's not super-clear how much of a role viral replication (or escape from deep lungs?) plays in fatality.}
 We therefore assume a tradeoff between the amount of presymptomatic transmission $\theta$ and the fatality rate $f$ that follows a power law: $f(\theta) = f_0 (1-\theta^a)$, which monotonically decreases from $f_0$ to 0.
@@
 and when all subclinical transmission is caused by presymptomatic transmission, we have the second model (Supplementary Figure S1).
-Across almost all ranges of scenarios we consider, we find that peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effect, $\delta > 0.6$ (Supplementary Figure S2);
-in particular, we find that this pattern is robust even in the absence of the tradeoff between the amount of presymptomatic transmission and the fatality rate.
-One exception is the case in which all subclinical transmission is caused by presymptomatic transmission as shown earlier.
-Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions are expected to hold for more general cases with respect to subclinical transmission.
+We find a wide variety of scenarios for which peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effect, $\delta > 0.6$ (Supplementary Figure S2);
+this pattern is robust even to the absence of a tradeoff between the amount of presymptomatic transmission and the fatality rate.
+One exception is the case discussed earlier in which all subclinical transmission is caused by presymptomatic transmission.
+Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of subclinical transmission.
 
@@
 The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$ (\fref{immune}).
-For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously; we do not consider the effect of immunity on transmission.
-We assume a relatively strong behaviroal effect $\delta=0.8$ for illustration.
+For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously; we do not consider the separate effect of immunity on transmission (beyond the effect on infection).
+We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
 

### 2022-05-30 main.tex
comments b4 meeting

@@
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
+\jd{The exact equivalance is sort of an assumption: we shold find an opportunity to say explicitly that we assume that asymptomatic infections in protected and unprotected people have the same \RR.}
 The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
 the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
-Therefore, for low values of $p$, protection against symptoms can inadvertently increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
-This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
+\jd{Do we need ff.~s.?} This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
 Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}C).
@@
 Points represent specific scenarios we assume for the first and second waves.
-Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using same parameters as \fref{base} and $\delta=0.8$.
-Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ (wildtype) for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune}.
+Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using parameters from \fref{base} with $\delta=0.8$.
+Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for two different values of $p$ as shown.
 (B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
@@
 In this case, protection against symptoms allows new variants to spread faster, resulting in larger outbreaks (\fref{variant}B).
-Although, a milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
+Although the milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
 The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
 In general, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms, given that asymptomatically infected individuals can transmit better than symptomatically infected individuals (due to either behavioral and intervention effects).
+\jd{Be more clear what comparison we are making here. If we are comparing immunity to symptoms with immunity to infection, this s.~can be sharper and clearer. If not, what are we comparing?}
 

### 2022-06-01 main.tex
some more thoughts

@@
 
-Next, we consider a more realistic scenario in which immunity provides protection against symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
-In this case, cross protection against infection has a disproportionately larger effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
+Next, we consider a more realistic scenario in which immunity provides protection against both symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
+In this case, cross protection against infection has a large effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
 Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level.
 
-In summary, using a simplified model we have shown that asymptomatic infections can represent a double-edge sword insofar as they represent a better outcome for some individuals but a mechanism for onwards transmission that leads to a worse outcome for the population as a whole.
+In summary, using a simplified model we have shown that asymptomatic infections can represent a double-edged sword by providing a better outcome for some individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variants:
-while protection against symptoms prevent diseases at the individual level, it can inadvertently lead to more infections, and potentially deaths, at the population level.
-For example, milder variants can result in even faster epidemic growth rates and larger outbreaks, even though it can lead to less fatalities if the variant is sufficiently mild.
+while protection against symptoms protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
 
 Our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
-Despite moderate levels of vaccine effectiveness against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity has provided limited protection against infections \citep{pearson2021omicron}.
-Its immune evasive property allowed the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
-therefore, even though the Omicron variant has been estimated to be milder \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant have remained high \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
+Despite moderate levels of vaccine effectiveness against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \citep{pearson2021omicron}.
+This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
+even though the Omicron variant is probably milder than XXX \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was high \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
 
 There are several limitations to our analysis.
-First of all, behavioral and intervention effects must sufficiently high in order for the fatality to peak at intermediate levels of asymptomaticity.
+First of all, behavioral and intervention effects must be large in order for the fatality to peak at intermediate levels of asymptomaticity.
 While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of subclinical transmission on population-level outcomes.
 \swp{what are some other limitations we should address?}
+\jd{Something about estimating protection against different endpoints. We should also highlight the importance of doing this estimation.}
 
@@
 As more variants continue to emerge, updating vaccines to prevent infections, and not just diseases, will be critical to controlling the course of the pandemic.
+\jd{This is complicated.
+Maybe regular waves of mild SC2 are a good way to control morbidity in the medium term.
+There is also an argument to make about symptom-focused responses and evolution toward mildness.}
 

### 2022-06-06 main.tex
some work before going to korea

@@
 we note that the parameter $\delta$ may also capture intervention measures that target symptomatically infected individuals, such as symptom-based isolation. 
-For our main simulations, we assume lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
+For our main simulations, we assume that asymptomatically infected individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
 We assume that asymptomatic individuals do not die, and evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
-\jd{It's probably friendlier to talk mostly about \RR, we can then mention $\gamma$ only briefly.}
-
 \begin{figure}[!ht]
@@
 such that $p>p_c$ is required for an outbreak.
-\jd{ff.~s.~seems unnecessary:} In the limit that $\delta=1$ (i.e., when symptomatically individuals does not transmit at all), then $p_c=1/\RR_a$.
 
@@
 While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to intervention and behavioral effects already as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
-Instead, \cite{sender2021unmitigated} recently showed that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) in the absence of intervention measures;
-while this study provides indirect support for the feasibility of high $\delta$ values, it is still uncertain whether similar effects can be found in more realistic settings.
+Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
+There are two implications for the discrepancy in the estimates of the proportion of presymptomatic transmission---first, a low proportion of presymptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) relative to the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of presymptomatic transmission from 20\% to 60\%.
 \jd{We need to be a bit more nuanced here: we get the low value by focusing on just a couple of weeks, so we're sort of suggesting a counter-factual where nobody notices COVID for a long time.}
+\swp{How's this?}
 
@@
 The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$ (\fref{immune}).
-For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously; we do not consider the separate effect of immunity on transmission (beyond the effect on infection).
+For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously.
+We also do not consider the separate effect of immunity on transmission (beyond the effect on infection).
+In other words, we assume that asymptomatic infections in protected and unprotected people have the same reproduction numbers (and likewise for the symptomatic infections).
+In practice, both asymptomatic and symptomatic infections in protected people are less likey to transmit than their unprotected counterparts: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral laod), in which case symptomaticity can be a poor proxy for transmission.
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
+\swp{JD: see if you like what I did in this P, esp in terms of boosting.}
 
@@
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
-\jd{The exact equivalance is sort of an assumption: we shold find an opportunity to say explicitly that we assume that asymptomatic infections in protected and unprotected people have the same \RR.}
 The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
@@
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
-\jd{Do we need ff.~s.?} This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+We note that the equivalence between protection against symptoms $\epsilon_s$ and fraction asymptomatic $p$ relies on our assumption that immunity does not provide protection against transmission.
 Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}C).
@@
 The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
-In general, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms, given that asymptomatically infected individuals can transmit better than symptomatically infected individuals (due to either behavioral and intervention effects).
-\jd{Be more clear what comparison we are making here. If we are comparing immunity to symptoms with immunity to infection, this s.~can be sharper and clearer. If not, what are we comparing?}
+In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms.
 

### 2022-06-06 main.tex
working before boarding

@@
 	\textbf\newline{
-		Intermediate levels of symptomaticity can lead to the worst population-level outcomes
+		Intermediate levels of asymptomaticit transmission can lead to the worst population-level outcomes
 	}
@@
 \\
-Joshua Weitz\textsuperscript{1,2,3},
+Joshua S. Weitz\textsuperscript{1,2,3},
 Sang Woo Park\textsuperscript{4},
@@
 SARS-CoV-2 has had devastating effects at the population level.
-However, many individuals have mild cases, making it harder to estimate the magnitude of spread and fatality rate. 
-The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \%--4\% \citep{rajgor2020many,VERITY2020669,yang2020early}.
-But many infections are never documented;
-the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for demographics similar to that of the United States \citep{levin2020assessing}.
+However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \citep{park2020time}. 
+The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \citep{rajgor2020many,VERITY2020669,yang2020early}.
+But many infections were never documented;
+the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to that of the United States \citep{levin2020assessing}.
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
@@
 Individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
-This means that the presence of asymptomatic infections has countervailing effects at the population level. 
+This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
 On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, and may therefore be more likely to infect others, since precautions are less likely to be taken.
-This means that the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and more likely to infect others.
+Altogether, the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
 
@@
 Asymptomatically and symptomatically infected individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
-Our key assumption is that symptomatically infected individuals take precautions (e.g., reducing contacts or mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
+Our key assumption is that symptomatically infected individuals take greater precautions than do asymptomatically infected individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
 we note that the parameter $\delta$ may also capture intervention measures that target symptomatically infected individuals, such as symptom-based isolation. 
@@
 \textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
-(Left) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
-(Middle) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-(Right) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+(Bottom left) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+(Bottom right) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
 We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$.
@@
 This relationship changes as $\delta$ increases.
-In particular, when $\delta > 1-\RR_a/\RR_s$ (in our case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic transmission rate (including behavioral response) is less than that the asymptomatic rate.
+In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic transmission rate (including behavioral response) is less than that the asymptomatic rate.
 For $\delta$ in this range, we can find a critical level of asymptomatic proportion, $p_c$:
@@
 although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
-In contrast, when $\delta$ is small enough that $\RR_s\geq\RR_a$, then total fatalities decrease with $p$ because the IFR ($(1-p)f$) decreases.
+In contrast, when $\delta$ is small enough such that $\RR_s\geq\RR_a$, then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
 
@@
 We therefore consider two additional of mathematical models with increasing complexities to answer a more general question:
-does intermediate amount of subclinical (including both asymptomatic and presymptomatic) transmission lead to peak fatalities?
-First, we analyze a model with presymptomatic transmission alone (Supplementary Figure S1).
+does intermediate amount of subclinical (including both asymptomatic and presymptomatic) transmission lead to a peak in fatalities?
+First, we analyze a modelin which there is a fraction $\theta$ of transmission occurs presymptomatically and all individuals eventually transition to a symptomatic phase (Supplementary Figure S1).
 When the reproduction number is fixed, increasing the proportion of presymptomatic transmission $\theta$ causes larger outbreaks, and therefore more deaths, in the presence of symptom-responsive transmission reduction ($\delta > 0$).
-However, standard compartmental models implicitly assume that the fatality rate $f$ is independent of the amount of presymptomatic transmission---in the extreme case where all transmission happens presymptomatically, infected individuals no longer shed infectious virus after symptom onset, meaning that they should be less likely to die from infection.
+However, standard compartmental models implicitly assume that the fatality rate $f$ is independent of the amount of presymptomatic transmission---in the extreme case where all transmission happens presymptomatically, we assume that the case fatality rate is far lower than in the limit when none of the transmission is in the presymptomatic phase.
 \jd{We need a bit more detail here.
@@
 Maybe I'm COVID-ignorant, but I want more info: many flu deaths are post-viral, and it's not super-clear how much of a role viral replication (or escape from deep lungs?) plays in fatality.}
-We therefore assume a tradeoff between the amount of presymptomatic transmission $\theta$ and the fatality rate $f$ that follows a power law: $f(\theta) = f_0 (1-\theta^a)$, which monotonically decreases from $f_0$ to 0.
+We therefore assume a tradeoff between the amount of presymptomatic transmission $\theta$ and the fatality rate $f$: $f(\theta) = f_0 (1-\theta^a)$, which monotonically decreases from $f_0$ to 0; the nonlinearity of the tradeoff is characterized by the power law exponent $a$.
 In the simple case where the fatality rate decreases linearly with $\theta$, the relationship between the amount of presymptomatic transmission $\theta$ and total fatalities is equivalent to that between the amount of asymptomatic transmission and total fatalities in the original model (\fref{base}).
-As we increase the exponent $a$, we obtain a more realistic tradeoff curve such that fatality rate $f$ remains roughly constant for low to intermediate values of $\theta$ and suddenly decreases to $0$ as $\theta$ approaches 1;
+As we increase the exponent $a$, we obtain a nonlinear tradeoff curve where fatality rate $f$ remains roughly constant for low to intermediate values of $\theta$ and suddenly decreases to $0$ as $\theta$ approaches 1;
 in these cases, peak fatalities occur at intermediate levels of presymptomatic transmission for even lower values of $\delta=0.6$, which now represents reduction in transmission after symptom onset.
@@
 We then extend our model to consider the effects of generalized \textit{subclinical} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S2).
-In particular, we fix the reproduction number of symptomatic individuals and calculate the proportion of deceased individuals for given values of proportion of total subclinical transmission and proportion of subclinical transmission that is caused by presymptomatic transmission.
-This model generalizes two earlier models:
-when all subclinical transmission is caused by asymptomatic transmission, we have the first model (\fref{base});
-and when all subclinical transmission is caused by presymptomatic transmission, we have the second model (Supplementary Figure S1).
-We find a wide variety of scenarios for which peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effect, $\delta > 0.6$ (Supplementary Figure S2);
+In particular, we fix the reproduction number of symptomatic individuals and calculate the proportion of fatalities as a function of the proportion of total subclinical transmission and the proportion of subclinical transmission that is caused by presymptomatic transmission.
+This generalized model extends the models proposed in \fref{base}, where all subclinical transmission is caused by asymptomatic transmission, and the model shown in Supplementary Figure S1 where all subclinical transmission is caused by presymptomatic transmission.
+Using the generalized subclinical transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2);
 this pattern is robust even to the absence of a tradeoff between the amount of presymptomatic transmission and the fatality rate.
@@
 
-We now apply our framework to understand the impact of immunity on the severity of the disease by dividing the population into two groups: immunologically naive and protected.
+We now apply our framework to understand the impact of immunity on total fatalities at the population scale by dividing the population into two groups: immunologically naive and protected.
 The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
@@
 Immunity may provide protection against infection, symptoms, or deaths.
-The dynamics of immunologically naive individuals are described in (\fref{base}).
+The dynamics of immunologically naive individuals are described in \fref{base}.
 (Bottom) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
@@
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
-The impact of protection against symptoms $\epsilon_s$ is exactly equivalent to changing fraction asymptomatic $p$ for the protected population:
-the fatality curves move left as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
+The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population:
+the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.

### 2022-06-08 main.tex
minor consensing

@@
 Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
-There are two implications for the discrepancy in the estimates of the proportion of presymptomatic transmission---first, a low proportion of presymptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) relative to the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of presymptomatic transmission from 20\% to 60\%.
-\jd{We need to be a bit more nuanced here: we get the low value by focusing on just a couple of weeks, so we're sort of suggesting a counter-factual where nobody notices COVID for a long time.}
-\swp{How's this?}
+There are two implications for the discrepancy in the estimates of the proportion of presymptomatic transmission---first, a low proportion of presymptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of presymptomatic transmission from 20\% to 60\%.
 

### 2022-06-11 figure_evidence.R
new figure and a posterior rds file

@@ -1,0 +1,78 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(gridExtra)
+library(shellpipes)
+
+## data from NBA
+data <- read.csv("indiv_data.csv") %>%
+  mutate(
+    symptomatic=factor(symptomatic, 
+                       labels=c("Asymptomatic", "Symptomatic"))
+  )
+
+## posterior from diamond
+posterior <- readRDS("primary.rds")
+
+gamma_p <- 1/2.1
+gamma_s <- 1/2.9
+gamma_a <- 1/5
+
+theta_a <- posterior$output$theta_a$value
+theta_p <- posterior$output$theta_p$value
+
+R_s <- 1/gamma_s + theta_p * 1/gamma_p
+R_a <- theta_a * 1/gamma_a
+
+Rratio <- data.frame(
+  value=R_a/R_s
+)
+
+chi <- posterior$output$chi
+
+g1 <- ggplot(chi) +
+  geom_histogram(aes(value, y=..density..), color="black", fill=NA) +
+  scale_x_continuous("Proportion of asymptomatic infection", limits=c(0, 1)) +
+  scale_y_continuous("Density") +
+  ggtitle("A. Diamond Princess cruise ship") +
+  theme(
+    panel.background = element_blank(),
+    legend.position = "none",
+    strip.background = element_blank(),
+    panel.grid = element_blank()
+  )
+
+g2 <- ggplot(Rratio) +
+  geom_histogram(aes(value, y=..density..), color="black", fill=NA) +
+  scale_x_continuous("Relative infectiousness of asymptomatic infection") +
+  scale_y_continuous("Density") +
+  ggtitle("B. Diamond Princess cruise ship") +
+  theme(
+    panel.background = element_blank(),
+    legend.position = "none",
+    strip.background = element_blank(),
+    panel.grid = element_blank()
+  )
+
+gcomb1 <- ggarrange(g1, g2, ncol=1, draw=FALSE)
+
+g3 <- ggplot(data) +
+  geom_point(aes(t, y, col=factor(symptomatic))) +
+  # geom_line(aes(t, y, col=factor(symptomatic), group=id), alpha=0.4) +
+  geom_smooth(aes(t, y, col=factor(symptomatic), fill=factor(symptomatic))) +
+  scale_x_continuous("Time since peak viral load") +
+  scale_y_reverse("Ct values") +
[…]

### 2022-06-13 main.tex
Update main.tex
asymptomatic is a better word than asymptomaticit, probably should mention epidemic in the title.
some more

@@
 	\textbf\newline{
-		Intermediate levels of asymptomaticit transmission can lead to the worst population-level outcomes
+		Intermediate levels of asymptomatic transmission can lead to the worst population-level epidemic outcomes
 	}
@@
 \\
-Joshua S. Weitz\textsuperscript{1,2,3},
-Sang Woo Park\textsuperscript{4},
-Jonathan Dushoff\textsuperscript{5,6,7}
+Sang Woo Park\textsuperscript{1},
+Jonathan Dushoff\textsuperscript{2,3,4},
+Joshua S. Weitz\textsuperscript{5,6,7}
 \\
 \bigskip
-\textbf{1} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
 \\
-\textbf{2} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{3} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
+\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{4} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
+\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{5} Department of Biology, McMaster University, Hamilton, ON, Canada
+\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
 \\
-\textbf{6} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
 \\
-\textbf{7} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
 \\
@@
 
-Individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{figure_evidence.ggp.pdf}
+\caption{
+\textbf{Available data on asymptomatic transmissibility of SARS-CoV-2.}
+(A) Previous estimates of the proportion of asymptomatic infections from the Diamond Princess Cruise Ship.
+(B) Previous estimates of the ratio $\theta_a$ of the transmisson rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship.
+Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an averae of 2.9 days, followed by a presymptomatic stage with an average of 2.1 days. 
+Aymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an averae of 5 days.
+Both estimates are publicly available from \cite{emery2020}.
+(C) Viral load trajectory data from National Basketball Association (NBA).
+Points represent each Ct measurement.
+Lines and shaded areas represent the smooth trajectories estimated via LOESS and the associated 95\% confidence intervals.
+Data are publicly available from \cite{Kissler2020}.
+}
+\label{fig:evidence}
+\end{figure}
+
+Early in the pandemic, a COVID-19 outbreak on the Diamond Princess Cruise Ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
+the outbreak occurred among 3711 passengers and crew, of which 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
+\cite{emery2020} estimated 74\% (95\% CI: 70\%--78\%) infections on the cruise ship to be asymptomatic (\fref{evidence}A), with about a half of total infections undetected;
+an earlier analysis of the same outbreak estimated a much lower asymptomatic proportion (17.9\%; 95\% CI: 15.5\%--20.2\%) but neglected undetected cases \citep{mizumoto2020estimating}.
+\cite{emery2020} found that the relative transmission rate of asymptomatic individuals were largely unidentifiable (\fref{evidence}B) but they were still able to rule out low transmissibility (0\%--25\%) as they require unrealistically high reproduction number of symptomatic individuals.
+Other modeling studies have also typically assumed low values for the ratio of the reproduction numbers between asymptomatic and symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
+Similarities in viral load trajectories of asymptomatic and symptomatic individuals also provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \cite{Kissler2020});
+however, differences between total viral load and infectious viral load add uncertainties to how well asymptomatic individuals can transmit.
+
+Despite large uncertainties in asymptomatic transmissibility, it is clear that individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
 This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
 On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and more likely to infect others.
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and more likely to infect others;
+asymptomatic SARS-CoV-2 infections present additional challenges due to the possibility of long COVID-19 \citep{xie2022long}.
 Altogether, the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
@@
 
-High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic---in practice, $\delta$ cannot be greater than the amount of post-symptomatic transmission.
+High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic.
+For this particular model, $\delta$ cannot be greater than the amount of post-symptomatic transmission because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
 While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to intervention and behavioral effects already as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
@@
 
-We therefore consider two additional of mathematical models with increasing complexities to answer a more general question:
-does intermediate amount of subclinical (including both asymptomatic and presymptomatic) transmission lead to a peak in fatalities?
-First, we analyze a modelin which there is a fraction $\theta$ of transmission occurs presymptomatically and all individuals eventually transition to a symptomatic phase (Supplementary Figure S1).
-When the reproduction number is fixed, increasing the proportion of presymptomatic transmission $\theta$ causes larger outbreaks, and therefore more deaths, in the presence of symptom-responsive transmission reduction ($\delta > 0$).
-However, standard compartmental models implicitly assume that the fatality rate $f$ is independent of the amount of presymptomatic transmission---in the extreme case where all transmission happens presymptomatically, we assume that the case fatality rate is far lower than in the limit when none of the transmission is in the presymptomatic phase.
-\jd{We need a bit more detail here.
-I have no idea what proportion of COVID deaths are post-viral.
-Maybe I'm COVID-ignorant, but I want more info: many flu deaths are post-viral, and it's not super-clear how much of a role viral replication (or escape from deep lungs?) plays in fatality.}
-We therefore assume a tradeoff between the amount of presymptomatic transmission $\theta$ and the fatality rate $f$: $f(\theta) = f_0 (1-\theta^a)$, which monotonically decreases from $f_0$ to 0; the nonlinearity of the tradeoff is characterized by the power law exponent $a$.
-In the simple case where the fatality rate decreases linearly with $\theta$, the relationship between the amount of presymptomatic transmission $\theta$ and total fatalities is equivalent to that between the amount of asymptomatic transmission and total fatalities in the original model (\fref{base}).
-As we increase the exponent $a$, we obtain a nonlinear tradeoff curve where fatality rate $f$ remains roughly constant for low to intermediate values of $\theta$ and suddenly decreases to $0$ as $\theta$ approaches 1;
-in these cases, peak fatalities occur at intermediate levels of presymptomatic transmission for even lower values of $\delta=0.6$, which now represents reduction in transmission after symptom onset.
-
-We then extend our model to consider the effects of generalized \textit{subclinical} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S2).
-In particular, we fix the reproduction number of symptomatic individuals and calculate the proportion of fatalities as a function of the proportion of total subclinical transmission and the proportion of subclinical transmission that is caused by presymptomatic transmission.
-This generalized model extends the models proposed in \fref{base}, where all subclinical transmission is caused by asymptomatic transmission, and the model shown in Supplementary Figure S1 where all subclinical transmission is caused by presymptomatic transmission.
-Using the generalized subclinical transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of subclinical transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2);
-this pattern is robust even to the absence of a tradeoff between the amount of presymptomatic transmission and the fatality rate.
-One exception is the case discussed earlier in which all subclinical transmission is caused by presymptomatic transmission.
-Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of subclinical transmission.
+We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S1), to ask the following question:
+does intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
+For this model, we assume that $\delta$ decreases transmission only after symptom onset.
+We then fix the reproduction number of symptomatic individuals and calculate the proportion of fatalities as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by presymptomatic transmission.
+Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2).
+One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by presymptomatic transmission;
+in this case, total infections and fatalities are maximized when all transmission is caused by presymptomatic transmission.
+Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
 
@@
 In other words, we assume that asymptomatic infections in protected and unprotected people have the same reproduction numbers (and likewise for the symptomatic infections).
-In practice, both asymptomatic and symptomatic infections in protected people are less likey to transmit than their unprotected counterparts: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral laod), in which case symptomaticity can be a poor proxy for transmission.
+In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
-\swp{JD: see if you like what I did in this P, esp in terms of boosting.}
 
@@
 
+We now consider each protection scenarios separately.
 The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}A).
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
-The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population:
-the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
+The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population because protected individuals are less likely to develop symptoms:
+therefore, the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
@@
 First of all, behavioral and intervention effects must be large in order for the fatality to peak at intermediate levels of asymptomaticity.
-While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of subclinical transmission on population-level outcomes.
+While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of non-symptomatic transmission on population-level outcomes.
 \swp{what are some other limitations we should address?}
@@
 \end{equation}
-Then, the proportion of subclinical transmission $\phi$ is given by:
+Then, the proportion of non-symptomatic transmission $\phi$ is given by:
 \begin{equation}
@@
 Consistent with previous assumptions, we also assume that asymptomatic reproduction number is lower than that of the symptomatic reproduction number: $\RR_a = \rho \RR_s$ where $\rho = 0.75$.
-Then, for a given value of the proportion of subclinical transmission $\phi$ and proportion of subclinical transmission caused by the presymptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
+Then, for a given value of the proportion of non-symptomatic transmission $\phi$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
 More specifically:
@@
 
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{diagram_pre.pdf}
-\caption{
-\textbf{Schematic diagram and simulations of a model with presymptomatic transmission and symptom-responsive transmission reduction.}
-(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
-See Methods for model details.
-(Bottom) Total deaths as a function of the proportion of presymptomatic transmission $\theta$ across a wide range scenarios for $\delta$ and a tradeoff between the proportion of presymptomatic infections $\theta$ and fatality rates $f$.
-}
-\end{center}
-\end{figure}
-
-\pagebreak
-
 \begin{figure}[!ht]
@@
 See Methods for model details.
-(Bottom) Total deaths as a function of the proportion of subclinical transmission $\phi$ across a wide range scenarios for $\delta$ and a tradeoff between the proportion of presymptomatic infections $\theta$ and fatality rates $f$.
+(Bottom) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta$ (between 0\% and 100\%).
 }

### 2022-06-13 figure_evidence.R

@@ -14,20 +14,7 @@ data <- read.csv("indiv_data.csv") %>%
 ## posterior from diamond
 posterior <- readRDS("primary.rds")
 
-gamma_p <- 1/2.1
-gamma_s <- 1/2.9
-gamma_a <- 1/5
-
-theta_a <- posterior$output$theta_a$value
-theta_p <- posterior$output$theta_p$value
-
-R_s <- 1/gamma_s + theta_p * 1/gamma_p
-R_a <- theta_a * 1/gamma_a
-
-Rratio <- data.frame(
-  value=R_a/R_s
-)
-
+theta_a <- posterior$output$theta_a
 chi <- posterior$output$chi
 
 g1 <- ggplot(chi) +
@@ -42,9 +29,9 @@ g1 <- ggplot(chi) +
     panel.grid = element_blank()
   )
 
-g2 <- ggplot(Rratio) +
+g2 <- ggplot(theta_a) +
   geom_histogram(aes(value, y=..density..), color="black", fill=NA) +
-  scale_x_continuous("Relative infectiousness of asymptomatic infection") +
+  scale_x_continuous("Relative asymptomatic transmission rate") +
   scale_y_continuous("Density") +
   ggtitle("B. Diamond Princess cruise ship") +
   theme(

### 2022-06-13 figure_sub.R

@@ -8,20 +8,21 @@ simulate_sub <- rdsRead()
 simulate_sub_max <- simulate_sub %>%
   group_by(sim) %>%
   filter(
-    time==max(time)
+    time==max(time),
+    power==-1
   ) %>%
   mutate(
-    power=paste0("f(theta)==f[0](1-theta^", power, ")"),
-    power=factor(power,
-                 levels=paste0("f(theta)==f[0](1-theta^", c(-1, 1, 2, 5), ")"),
-                 labels=c("No~tradeoff", paste0("f(theta)==f[0](1-theta^", c(1, 2, 5), ")"))),
+    # power=paste0("f(theta)==f[0](1-theta^", power, ")"),
+    # power=factor(power,
+    #              levels=paste0("f(theta)==f[0](1-theta^", c(-1, 1, 2, 5), ")"),
+    #              labels=c("No~tradeoff", paste0("f(theta)==f[0](1-theta^", c(1, 2, 5), ")"))),
     p_pre_ratio=factor(paste0(p_pre_ratio*100, "*\'%\'"),
                        levels=paste0(0:5*0.2*100, "*\'%\'"))
   )
 
 g1 <- ggplot(simulate_sub_max) +
   geom_line(aes(p_sub, D, col=factor(delta), group=factor(delta))) +
-  scale_x_continuous(expression(Proportion~subclinical~"transmission,"~phi), 
+  scale_x_continuous(expression(Proportion~"non-symptomatic"~"transmission,"~phi), 
                      breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
                      labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
                      expand=c(0, 0)) +
@@ -29,12 +30,11 @@ g1 <- ggplot(simulate_sub_max) +
                      limits=c(0, 0.0103),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(delta)) +
-  facet_grid(p_pre_ratio~power, labeller = label_parsed) +
+  facet_wrap(~p_pre_ratio, labeller = label_parsed) +
   theme(
     panel.grid = element_blank(),
     strip.background = element_blank(),
     legend.position = "right"
   )
 
-saveGG(g1, width=8, height=8)
-
+saveGG(g1, width=6, height=4)

### 2022-06-20 diagram_sub.tex
some more
changes

@@
 
-\begin{tikzpicture}[node distance=3.5cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+\begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
     \node [int, ultra thick] (S)             {$S$};
@@
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [below=15cm of S, anchor=west] (pic) {\includegraphics{figure_sub.ggp.pdf}};
+    \node [below=12cm of S, anchor=west] (pic) {\includegraphics{figure_sub.ggp.pdf}};
 \end{tikzpicture}

### 2022-06-20 main.tex
changes

@@
 (A) Previous estimates of the proportion of asymptomatic infections from the Diamond Princess Cruise Ship.
-(B) Previous estimates of the ratio $\theta_a$ of the transmisson rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship.
-Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an averae of 2.9 days, followed by a presymptomatic stage with an average of 2.1 days. 
-Aymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an averae of 5 days.
+(B) Previous estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship.
+Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a presymptomatic stage with an average of 2.1 days. 
+Asymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an average of 5 days.
 Both estimates are publicly available from \cite{emery2020}.
@@
 
-We now consider each protection scenarios separately.
+We now consider each protection effect separately---we will consider joint effects later on.
 The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}A).
@@
 First, we consider a scenario in which immunity only provides protection against symptoms, $\epsilon_s = 0.4$ (\fref{variant}A--C).
-In this case, protection against symptoms allows new variants to spread faster, resulting in larger outbreaks (\fref{variant}B).
+In this case, protection against symptoms allows new variants to spread faster by increasing the amount of asymptomatic infections, resulting in larger outbreaks (\fref{variant}B).
 Although the milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
-The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time.
+The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic.
 In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms.
@@
 
-In summary, using a simplified model we have shown that asymptomatic infections can represent a double-edged sword by providing a better outcome for some individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
-Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variants:
-while protection against symptoms protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
+In summary, using a series of simplified models, we have shown that asymptomatic infections (more generally, non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
+For example, while protection against symptoms (but not against transmission) protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level \swp{JSW: what's the Koelle and Lopman paper about vaccination that you were thinking?}.
 
@@
 This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
-even though the Omicron variant is probably milder than XXX \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was high \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
+even though the Omicron variant is probably milder than the Delta variant \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
 
 There are several limitations to our analysis.
-First of all, behavioral and intervention effects must be large in order for the fatality to peak at intermediate levels of asymptomaticity.
-While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of non-symptomatic transmission on population-level outcomes.
-\swp{what are some other limitations we should address?}
-\jd{Something about estimating protection against different endpoints. We should also highlight the importance of doing this estimation.}
+First of all, behavioral and intervention effects must be sufficiently large in order for the fatality to peak at intermediate levels of asymptomaticity.
+While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of non-symptomatic transmission on population-level outcomes using parameters informed by the SARS-CoV-2 outbreak.
+We also neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
+In particular, if immunity provides stronger protection against transmission among asymptomatically infected individuals (than among symptomatically infected individuals), asymptomatic infections will not necessarily make population-level outcomes worse.
+However, there is considerable uncertainty in how symptomaticity and transmissibility are correlated among protected people because symptoms can indicate strong immune response, rather than high viral load.
+Estimating protection against different endpoints (e.g., infection, symptom, death, and transmission) can help narrow this uncertainty.
+Finally, we assumed that asymptomatic and symptomatic individuals are infected for the same amount of time.
+On one hand, earlier analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \citep{Kissler2020}.
+On the other hand, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
+The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \citep{park2020time,harris2022time};
+despite these limitations, our qualitative predictions on the final size of the epidemic and total fatalities are likely robust.
 
 SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms. 
-Although mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and in the event that individuals are at risk for severe outcomes.
-As more variants continue to emerge, updating vaccines to prevent infections, and not just diseases, will be critical to controlling the course of the pandemic.
-\jd{This is complicated.
-Maybe regular waves of mild SC2 are a good way to control morbidity in the medium term.
-There is also an argument to make about symptom-focused responses and evolution toward mildness.}
+% Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
+Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via asymptomatic testing programs \citep{gibson2022surv}, mask-wearing indoors and in crowded environments \citep{howard2021ev}, and through improvements in ventilation \citep{wang2021airborne}) while improving the treatment of symptomatic cases.
+Given the link between age and asymptomatic infections, interventions may also consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
+As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections) on preventing infections, and not just diseases, will be critical to controlling the course of the pandemic.
 
@@
 
-\subsubsection*{Models without immunity}
+\subsubsection*{Epidemic models with asymptomatic infection and transmission in the absence of immunity}
 
-First, we consider a simple, compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
+First, we consider a compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
 The model dynamics are as follows:
@@
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
-Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\mu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$.
+Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\mu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
+Under this parameterization, we have symptomatic and asymptomatic reproduction numbers of $\RR_s = 4$ and $\RR_a = 3$.
 
-We then consider a model with presymptomatic transmission (and without asymptomatic transmission), where all infected individuals eventually develop symptoms:
-\begin{align}
-\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
-\dot{I}_p &= \mu E - \sigma I_p\\
-\dot{I}_s &= \sigma I_p -\gamma_s I_s\\
-\dot{R} &= (1-f) \gamma_s I_s \\
-\dot{D} &= f \gamma_s I_s
-\end{align}
-For this model, the presymptomatic $\RR_p$ and symptomatic $\RR_s$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$ and $\RR_s=\beta_s/\gamma_s$ in the absence of the behavioral effect; and the basic reproduction number is equal to the sum of the two.
-Then, the intrinsic proportion of presymptomatic transmission is given by: 
-\begin{equation}
-\theta = \frac{\RR_p}{\RR_p + \RR_s}.
-\end{equation}
-We assume there is a tradeoff between the proportion of presymptomatic transmission and fatality rate: 
-\begin{equation}
-f(\theta) = f_0 (1-\theta^a),
-\end{equation}
-where $f_0=0.01$ represents the baseline fatality rate and the exponent $a$ is varied between 1 and 5.
-Throughout simulations, we assume $\mathcal R_0 = 4$, $1/\sigma=2\,\mathrm{days}$, and $1/\gamma_s=3\,\mathrm{days}$.
-All other parameters are same as before.
-
-Finally, we combine both models to include both presymptomatic and asymptomatic transmission:
+We then extend the first model to include both presymptomatic and asymptomatic transmission:
 \begin{align}
@@
 \end{align}
-For this model, the reproduction number of individuals who will eventually develop symptoms is equal to: $\RR_p + \RR_s$;
+For this model, the presymptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
+these reproduction numbers represent the average number of secondary cases caused by an infected individual in each compartment.
+Then, the reproduction number of individuals who will eventually develop symptoms is equal to: $\RR_p + \RR_s$;
 similarly, the reproduction number of individuals who remain asymptomatic is equal to: $\RR_p + \RR_a$.
@@
 
-\subsubsection*{Model with immunity}
+\subsubsection*{Epidemic models with asymptomatic infection and transmission in the presence of immunity}
 
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{diagram_sub.pdf}
+\includegraphics[width=0.8\textwidth]{diagram_sub.pdf}
 \caption{

### 2022-06-29 main.tex
change
commit before dinner

@@
 \caption{
-\textbf{Available data on asymptomatic transmissibility of SARS-CoV-2.}
-(A) Previous estimates of the proportion of asymptomatic infections from the Diamond Princess Cruise Ship.
-(B) Previous estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship.
+\textbf{Asymptomatic transmissibility of SARS-CoV-2.}
+(A) Posterior estimates of the proportion of asymptomatic infections from the Diamond Princess Cruise Ship \cite{emery2020}.
+(B) Posterior estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship \cite{emery2020}.
+Bar charts represent the posterior distributions.
 Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a presymptomatic stage with an average of 2.1 days. 
 Asymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an average of 5 days.
-Both estimates are publicly available from \cite{emery2020}.
-(C) Viral load trajectory data from National Basketball Association (NBA).
+Both estimates are publicly available with further details in \cite{emery2020}.
+(C) Viral load trajectory data from players, staff, and vendors of the National Basketball Association (NBA).
 Points represent each Ct measurement.
 Lines and shaded areas represent the smooth trajectories estimated via LOESS and the associated 95\% confidence intervals.
-Data are publicly available from \cite{Kissler2020}.
+Data are publicly available in \cite{Kissler2020}.
 }
@@
 the outbreak occurred among 3711 passengers and crew, of which 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
-\cite{emery2020} estimated 74\% (95\% CI: 70\%--78\%) infections on the cruise ship to be asymptomatic (\fref{evidence}A), with about a half of total infections undetected;
-an earlier analysis of the same outbreak estimated a much lower asymptomatic proportion (17.9\%; 95\% CI: 15.5\%--20.2\%) but neglected undetected cases \citep{mizumoto2020estimating}.
-\cite{emery2020} found that the relative transmission rate of asymptomatic individuals were largely unidentifiable (\fref{evidence}B) but they were still able to rule out low transmissibility (0\%--25\%) as they require unrealistically high reproduction number of symptomatic individuals.
-Other modeling studies have also typically assumed low values for the ratio of the reproduction numbers between asymptomatic and symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
-Similarities in viral load trajectories of asymptomatic and symptomatic individuals also provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \cite{Kissler2020});
-however, differences between total viral load and infectious viral load add uncertainties to how well asymptomatic individuals can transmit.
+Early response work estimated that 17.9\% (95\% CI: 15.5\%--20.2\%) of confirmed infections on the Diamond Princess Cruise Ship were asymptomatic; but such estimates neglected undetected cases \citep{mizumoto2020estimating}.
+In contrast, re-analysis estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
+The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissability (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out becase it required unrealistically high transmissability for symptomatically infected individuals (\fref{evidence}B).
+Modeling studies have typically assumed that transmissability is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
+Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \cite{Kissler2020});
+however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals.
 
-Despite large uncertainties in asymptomatic transmissibility, it is clear that individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
+Despite quantitative uncertainties in asymptomatic transmissibility, it is clear that individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
 This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
@@
 To explore this idea, we propose a simple epidemic model,
-in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively (\fref{base}).  
+in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
 Asymptomatically infected individuals always recover, whereas a fraction $f$ of symptomatically infected individuals die.
@@
 For our main simulations, we assume that asymptomatically infected individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
-We assume that asymptomatic individuals do not die, and evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
+We evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
 \begin{figure}[!ht]
@@
 
-\fref{base} shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
-In the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number:
+\fref{base}B--C shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
+For this model, the basic reproduction number is given by:
 \begin{equation}
@@
 \end{equation}
-where $\RR_s = \beta_s/\gamma_s$ and $\RR_a = \beta_a/\gamma_a$ represent the reproduction numbers of asymptomatic and symptomatic individuals.
+where $\RR_s = \beta_s/\gamma_s$ and $\RR_a = \beta_a/\gamma_a$ represent the reproduction numbers of asymptomatic and symptomatic individuals (i.e., the average number of secondary infections caused by asymptomatic and symptomatic individuals);
+therefore, in the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number.
 This relationship changes as $\delta$ increases.
@@
 \end{equation}
-such that $p>p_c$ is required for an outbreak.
+such that $p>p_c$ is required for an outbreak (see threshold effects for large values of $\delta$ in \fref{base}B).
 
-When behavioral protection is high, the effect of asymptomatic proportion on fatalities shows countervailing effects of individual-level protection and population-level risk.
+When behavioral protection is high, the effect of asymptomatic proportion on fatalities shows countervailing effects of individual-level protection and population-level risk (\fref{base}C).
 For high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
@@
 We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S1), to ask the following question:
-does intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
+does an intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
 For this model, we assume that $\delta$ decreases transmission only after symptom onset.
-We then fix the reproduction number of symptomatic individuals and calculate the proportion of fatalities as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by presymptomatic transmission.
+We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by presymptomatic transmission.
 Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2).
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{diagram_sub.pdf}
+\includegraphics[width=0.75\textwidth]{diagram_sub.pdf}
 \caption{

### 2022-07-06 main.tex
updates

@@
 In other words, we assume that asymptomatic infections in protected and unprotected people have the same reproduction numbers (and likewise for the symptomatic infections).
-In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
+In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \citep{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
@@
 
-We now consider each protection effect separately---we will consider joint effects later on.
+We consider each protection effect---$\epsilon_i$, $\epsilon_s$, and $\epsilon_d$---separately and consider joint effects later on.
 The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}A).
@@
 Finally, we use our framework to understand the impact of behavioral effects on invading variants (\fref{variant}).
-In doing so, we first simulate the dynamics of a wildtype variant for 1 year using our base model (\fref{base}).
-We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the wildtype variant.
+In doing so, we first simulate the dynamics of a wildtype variant for 1 year using our base model using identical parameters as in \fref{base}.
+We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
 We consider two types of variants (which are simulated separately): one with the same severity $p$ (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
@@
 The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic.
-In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms.
+In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms but not against infection or transmission.
 
@@
 
-In summary, using a series of simplified models, we have shown that asymptomatic infections (more generally, non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+\section*{Discussion}
+
+In summary, using a series of simplified models, we have shown that asymptomatic infections (including the more general category of non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
@@
 
-Our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
-Despite moderate levels of vaccine effectiveness against symptomatic and severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \citep{pearson2021omicron}.
-This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron};
-even though the Omicron variant is probably milder than the Delta variant \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
+The outcomes in our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
+Despite moderate levels of vaccine effectiveness against symptomatic and reduced levels of severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \citep{pearson2021omicron}.
+This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron}.
+Moreover, even though the Omicron variant is probably milder than the Delta variant \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
 
 There are several limitations to our analysis.
-First of all, behavioral and intervention effects must be sufficiently large in order for the fatality to peak at intermediate levels of asymptomaticity.
-While we are able to generalize our results using more realistic models, incorporating both presymptomatic and asymptomatic transmission, the transmission rate needs to be reduced by at least 60\% after symptom onset for us to see the nonlinear effects of non-symptomatic transmission on population-level outcomes using parameters informed by the SARS-CoV-2 outbreak.
-We also neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
+First, while we are able to generalize the model to include both presymptomatic and asymptomatic transmission, behavioral and intervention effects must be sufficiently large in order for the fatality to peak at intermediate levels of asymptomaticity (e.g., a reduction in transmission rate of typically 60\% or more for the given choice of epidemiological parameters).
+Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity. 
+However, we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
 In particular, if immunity provides stronger protection against transmission among asymptomatically infected individuals (than among symptomatically infected individuals), asymptomatic infections will not necessarily make population-level outcomes worse.
@@
 Finally, we assumed that asymptomatic and symptomatic individuals are infected for the same amount of time.
-On one hand, earlier analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \citep{Kissler2020}.
-On the other hand, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
+Analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \citep{Kissler2020};
+however, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
 The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \citep{park2020time,harris2022time};
-despite these limitations, our qualitative predictions on the final size of the epidemic and total fatalities are likely robust.
+despite these limitations, we expect that our qualitative predictions on the final size of the epidemic and total fatalities are robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
 
@@
 % Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
-Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via asymptomatic testing programs \citep{gibson2022surv}, mask-wearing indoors and in crowded environments \citep{howard2021ev}, and through improvements in ventilation \citep{wang2021airborne}) while improving the treatment of symptomatic cases.
+Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \citep{chande2020real,sinclair2021pairing}, asymptomatic testing programs \citep{mina2021covid,gibson2022surv}, mask-wearing indoors and in crowded environments \citep{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \citep{allen2021indoor,wang2021airborne}) while improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
 Given the link between age and asymptomatic infections, interventions may also consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).

### 2022-07-07 diagram_base.tex
working before boarding
done?

@@
 
-\begin{tikzpicture}[node distance=3.5cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+\begin{tikzpicture}[node distance=3.2cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+    \tikzstyle{every node}=[font=\Large]
     \node [int, ultra thick] (S)             {$S$};
@@
     \node [int, right=of Is, ultra thick] (D)             {$D$};
+    \node [left=8cm of Ia] (A)             {\huge \textbf{A}};
     \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
@@
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [right=8.5cm of E] (pic) {\includegraphics{figure_base.ggp.pdf}};
+    \node [below=1cm of Is] (pic) {\includegraphics{figure_base.ggp.pdf}};
 \end{tikzpicture}

### 2022-07-07 diagram_immune.tex
done?

@@
 
-\begin{tikzpicture}[node distance=3.5cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+\begin{tikzpicture}[node distance=4cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+    \tikzstyle{every node}=[font=\large]
     \node [int, ultra thick] (S)             {$S_p$};
@@
     \node [int, right=of Is, ultra thick] (D)             {$D_p$};
+    \node [left=8cm of Ia] (A)             {\huge \textbf{A}};
     \path[->, ultra thick] (S) edge node [above] {${\color{red}(1-  \epsilon_i)} \lambda(t)$} (E);
@@
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [below=8.5cm of S, anchor=west] (pic) {\includegraphics{figure_immune.ggp.pdf}};
+    \node [below=1cm of Is] (pic) {\includegraphics{figure_immune.ggp.pdf}};
 \end{tikzpicture}

### 2022-07-07 main.tex
done?

@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission can lead to the worst population-level epidemic outcomes
+		Intermediate levels of asymptomatic transmission can lead to highest levels of epidemic fatalities
 	}
@@
 \textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
-(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
-(Bottom left) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-(Bottom right) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+(B) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+(C) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
 We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$.
@@
 \textbf{Schematic diagram and simulations of a model with symptom-responsive transmission reduction and immunity.}
-(Top) The subscript $p$ represents protected individuals. 
+(A) The subscript $p$ represents protected individuals. 
 Immunity may provide protection against infection, symptoms, or deaths.
 The dynamics of immunologically naive individuals are described in \fref{base}.
-(Bottom) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$.
+(B--D) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$ (B), symptoms $\epsilon_s$ (C), and deaths $\epsilon_d$ (D).
 We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
@@
 \begin{align}
-\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
+\dot{S} &= -\beta_p I_p -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
 \dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
@@
 similarly, the reproduction number of individuals who remain asymptomatic is equal to: $\RR_p + \RR_a$.
-Since proportion $p$ of all infections is asymptomatic, the basic reproduction number is given by the weighted average of these two reproduction numbers: 
+Since the proportion $p$ of all infections is asymptomatic, the basic reproduction number is given by the weighted average of these two reproduction numbers: 
 \begin{equation}
@@
 where $y = (1/\phi - 1)/\eta + (1/\eta - 1)/\rho$.
+In order to keep the mean infectious period fixed, we assume $1/\sigma=2\,\mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\,\mathrm{days}$.
 All other parameters are same as before.
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{diagram_sub.pdf}
+\includegraphics[width=0.8\textwidth]{diagram_sub.pdf}
 \caption{
 \textbf{Schematic diagram and simulations of a model with presymptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
-(Top) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
-(Bottom) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta$ (between 0\% and 100\%).
+(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta$ (between 0\% and 100\%).
 }

### 2022-07-07 figure_base.R

@@ -21,6 +21,7 @@ g1 <- ggplot(simulate_base_max) +
                      limits=c(0, 1),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(delta)) +
+  ggtitle("B") +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.8, 0.4),
@@ -37,6 +38,7 @@ g2 <- ggplot(simulate_base_max) +
                      limits=c(0, 0.01),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(delta)) +
+  ggtitle("C") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -44,4 +46,4 @@ g2 <- ggplot(simulate_base_max) +
 
 gfinal <- ggarrange(g1, g2, nrow=1, draw=FALSE)
 
-saveGG(gfinal, width=8, height=3)
+saveGG(gfinal, width=8, height=3.2)

### 2022-07-07 figure_immune.R

@@ -16,7 +16,7 @@ g1 <- ggplot(simulate_immune_i_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d("Protection", option="B", end=0.9) +
-  ggtitle("Protection against\ninfection") +
+  ggtitle("B. Protection against\ninfection") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -32,7 +32,7 @@ g2 <- ggplot(simulate_immune_s_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d("Protection", option="B", end=0.9) +
-  ggtitle("Protection against\nsymptoms") +
+  ggtitle("C. Protection against\nsymptoms") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -48,7 +48,7 @@ g3 <- ggplot(simulate_immune_d_summarize) +
                      limits=c(0, ymax),
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(epsilon), option="B", end=0.9) +
-  ggtitle("Protection against\ndeaths") +
+  ggtitle("D. Protection against\ndeaths") +
   theme(
     panel.grid = element_blank(),
     legend.position = "right"

### 2022-07-07 figure_sub.R

@@ -31,6 +31,7 @@ g1 <- ggplot(simulate_sub_max) +
                      expand=c(0, 0)) +
   scale_color_viridis_d(expression(delta)) +
   facet_wrap(~p_pre_ratio, labeller = label_parsed) +
+  ggtitle("B") +
   theme(
     panel.grid = element_blank(),
     strip.background = element_blank(),

### 2022-07-07 main.tex
done

@@
 
-\section*{Discussion}
-
-In summary, using a series of simplified models, we have shown that asymptomatic infections (including the more general category of non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
-Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
-For example, while protection against symptoms (but not against transmission) protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level \swp{JSW: what's the Koelle and Lopman paper about vaccination that you were thinking?}.
-
 The outcomes in our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
@@
 despite these limitations, we expect that our qualitative predictions on the final size of the epidemic and total fatalities are robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
+
+In summary, using a series of simplified models, we have shown that asymptomatic infections (including the more general category of non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
+For example, while protection against symptoms (but not against transmission) protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
+Similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \citep{koelle2022changing}.
 

### 2022-07-11 main.tex
Still need to do supp

@@
 However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \citep{park2020time}. 
+\jd{This is an awkward place to cite ourselves (we do it again later) and it makes the sentence a bit awkward, too.
+Can we find a mainstream cite and say something more like “total number of infections”?}
 The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \citep{rajgor2020many,VERITY2020669,yang2020early}.
 But many infections were never documented;
-the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to that of the United States \citep{levin2020assessing}.
+the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \citep{levin2020assessing}.
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
@@
 Bar charts represent the posterior distributions.
-Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a presymptomatic stage with an average of 2.1 days. 
+Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a pre-symptomatic stage with an average of 2.1 days. 
 Asymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an average of 5 days.
@@
 Early in the pandemic, a COVID-19 outbreak on the Diamond Princess Cruise Ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
-the outbreak occurred among 3711 passengers and crew, of which 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
-Early response work estimated that 17.9\% (95\% CI: 15.5\%--20.2\%) of confirmed infections on the Diamond Princess Cruise Ship were asymptomatic; but such estimates neglected undetected cases \citep{mizumoto2020estimating}.
+the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
+\jd{Do we need this s.?} Early response work estimated that 17.9\% (95\% CI: 15.5\%--20.2\%) of confirmed infections on the Diamond Princess Cruise Ship were asymptomatic; but such estimates neglected undetected cases \citep{mizumoto2020estimating}.
 In contrast, re-analysis estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
-The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissability (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out becase it required unrealistically high transmissability for symptomatically infected individuals (\fref{evidence}B).
-Modeling studies have typically assumed that transmissability is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
+The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissibility (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out becase it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
+Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
 Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \cite{Kissler2020});
-however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals.
+however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \jd{Is there a citation?
+Can this be clarified?
+Anything about the role of symptoms (coughing, sneezing) in potentially spreading disease?}.
 
@@
 This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
-On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and fatality. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and more likely to infect others;
-asymptomatic SARS-CoV-2 infections present additional challenges due to the possibility of long COVID-19 \citep{xie2022long}.
+On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and death. 
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and relatively more likely to infect others;
+\jd{If this is about the possibility of \emph{transmission} via long COVID, we need to say so} asymptomatic SARS-CoV-2 infections present additional challenges due to the possibility of long COVID \citep{xie2022long}.
 Altogether, the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
@@
 (C) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$.
-We assume that $10^{-4}$ proportion of individuals are initially infected.
+We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
+\jd{Can we change $\mu$ to $\nu$ to follow convention? Don't feel strongly about $\nu$ but $\mu$ has an accepted meaning here.}
 }
@@
 This relationship changes as $\delta$ increases.
-In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic transmission rate (including behavioral response) is less than that the asymptomatic rate.
+In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic reproductive number (including behavioral response) is less than that the asymptomatic reproductive number.
+\jd{The ff.~is not quite correct: we need $\delta > 1-/\RR_s$ for $p_c>0$ (and we're also assuming $\RR_a>1$.}
 For $\delta$ in this range, we can find a critical level of asymptomatic proportion, $p_c$:
@@
 although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
-In contrast, when $\delta$ is small enough such that $\RR_s\geq\RR_a$, then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
+In contrast, when $\delta$ is small enough such that $\RR_s\geq\RR_a$, \jd{Note that this is the same as the $\delta > 0.25 $ criterion above} then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
 
 High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic.
-For this particular model, $\delta$ cannot be greater than the amount of post-symptomatic transmission because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
-While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of them were likely subject to intervention and behavioral effects already as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
+For this particular model, it does not make biological sense for $\delta$ to be greater than the amount of post-symptomatic transmission, because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
+While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of these were likely affected by intervention and behavioral effects, as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
 Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
-There are two implications for the discrepancy in the estimates of the proportion of presymptomatic transmission---first, a low proportion of presymptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of presymptomatic transmission from 20\% to 60\%.
+There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of pre-symptomatic transmission from 80\% to as low as 40\%.
 
-We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both presymptomatic and asymptomatic transmission (Supplementary Figure S1), to ask the following question:
+We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission (Supplementary Figure S1), to ask the following question:
 does an intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
 For this model, we assume that $\delta$ decreases transmission only after symptom onset.
-We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by presymptomatic transmission.
+We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission.
 Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2).
-One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by presymptomatic transmission;
-in this case, total infections and fatalities are maximized when all transmission is caused by presymptomatic transmission.
+One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission;
+in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission.
+\jd{What??
+If all transmission is pre-symptomatic, then we get no fatalities in our model.}
 Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
+\jd{This is continuing to seem tangled to me. Should we reconsider the approach?}
 
@@
 The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
-The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms $\epsilon_s$, and deaths $\epsilon_d$ (\fref{immune}).
-For simplicity, we assume that the population is exactly split in half (50\% naive and 50\% protected) and mixes homogeneously.
+The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms (given infection) $\epsilon_s$, and deaths (given symptoms) $\epsilon_d$ (\fref{immune}).
+For simplicity, we assume that the population is split in half (50\% naive and 50\% protected) and mixes homogeneously.
 We also do not consider the separate effect of immunity on transmission (beyond the effect on infection).
@@
 In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \citep{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
+\jd{Should these caveats bother us more?
+Is this part of our model misleading?}
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
@@
 There are several limitations to our analysis.
-First, while we are able to generalize the model to include both presymptomatic and asymptomatic transmission, behavioral and intervention effects must be sufficiently large in order for the fatality to peak at intermediate levels of asymptomaticity (e.g., a reduction in transmission rate of typically 60\% or more for the given choice of epidemiological parameters).
-Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity. 
-However, we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
+First, while we are able to generalize the model to include both pre-symptomatic and asymptomatic transmission, behavioral and intervention effects must be relatively large in order for the fatality to peak at intermediate levels of asymptomaticity (typically requiring a reduction in transmission rate of 60\% or more for most of our chosen parameter sets).
+Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity,
+but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
 In particular, if immunity provides stronger protection against transmission among asymptomatically infected individuals (than among symptomatically infected individuals), asymptomatic infections will not necessarily make population-level outcomes worse.
@@
 The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \citep{park2020time,harris2022time};
-despite these limitations, we expect that our qualitative predictions on the final size of the epidemic and total fatalities are robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
+nonetheless, we expect that predictions on the final size of the epidemic and total fatalities will be robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
 
-In summary, using a series of simplified models, we have shown that asymptomatic infections (including the more general category of non-symptomatic transmissions) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generallly, non-symptomatic transmission) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
-For example, while protection against symptoms (but not against transmission) protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
+For example, while protection against symptoms unaccompanied by protection against transmission protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
 Similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \citep{koelle2022changing}.
@@
 
-We then extend the first model to include both presymptomatic and asymptomatic transmission:
+We then extend the first model to include both pre-symptomatic and asymptomatic transmission:
 \begin{align}
@@
 \end{align}
-For this model, the presymptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
+For this model, the pre-symptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
 these reproduction numbers represent the average number of secondary cases caused by an infected individual in each compartment.
@@
 Consistent with previous assumptions, we also assume that asymptomatic reproduction number is lower than that of the symptomatic reproduction number: $\RR_a = \rho \RR_s$ where $\rho = 0.75$.
-Then, for a given value of the proportion of non-symptomatic transmission $\phi$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
+Then, for a given value of the proportion of non-symptomatic transmission $\phi$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
 More specifically:
@@
 \caption{
-\textbf{Schematic diagram and simulations of a model with presymptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
-(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents presymptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+\textbf{Schematic diagram and simulations of a model with pre-symptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
+(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents pre-symptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
-(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the presymptomatic transmission, $\eta$ (between 0\% and 100\%).
+(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
 }

### 2022-07-12 diagram_immune.tex
done

@@
     \path[->, ultra thick] (S) edge node [above] {${\color{red}(1-  \epsilon_i)} \lambda(t)$} (E);
-    \path[->, ultra thick] (E) edge node [sloped, above] {$\big[1-{\color{red}(1-\epsilon_s)} (1-p)\big] \mu$} (Ia);
-    \path[->, ultra thick] (E) edge node [sloped, above] {${\color{red}(1-\epsilon_s)} (1-p) \mu$} (Is);
+    \path[->, ultra thick] (E) edge node [sloped, above] {$\big[1-{\color{red}(1-\epsilon_s)} (1-p)\big] \nu$} (Ia);
+    \path[->, ultra thick] (E) edge node [sloped, above] {${\color{red}(1-\epsilon_s)} (1-p) \nu$} (Is);
     \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);

### 2022-07-12 diagram_sub.tex
done?
done

@@
 \begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},]
+    \tikzstyle{every node}=[font=\Large]
     \node [int, ultra thick] (S)             {$S$};
@@
     \node [int, right=of Is, ultra thick] (D)             {$D$};
+    \node [left=10cm of Ia] (A)             {\huge \textbf{A}};
     \path[->, ultra thick] (S) edge node [above] {$\lambda(t)$} (E);
     \path[->, ultra thick] (E) edge node [above] {$\sigma$} (Ip);
-    \path[->, ultra thick] (Ip) edge node [sloped, above] {$p \mu$} (Ia);
-    \path[->, ultra thick] (Ip) edge node [sloped, above] {$(1-p) \mu$} (Is);
+    \path[->, ultra thick] (Ip) edge node [sloped, above] {$p \nu$} (Ia);
+    \path[->, ultra thick] (Ip) edge node [sloped, above] {$(1-p) \nu$} (Is);
     \path[->, ultra thick] (Ia) edge node [above] {$\gamma_a$} (R);
@@
     \path[->, ultra thick, dashed] (Is.west) edge [out=180, in=-90] node [sloped, above] {$(1-\delta) \beta_s$} (Ts.south);
-    \node [below=12cm of S, anchor=west] (pic) {\includegraphics{figure_sub.ggp.pdf}};
+    \node [below=9cm of S, anchor=west] (pic) {\includegraphics{figure_sub.ggp.pdf}};
 \end{tikzpicture}

### 2022-07-12 main.tex
done

@@
 SARS-CoV-2 has had devastating effects at the population level.
-However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \citep{park2020time}. 
-\jd{This is an awkward place to cite ourselves (we do it again later) and it makes the sentence a bit awkward, too.
-Can we find a mainstream cite and say something more like “total number of infections”?}
+However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \citep{nogrady2020data}.
 The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \citep{rajgor2020many,VERITY2020669,yang2020early}.
@@
 the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
-\jd{Do we need this s.?} Early response work estimated that 17.9\% (95\% CI: 15.5\%--20.2\%) of confirmed infections on the Diamond Princess Cruise Ship were asymptomatic; but such estimates neglected undetected cases \citep{mizumoto2020estimating}.
-In contrast, re-analysis estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
+It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}, but there has been a considerable heterogeneity in the estimates of asymptomatic prevalence in other settings \citep{alene2021magnitude}.
 The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissibility (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out becase it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
 Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
-Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \cite{Kissler2020});
-however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \jd{Is there a citation?
-Can this be clarified?
-Anything about the role of symptoms (coughing, sneezing) in potentially spreading disease?}.
+Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \citep{Kissler2020});
+however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
 
@@
 On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and relatively more likely to infect others;
-\jd{If this is about the possibility of \emph{transmission} via long COVID, we need to say so} asymptomatic SARS-CoV-2 infections present additional challenges due to the possibility of long COVID \citep{xie2022long}.
+asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \citep{xie2022long}.
 Altogether, the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
@@
 (C) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
-\jd{Can we change $\mu$ to $\nu$ to follow convention? Don't feel strongly about $\nu$ but $\mu$ has an accepted meaning here.}
+We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
 }
@@
 In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic reproductive number (including behavioral response) is less than that the asymptomatic reproductive number.
-\jd{The ff.~is not quite correct: we need $\delta > 1-/\RR_s$ for $p_c>0$ (and we're also assuming $\RR_a>1$.}
-For $\delta$ in this range, we can find a critical level of asymptomatic proportion, $p_c$:
+For high values of $\delta$, we can find a critical level of asymptomatic proportion, $p_c$:
 \begin{equation}
@@
 although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
-In contrast, when $\delta$ is small enough such that $\RR_s\geq\RR_a$, \jd{Note that this is the same as the $\delta > 0.25 $ criterion above} then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
+In contrast, when $\delta$ is small enough such that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
 
@@
 \jd{What??
-If all transmission is pre-symptomatic, then we get no fatalities in our model.}
+If all transmission is pre-symptomatic, then we get no fatalities in our model.} \swp{what do you mean?
+if all non-symptomatic transmission is caused by pre-symptomatic transmission, then everyone eventually develops symptoms...}
 Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
@@
 \jd{Should these caveats bother us more?
-Is this part of our model misleading?}
+Is this part of our model misleading?} \swp{I don't think they really bother us...
+transmission blocking should have qualitatively similar effects as infection blocking.
+If transmission rate is sufficiently reduced then we're good.
+Just like how we showed that protection against symptoms + infection is good.}
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
@@
 (B--D) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$ (B), symptoms $\epsilon_s$ (C), and deaths $\epsilon_d$ (D).
-We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\mu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
+We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
 We assume that $10^{-4}$ proportion of individuals are initially infected.
@@
 \dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
-\dot{I}_a &= p \mu E - \gamma_a I_a\\
-\dot{I}_s &= (1-p) \mu E -\gamma_s I_s\\
+\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \nu E\\
+\dot{I}_a &= p \nu E - \gamma_a I_a\\
+\dot{I}_s &= (1-p) \nu E -\gamma_s I_s\\
 \dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
@@
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
-Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\mu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
+Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
 Under this parameterization, we have symptomatic and asymptomatic reproduction numbers of $\RR_s = 4$ and $\RR_a = 3$.
@@
 \dot{S} &= -\beta_p I_p -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \mu E\\
-\dot{I}_p &= \mu E - \sigma I_p\\
+\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \nu E\\
+\dot{I}_p &= \nu E - \sigma I_p\\
 \dot{I}_a &= p \sigma I_p - \gamma_a I_a\\
@@
 \dot{S} &= -\lambda (t) S \\
-\dot{E} &= \lambda (t) S - \mu E\\
-\dot{I}_a &= p \mu E - \gamma_a I_a\\
-\dot{I}_s &= (1-p) \mu E -\gamma_s I_s\\
+\dot{E} &= \lambda (t) S - \nu E\\
+\dot{I}_a &= p \nu E - \gamma_a I_a\\
+\dot{I}_s &= (1-p) \nu E -\gamma_s I_s\\
 \dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
@@
 \dot{S}_p &= - (1-\epsilon_i) \lambda (t) S_p \\
-\dot{E}_p &= (1-\epsilon_i) \lambda (t) S_p - \mu E_p\\
-\dot{I}_{p, a} &= (1 - (1-\epsilon_s) (1-p)) \mu E_p - \gamma_a I_{p,a}\\
-\dot{I}_{p, s} &= (1-\epsilon_s) (1-p) \mu E_p -\gamma_s I_{p,s}\\
+\dot{E}_p &= (1-\epsilon_i) \lambda (t) S_p - \nu E_p\\
+\dot{I}_{p, a} &= (1 - (1-\epsilon_s) (1-p)) \nu E_p - \gamma_a I_{p,a}\\
+\dot{I}_{p, s} &= (1-\epsilon_s) (1-p) \nu E_p -\gamma_s I_{p,s}\\
 \dot{R}_p &= \gamma_a I_{p,a} + (1-(1-\epsilon_d) f) \gamma_s I_{p,s} \\

### 2022-07-14 main.tex
de-confuse self

@@
 Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2).
-One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission;
+One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-\jd{What??
-If all transmission is pre-symptomatic, then we get no fatalities in our model.} \swp{what do you mean?
-if all non-symptomatic transmission is caused by pre-symptomatic transmission, then everyone eventually develops symptoms...}
 Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
-\jd{This is continuing to seem tangled to me. Should we reconsider the approach?}
 

### 2022-07-19 main.tex
abstract

@@
 \end{flushleft}
+
+\section*{Abstract}
+
+Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
+Even though asymptomatic infections reduce the severity at the individual level, they can still make population-level outcomes worse by transmitting to others unknowingly.
+Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities in the presence of a strong reduction in symptomatic transmission rates, which can be caused by either behavioral changes or mitigation efforts.
+We further generalize this result to show that intermediate levels of non-symptomatic transmission (including both presymptomatic and asymptomatic transmission) can also lead to the highest levels of fatalities.
+Finally, we extend our framework to illustrate the importance of immunity profile in determining epidemic trajectories of future variants.
+
+\pagebreak
 

### 2022-07-22 main.tex
changes to abstract

--- 2022-07-14 main.tex
@@
 \end{flushleft}
+
+\section*{Abstract}
+
+Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
+Even though asymptomatic infections reduce severity at the individual level, they can still make population-level outcomes worse if asymptomatic individuals transmit more because they are unaware that they are infected.
+Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when this increase in transmission, due either to individual behavior or mitigation efforts, is strong.
+We generalize this result to include presymptomatic transmission, and show that in this case as well intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
+Finally, we extend our framework to illustrate the importance of immunity profile in determining epidemic trajectories of future variants.
+
+\pagebreak
 

### 2022-07-23 main.tex
Wasting time
commit

--- 2022-07-19 main.tex
@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission can lead to highest levels of epidemic fatalities
+		Intermediate levels of asymptomatic transmission can lead to the highest levels of epidemic fatalities
 	}
@@
 Data are publicly available in \cite{Kissler2020}.
+(D) Inferred proportion of infections that are subclinical for each age group using case data from six countries (China, Italy, Japan, Singapore, South Korea, and Canada) \cite{davies2020}.
 }
@@
 however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
+We note asymptomaticity is expected to be more heterogeneous in real outbreaks.
+For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020};
+differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
+For now, we primarily focus on a homogeneous population and return to the age effect in the discussion.
 
@@
 asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \citep{xie2022long}.
-Altogether, the prevalence of asymptomatic infections can paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
+Altogether, the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
 
@@
 For high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
-although less individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
+although fewer individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
 In contrast, when $\delta$ is small enough such that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
@@
 While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of these were likely affected by intervention and behavioral effects, as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
-Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission can be as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
+Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission could have been as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
 There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of pre-symptomatic transmission from 80\% to as low as 40\%.
@@
 We now apply our framework to understand the impact of immunity on total fatalities at the population scale by dividing the population into two groups: immunologically naive and protected.
+For simplicity, we do not distinguish whether the immunity is derived from natural infections or vaccines.
 The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
@@
 In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \citep{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
-\jd{Should these caveats bother us more?
-Is this part of our model misleading?} \swp{I don't think they really bother us...
-transmission blocking should have qualitatively similar effects as infection blocking.
-If transmission rate is sufficiently reduced then we're good.
-Just like how we showed that protection against symptoms + infection is good.}
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
@@
 We consider each protection effect---$\epsilon_i$, $\epsilon_s$, and $\epsilon_d$---separately and consider joint effects later on.
-The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}A).
+The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}B).
 We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
 The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population because protected individuals are less likely to develop symptoms:
-therefore, the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}B).
+therefore, the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}C).
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
@@
 We note that the equivalence between protection against symptoms $\epsilon_s$ and fraction asymptomatic $p$ relies on our assumption that immunity does not provide protection against transmission.
-Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}C).
+Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}D).
 
@@
 In particular, if immunity provides stronger protection against transmission among asymptomatically infected individuals (than among symptomatically infected individuals), asymptomatic infections will not necessarily make population-level outcomes worse.
-However, there is considerable uncertainty in how symptomaticity and transmissibility are correlated among protected people because symptoms can indicate strong immune response, rather than high viral load.
 Estimating protection against different endpoints (e.g., infection, symptom, death, and transmission) can help narrow this uncertainty.
@@
 nonetheless, we expect that predictions on the final size of the epidemic and total fatalities will be robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
+
+Even though we assumed a homogeneous population throughout, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity.
+For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make their individual-level outcomes better.
+However, asymptomatically infected individuals, especially among the younger population with high contact rates, can still transmit to other, potentially making the population-level outcomes worse than they would be if intervention measures were distributed differently.
+We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.
 
@@
 Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \citep{chande2020real,sinclair2021pairing}, asymptomatic testing programs \citep{mina2021covid,gibson2022surv}, mask-wearing indoors and in crowded environments \citep{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \citep{allen2021indoor,wang2021airborne}) while improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
-Given the link between age and asymptomatic infections, interventions may also consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
+Given the link between age and asymptomatic infections \citep{davies2020}, interventions may consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
+Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
 As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections) on preventing infections, and not just diseases, will be critical to controlling the course of the pandemic.

### 2022-07-23 figure_evidence.R

@@ -11,6 +11,13 @@ data <- read.csv("indiv_data.csv") %>%
                        labels=c("Asymptomatic", "Symptomatic"))
   )
 
+data_age <- data.frame(
+  age=c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70+"),
+  mean=c(0.29, 0.21, 0.27, 0.33, 0.4, 0.49, 0.63, 0.69),
+  lwr=c(0.18, 0.12, 0.18, 0.24, 0.28, 0.37, 0.49, 0.57),
+  upr=c(0.44, 0.31, 0.38, 0.43, 0.52, 0.6, 0.76, 0.82)
+)
+
 ## posterior from diamond
 posterior <- readRDS("primary.rds")
 
@@ -60,6 +67,17 @@ g3 <- ggplot(data) +
     panel.grid = element_blank()
   )
 
-gfinal <- arrangeGrob(gcomb1, g3, nrow=1)
+g4 <- ggplot(data_age) +
+  geom_point(aes(age, 1-mean)) +
+  geom_line(aes(age, 1-mean, group=1), lty=2) +
+  geom_errorbar(aes(age, ymin=1-lwr, ymax=1-upr), width=0.2) +
+  scale_x_discrete("Age") +
+  scale_y_continuous("Inferred proportion of infections that are subclinical") +
+  ggtitle("D. Surveillance data from 6 countries") +
+  theme(
+    panel.grid = element_blank()
+  )
 
-saveGG(gfinal, width=8, height=6)
+gfinal <- arrangeGrob(gcomb1, g3, g4, nrow=1)
+
+saveGG(gfinal, width=12, height=6)

### 2022-07-23 main.tex

@@
 \newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
-\newcommand{\bmb}[1]{\comment{blue}{BMB}{#1}}
-\newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}
+\newcommand{\weitz}[1]{\comment{blue}{WEITZ}{#1}}
 
@@
 Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
-Even though asymptomatic infections reduce the severity at the individual level, they can still make population-level outcomes worse by transmitting to others unknowingly.
-Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities in the presence of a strong reduction in symptomatic transmission rates, which can be caused by either behavioral changes or mitigation efforts.
-We further generalize this result to show that intermediate levels of non-symptomatic transmission (including both presymptomatic and asymptomatic transmission) can also lead to the highest levels of fatalities.
+Even though asymptomatic infections reduce severity at the individual level, they can still make population-level outcomes worse if asymptomatic individuals transmit more because they are unaware that they are infected.
+Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when this increase in transmission, due either to individual behavior or mitigation efforts, is strong.
+We generalize this result to include presymptomatic transmission, and show that in this case as well intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
 Finally, we extend our framework to illustrate the importance of immunity profile in determining epidemic trajectories of future variants.

### 2022-07-23 main.tex
spelling

@@
 Sang Woo Park\textsuperscript{1},
-Jonathan Dushoff\textsuperscript{2,3,4},
-Joshua S. Weitz\textsuperscript{5,6,7}
+Bryan T. Grenfell\textsuperscript{1,2},
+Jonathan Dushoff\textsuperscript{3,4,5},
+Joshua S. Weitz\textsuperscript{6,7,8}
 \\
@@
 \\
-\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
+\textbf{2} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
 \\
-\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{6} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
 \\
-\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
+\textbf{7} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\\
+\textbf{8} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
 \\
@@
 It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}, but there has been a considerable heterogeneity in the estimates of asymptomatic prevalence in other settings \citep{alene2021magnitude}.
-The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissibility (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out becase it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
+The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissibility (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
 Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
@@
 
-In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generallly, non-symptomatic transmission) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 

### 2022-07-25 main.tex
minor edits

@@
 Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
-Even though asymptomatic infections reduce severity at the individual level, they can still make population-level outcomes worse if asymptomatic individuals transmit more because they are unaware that they are infected.
+Even though asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals transmit more than symptomatic individuals because asymptomatic individuals are unaware that they are infected.
 Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when this increase in transmission, due either to individual behavior or mitigation efforts, is strong.
-We generalize this result to include presymptomatic transmission, and show that in this case as well intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
-Finally, we extend our framework to illustrate the importance of immunity profile in determining epidemic trajectories of future variants.
+We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
+Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories of future variants.
 
@@
 The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \citep{rajgor2020many,VERITY2020669,yang2020early}.
-But many infections were never documented;
+But many infections are never documented;
 the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \citep{levin2020assessing}.
@@
 On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and death. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that they are less likely to take precautions and relatively more likely to infect others;
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
 asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \citep{xie2022long}.
@@
 The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}B).
-We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size.
+We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size of the outbreak.
 The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population because protected individuals are less likely to develop symptoms:
@@
 In doing so, we first simulate the dynamics of a wildtype variant for 1 year using our base model using identical parameters as in \fref{base}.
-We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
+We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}A), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
 We consider two types of variants (which are simulated separately): one with the same severity $p$ (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
@@
 In this case, cross protection against infection has a large effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
-Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level.
+Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level;
+we note that the second wave of deaths is still high (and having higher peaks in some cases) even if the overall deaths are lower.
 
@@
 but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
-In particular, if immunity provides stronger protection against transmission among asymptomatically infected individuals (than among symptomatically infected individuals), asymptomatic infections will not necessarily make population-level outcomes worse.
+In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts 
 Estimating protection against different endpoints (e.g., infection, symptom, death, and transmission) can help narrow this uncertainty.
@@
 For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make their individual-level outcomes better.
-However, asymptomatically infected individuals, especially among the younger population with high contact rates, can still transmit to other, potentially making the population-level outcomes worse than they would be if intervention measures were distributed differently.
+However, asymptomatically infected individuals, especially among younger individuals with high contact rates, can still transmit to other, older individuals, potentially making the population-level outcomes worse than they would be if intervention measures were distributed differently.
 We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.
 
-In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword by providing a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
-Extending our framework further shows that the immunity profile plays a critical role in determining the dynamics of future variant. 
+In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword leading to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants.
 For example, while protection against symptoms unaccompanied by protection against transmission protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
-Similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \citep{koelle2022changing}.
+A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \citep{koelle2022changing}.
 
@@
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
-Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\,\mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\,\mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
+Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\ \mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\ \mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
 Under this parameterization, we have symptomatic and asymptomatic reproduction numbers of $\RR_s = 4$ and $\RR_a = 3$.
@@
 where $y = (1/\phi - 1)/\eta + (1/\eta - 1)/\rho$.
-In order to keep the mean infectious period fixed, we assume $1/\sigma=2\,\mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\,\mathrm{days}$.
+In order to keep the mean infectious period fixed, we assume $1/\sigma=2\ \mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\ \mathrm{days}$.
 All other parameters are same as before.

### 2022-07-25 main.tex
nothing very interesting

@@
 It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}, but there has been a considerable heterogeneity in the estimates of asymptomatic prevalence in other settings \citep{alene2021magnitude}.
-The relative transmission rate of asymptomatically infected individuals was not well constrained, but relatively low transmissibility (e.g., between 0\%-25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
-Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals, but their assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
+The relative transmission rate of asymptomatically infected individuals was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
+Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
 Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \citep{Kissler2020});
 however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
-We note asymptomaticity is expected to be more heterogeneous in real outbreaks.
+
+We note also that asymptomaticity is expected to be more heterogeneous in real outbreaks.
 For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020};
@@
 Even though we assumed a homogeneous population throughout, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity.
-For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make their individual-level outcomes better.
-However, asymptomatically infected individuals, especially among younger individuals with high contact rates, can still transmit to other, older individuals, potentially making the population-level outcomes worse than they would be if intervention measures were distributed differently.
+For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make individual-level outcomes better.
+However, asymptomatically infected individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
 We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.

### 2022-07-31 main.tex
minor clean up

@@
 Data are publicly available in \cite{Kissler2020}.
-(D) Inferred proportion of infections that are subclinical for each age group using case data from six countries (China, Italy, Japan, Singapore, South Korea, and Canada) \cite{davies2020}.
+(D) Inferred proportion of infections that are subclinical for each age group using surveillance data from six countries (China, Italy, Japan, Singapore, South Korea, and Canada) \cite{davies2020}.
 }
@@
 It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}, but there has been a considerable heterogeneity in the estimates of asymptomatic prevalence in other settings \citep{alene2021magnitude}.
-The relative transmission rate of asymptomatically infected individuals was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatically infected individuals (\fref{evidence}B).
+The relative transmission rate of asymptomatic individuals was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
 Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
-Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatically infected individuals (\fref{evidence}C, \citep{Kissler2020});
+Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatic individuals (\fref{evidence}C, \citep{Kissler2020});
 however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
@@
 We note also that asymptomaticity is expected to be more heterogeneous in real outbreaks.
-For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020};
-differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
+For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
+Differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
 For now, we primarily focus on a homogeneous population and return to the age effect in the discussion.
@@
 To explore this idea, we propose a simple epidemic model,
-in which infected individuals can be asymptomatically or symptomatically infected, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
-Asymptomatically infected individuals always recover, whereas a fraction $f$ of symptomatically infected individuals die.
-Asymptomatically and symptomatically infected individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
-Our key assumption is that symptomatically infected individuals take greater precautions than do asymptomatically infected individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
-we note that the parameter $\delta$ may also capture intervention measures that target symptomatically infected individuals, such as symptom-based isolation. 
-For our main simulations, we assume that asymptomatically infected individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatically infected individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
+in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
+Asymptomatic individuals always recover, whereas a fraction $f$ of symptomatic individuals die.
+Asymptomatic and symptomatic individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
+Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
+we note that the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
+For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
 We evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
@@
 \textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
-(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatic individuals; $I_s$ represents symptomatic individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.
@@
 therefore, the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}C).
-Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatically infected individuals, who can readily transmit infections to other individuals.
+Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatic individuals, who can readily transmit infections to other individuals.
 This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
@@
 For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make individual-level outcomes better.
-However, asymptomatically infected individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
+However, asymptomaticindividuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
 We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.
@@
 \end{align}
-where the transmission rate $\beta$ and recovery rate $\gamma$ can be potentially differ between asymptomatic and symptomatically infected individuals.  
-Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatically infected individuals.
+where the transmission rate $\beta$ and recovery rate $\gamma$ can be potentially differ between asymptomatic and symptomatic individuals.  
+Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatic individuals.
 Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\ \mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\ \mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
@@
 \textbf{Schematic diagram and simulations of a model with pre-symptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
-(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents pre-symptomatically infected individuals; $I_a$ represents symptomatically infected individuals; $I_s$ represents symptomatically infected individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
+(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents pre-symptomatic individuals; $I_a$ represents symptomatic individuals; $I_s$ represents symptomatic individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
 See Methods for model details.

### 2022-07-31 main.tex
done

@@
 Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
-we note that the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
+the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
+We note that intervention measures that target asymptomatic infections would reduce the effective value of delta---for example, frequent testing and isolation may effectively increase the recovery rate $\gamma_a$ of asymptomatic individuals.
 For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
@@
 We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
+See Supplementary Text for model details and Supplementary Table S1 for parameter descriptions and values.
 }
@@
 
-We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission (Supplementary Figure S1), to ask the following question:
+We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission, to ask the following question:
 does an intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
 For this model, we assume that $\delta$ decreases transmission only after symptom onset.
-We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission.
-Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S2).
+We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Supplementary Text for model details and Supplementary Table S2 for parameter descriptions and values).
+Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1).
 One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
@@
 We assume that $10^{-4}$ proportion of individuals are initially infected.
+See Supplementary Text for model details and Supplementary Table S3 for parameter descriptions and values.
 }
@@
 For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make individual-level outcomes better.
-However, asymptomaticindividuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
+However, asymptomatic individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
 We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.
@@
 
-\section*{Supplementary Materials}
+\section*{Supplementary Text}
 \setcounter{figure}{0}
 \renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 
-\subsection*{Methods}
-
-\subsubsection*{Epidemic models with asymptomatic infection and transmission in the absence of immunity}
+\subsection*{Epidemic models with asymptomatic infection and transmission in the absence of immunity}
 
@@
 
-\subsubsection*{Epidemic models with asymptomatic infection and transmission in the presence of immunity}
+\subsection*{Epidemic models with asymptomatic infection and transmission in the presence of immunity}
 
@@
 
+\section*{Supplementary Tables}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
+      \hline
+      $\beta_s$ & Symptomatic transmission rate & $0.8/\mathrm{days}$\\
+      \hline
+      $\beta_a$ & Asymptomatic transmission rate & $0.75 \beta_s$ \\
+      \hline
+      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_s$ & Mean symptomatic infectious period & $5\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_a$ & Mean asymptomatic infectious period & $5\ \mathrm{days}$\\
+      \hline
+      $p$ & Proportion asymptomatic & 0--1\\
+      \hline
+      $f$ & Fatality rate for symptomatic case & 0.01\\
+      \hline
+      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
+    \end{tabular}
+    \caption{Paramter descriptions and values for the basic asymptomatic model.}
+    \label{tab:table1}
+  \end{center}
+\end{table}
+
+\pagebreak
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
+      \hline
+      $\beta_s$ & Symptomatic transmission rate & See Supplementary Text\\
+      \hline
+      $\beta_a$ & Asymptomatic transmission rate & See Supplementary Text \\
+      \hline
+      $\beta_p$ & Presymptomatic transmission rate & See Supplementary Text \\
+      \hline
+      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
+      \hline
+      $1/\sigma$ & Mean presymptomatic infectious period & $2\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_s$ & Mean symptomatic infectious period & $3\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_a$ & Mean asymptomatic infectious period & $3\ \mathrm{days}$\\
+      \hline
+      $p$ & Proportion asymptomatic & 0--1\\
+      \hline
+      $f$ & Fatality rate for symptomatic case & 0.01\\
+      \hline
+      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
+    \end{tabular}
+    \caption{Paramter descriptions and values for the generalized asymptomatic model.}
+    \label{tab:table1}
+  \end{center}
+\end{table}
+
+\pagebreak
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
+      \hline
+      $\beta_s$ & Symptomatic transmission rate & $0.8/\mathrm{days}$\\
+      \hline
+      $\beta_a$ & Asymptomatic transmission rate & $0.75 \beta_s$ \\
+      \hline
+      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_s$ & Mean symptomatic infectious period & $5\ \mathrm{days}$\\
+      \hline
+      $1/\gamma_a$ & Mean asymptomatic infectious period & $5\ \mathrm{days}$\\
+      \hline
+      $p$ & Proportion asymptomatic & 0--1\\
+      \hline
+      $f$ & Fatality rate for symptomatic case & 0.01\\
+      \hline
+      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
+      \hline
+      $\epsilon_i$ & Protection against infection & 0--0.8\\
+      \hline
+      $\epsilon_s$ & Protection against symptoms & 0--0.8\\
+      \hline
+      $\epsilon_d$ & Protection against deaths & 0--0.8\\
+    \end{tabular}
+    \caption{Paramter descriptions and values for the asymptomatic model with immunity.}
+    \label{tab:table1}
+  \end{center}
+\end{table}
+
+\pagebreak
+
 \section*{Supplementary Figures}
@@
 (B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
+
+See Supplementary Text for model details and Supplementary Table S2 for parameter descriptions and values.
 }

### 2022-08-01 main.tex
remove space
done

@@
 Jonathan Dushoff\textsuperscript{3,4,5},
-Joshua S. Weitz\textsuperscript{6,7,8}
+Joshua S. Weitz\textsuperscript{6,7,8,*}
 \\
@@
 
+*Corresponding author: jsweitz@gatech.edu
 \bigskip
@@
 Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
-Even though asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals transmit more than symptomatic individuals because asymptomatic individuals are unaware that they are infected.
-Here, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when this increase in transmission, due either to individual behavior or mitigation efforts, is strong.
+Even though asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals---unaware they are infected---transmit more than symptomatic individuals.
+Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the increase in asymptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
 We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
-Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories of future variants.
+Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
 
@@
 \textbf{Asymptomatic transmissibility of SARS-CoV-2.}
-(A) Posterior estimates of the proportion of asymptomatic infections from the Diamond Princess Cruise Ship \cite{emery2020}.
+(A) Posterior estimates of the proportion of asymptomatic infections from the Diamond Princess cruise ship \cite{emery2020}.
 (B) Posterior estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship \cite{emery2020}.
@@
 
-Early in the pandemic, a COVID-19 outbreak on the Diamond Princess Cruise Ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
+Early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
 the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
-It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}, but there has been a considerable heterogeneity in the estimates of asymptomatic prevalence in other settings \citep{alene2021magnitude}.
-The relative transmission rate of asymptomatic individuals was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
+It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}.
+The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
 Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
@@
 however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
-
-We note also that asymptomaticity is expected to be more heterogeneous in real outbreaks.
+We note also that asymptomaticity is expected to be more heterogeneous in a diversity of outbreak settings \citep{alene2021magnitude}.
 For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
 Differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
-For now, we primarily focus on a homogeneous population and return to the age effect in the discussion.
+For now, we primarily focus on a homogeneous population and return to the age effect in discussing our model-based findings.
 
-Despite quantitative uncertainties in asymptomatic transmissibility, it is clear that individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
+Despite quantitative uncertainties in asymptomatic transmissibility, individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
 This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
@@
 Asymptomatic individuals always recover, whereas a fraction $f$ of symptomatic individuals die.
-Asymptomatic and symptomatic individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and recovery rates ($\gamma_a$ and $\gamma_s$).
+Asymptomatic and symptomatic individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and removal rates ($\gamma_a$ and $\gamma_s$).
 Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
 the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
-We note that intervention measures that target asymptomatic infections would reduce the effective value of delta---for example, frequent testing and isolation may effectively increase the recovery rate $\gamma_a$ of asymptomatic individuals.
-For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal recovery rates ($\gamma_a = \gamma_s$).
+We note that intervention measures that target asymptomatic infections would reduce the effective value of $\delta$---for example, frequent testing and isolation may effectively increase the removal rate $\gamma_a$ of asymptomatic individuals.
+For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal removal rates ($\gamma_a = \gamma_s$).
 We evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
@@
 The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population because protected individuals are less likely to develop symptoms:
-therefore, the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}C).
+the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}C).
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatic individuals, who can readily transmit infections to other individuals.
@@
 The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic.
-In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread better and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms but not against infection or transmission.
+In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread more effectively and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms but not against infection or transmission.
 
 Next, we consider a more realistic scenario in which immunity provides protection against both symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
-In this case, cross protection against infection has a large effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
+In this case, cross-protection against infection has a large effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
 Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level;
@@
 but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
-In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts 
+In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts.
 Estimating protection against different endpoints (e.g., infection, symptom, death, and transmission) can help narrow this uncertainty.
@@
 
-Even though we assumed a homogeneous population throughout, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity.
-For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and make individual-level outcomes better.
+Even though we assumed a homogeneous population throughout, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity (as shown in \fref{evidence}D).
+For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and improve individual-level outcomes.
 However, asymptomatic individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
@@
 Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
-As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections) on preventing infections, and not just diseases, will be critical to controlling the course of the pandemic.
+As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections) on preventing infections, and not just diseases, will be critical to controlling the course of the pandemic \citep{lopman2021framework}.
+
+\section*{Acknowledgements}
+
+JSW acknowledges support from the Chaire Blaise Pascal Program of the Île-de-France region.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
+\section*{Data availability}
+
+All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/asymptomaticvariant}).
 
@@
 \end{align}
-where the transmission rate $\beta$ and recovery rate $\gamma$ can be potentially differ between asymptomatic and symptomatic individuals.  
+where the transmission rate $\beta$ and removal rate $\gamma$ can be potentially differ between asymptomatic and symptomatic individuals.  
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatic individuals.
@@
 \end{align}
-where $\epsilon$ represents the degree of protection against infection, symptoms and death. 
+where $0\leq \epsilon \leq 1$ represents the degree of protection against infection, symptoms and death. 
 The force of infection $\lambda(t)$ is given by:
@@
 (B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
-
 See Supplementary Text for model details and Supplementary Table S2 for parameter descriptions and values.

### 2022-08-01 main.tex
ttweak

@@
 Sang Woo Park\textsuperscript{1},
-Bryan T. Grenfell\textsuperscript{1,2},
-Jonathan Dushoff\textsuperscript{3,4,5},
+Jonathan Dushoff\textsuperscript{2,3,4},
+Bryan T. Grenfell\textsuperscript{1,5},
 Joshua S. Weitz\textsuperscript{6,7,8,*}
@@
 \\
-\textbf{2} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
+\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\textbf{5} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
 \\

### 2022-08-03 main.tex
citations

@@
 On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and death. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
+On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors,moghadas2020implications}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
 asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \citep{xie2022long}.
@@
 Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity,
-but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune}.
+but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune,gallagher2021indirect}.
 In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts.
@@
 In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword leading to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
-Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants.
+Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants;
+these results extend previous work on post-pandemic trajectories that focus primarily on cross-immunity \citep{kissler2020projecting,lavine2021immunological}.
 For example, while protection against symptoms unaccompanied by protection against transmission protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.

### 2022-08-03 main.tex
done

@@
 We note also that asymptomaticity is expected to be more heterogeneous in a diversity of outbreak settings \citep{alene2021magnitude}.
-For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals are less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
+For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals were less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
 Differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
@@
 
-To explore this idea, we propose a simple epidemic model,
+To explore this idea, we propose an epidemic model,
 in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
@@
 Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
-As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections) on preventing infections, and not just diseases, will be critical to controlling the course of the pandemic \citep{lopman2021framework}.
+As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \citep{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of the pandemic.
 

### 2022-08-03 main.tex
word count

@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission can lead to the highest levels of epidemic fatalities
+		Intermediate levels of asymptomatic transmission lead to the highest levels of fatalities \swp{89/96 characters}
 	}
@@
 
+\section*{Short title}
+
+Asymptomatic transmission and fatalities \swp{40/40 characters}
+
+\section*{One-sentence summary}
+
+Asymptomatic infections improve individual-level outcomes but increase overall transmission, thereby causing more deaths. \swp{121/125 characters}
+
 \section*{Abstract}
 
-Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the ongoing pandemic.
-Even though asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals---unaware they are infected---transmit more than symptomatic individuals.
+Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the pandemic.
+Although asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals---unaware they are infected---transmit more than symptomatic individuals.
 Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the increase in asymptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
-We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission can lead to the highest levels of fatalities.
+We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission lead to the highest levels of fatalities.
 Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
+\swp{125/125 words}
 
@@
 We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
-See Supplementary Text for model details and Supplementary Table S1 for parameter descriptions and values.
+See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
 }
@@
 For this model, we assume that $\delta$ decreases transmission only after symptom onset.
-We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Supplementary Text for model details and Supplementary Table S2 for parameter descriptions and values).
+We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).
 Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1).
@@
 We assume that $10^{-4}$ proportion of individuals are initially infected.
-See Supplementary Text for model details and Supplementary Table S3 for parameter descriptions and values.
+See Materials and Methods for model details and Supplementary Table S3 for parameter descriptions and values.
 }
@@
 
-\section*{Supplementary Text}
+\section*{Materials and Methods}
 \setcounter{figure}{0}
@@
       \hline
-      $\beta_s$ & Symptomatic transmission rate & See Supplementary Text\\
+      $\beta_s$ & Symptomatic transmission rate & See Materials and Methods\\
       \hline
-      $\beta_a$ & Asymptomatic transmission rate & See Supplementary Text \\
+      $\beta_a$ & Asymptomatic transmission rate & See Materials and Methods \\
       \hline
-      $\beta_p$ & Presymptomatic transmission rate & See Supplementary Text \\
+      $\beta_p$ & Presymptomatic transmission rate & See Materials and Methods \\
       \hline
@@
 (B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
-See Supplementary Text for model details and Supplementary Table S2 for parameter descriptions and values.
+See Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values.
 }

### 2022-08-03 main.tex
bib and title

@@
 
-\usepackage{natbib}
-\bibliographystyle{unsrt}
+% \usepackage{natbib}
+\bibliographystyle{Science}
 \date{\today}
@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission lead to the highest levels of fatalities \swp{89/96 characters}
+		Intermediate levels of asymptomatic transmission can lead to the highest levels of fatalities
 	}
@@
 
-Asymptomatic transmission and fatalities \swp{40/40 characters}
+Asymptomatic transmission and fatalities
 
@@
 
-Asymptomatic infections improve individual-level outcomes but increase overall transmission, thereby causing more deaths. \swp{121/125 characters}
+Asymptomatic infections improve individual-level outcomes but increase overall transmission, thereby causing more deaths.
 
@@
 Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
-\swp{125/125 words}
 
@@
 SARS-CoV-2 has had devastating effects at the population level.
-However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \citep{nogrady2020data}.
-The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \citep{rajgor2020many,VERITY2020669,yang2020early}.
+However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \cite{nogrady2020data}.
+The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \cite{rajgor2020many,VERITY2020669,yang2020early}.
 But many infections are never documented;
-the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \citep{levin2020assessing}.
+the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \cite{levin2020assessing}.
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
@@
 Early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
-the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \citep{mizumoto2020estimating}.
-It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \citep{emery2020}.
+the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \cite{mizumoto2020estimating}.
+It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
 The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
-Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \citep{ferretti2020quantifying,lavezzo2020}.
-Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatic individuals (\fref{evidence}C, \citep{Kissler2020});
-however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \citep{romero2021ct}.
-We note also that asymptomaticity is expected to be more heterogeneous in a diversity of outbreak settings \citep{alene2021magnitude}.
+Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \cite{ferretti2020quantifying,lavezzo2020}.
+Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatic individuals (\fref{evidence}C, \cite{Kissler2020});
+however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \cite{romero2021ct}.
+We note also that asymptomaticity is expected to be more heterogeneous in a diversity of outbreak settings \cite{alene2021magnitude}.
 For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals were less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
@@
 On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and death. 
-On the other hand, asymptomatic infections are less likely to be detected \citep{fraser2004factors,moghadas2020implications}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
-asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \citep{xie2022long}.
+On the other hand, asymptomatic infections are less likely to be detected \cite{fraser2004factors,moghadas2020implications}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
+asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \cite{xie2022long}.
 Altogether, the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
@@
 For this particular model, it does not make biological sense for $\delta$ to be greater than the amount of post-symptomatic transmission, because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
-While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of these were likely affected by intervention and behavioral effects, as they were conducted after SARS-CoV-2 awareness became widespread \citep{he2020temporal}.
+While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of these were likely affected by intervention and behavioral effects, as they were conducted after SARS-CoV-2 awareness became widespread \cite{he2020temporal}.
 Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission could have been as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
@@
 In other words, we assume that asymptomatic infections in protected and unprotected people have the same reproduction numbers (and likewise for the symptomatic infections).
-In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \citep{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
+In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \cite{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
 We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
@@
 The outcomes in our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
-Despite moderate levels of vaccine effectiveness against symptomatic and reduced levels of severe cases caused by the Omicron variant, especially after booster shots \citep{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \citep{pearson2021omicron}.
-This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \citep{sun2022omicron}.
-Moreover, even though the Omicron variant is probably milder than the Delta variant \citep{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \citep{Iacobuccio254,faust2022omicron,sigal2022estimating}.
+Despite moderate levels of vaccine effectiveness against symptomatic and reduced levels of severe cases caused by the Omicron variant, especially after booster shots \cite{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \cite{pearson2021omicron}.
+This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \cite{sun2022omicron}.
+Moreover, even though the Omicron variant is probably milder than the Delta variant \cite{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \cite{Iacobuccio254,faust2022omicron,sigal2022estimating}.
 
@@
 Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity,
-but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \citep{saad2020immune,gallagher2021indirect}.
+but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \cite{saad2020immune,gallagher2021indirect}.
 In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts.
@@
 Finally, we assumed that asymptomatic and symptomatic individuals are infected for the same amount of time.
-Analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \citep{Kissler2020};
+Analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \cite{Kissler2020};
 however, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
-The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \citep{park2020time,harris2022time};
+The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \cite{park2020time,harris2022time};
 nonetheless, we expect that predictions on the final size of the epidemic and total fatalities will be robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
@@
 However, asymptomatic individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
-We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \citep{moore2021modelling}.
+We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \cite{moore2021modelling}.
 
@@
 Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants;
-these results extend previous work on post-pandemic trajectories that focus primarily on cross-immunity \citep{kissler2020projecting,lavine2021immunological}.
+these results extend previous work on post-pandemic trajectories that focus primarily on cross-immunity \cite{kissler2020projecting,lavine2021immunological}.
 For example, while protection against symptoms unaccompanied by protection against transmission protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
-A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \citep{koelle2022changing}.
+A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \cite{koelle2022changing}.
 
@@
 % Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
-Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \citep{chande2020real,sinclair2021pairing}, asymptomatic testing programs \citep{mina2021covid,gibson2022surv}, mask-wearing indoors and in crowded environments \citep{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \citep{allen2021indoor,wang2021airborne}) while improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
-Given the link between age and asymptomatic infections \citep{davies2020}, interventions may consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
+Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \cite{Bavel2020,chande2020real,sinclair2021pairing}, asymptomatic testing programs \cite{mina2021covid,gibson2022surv,Ranoa2022}, mask-wearing indoors and in crowded environments \cite{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \cite{allen2021indoor,wang2021airborne}) while improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
+Given the link between age and asymptomatic infections \cite{davies2020}, interventions may consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
 Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
-As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \citep{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of the pandemic.
+As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \cite{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of the pandemic.
 
@@
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatic individuals.
-Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\ \mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\ \mathrm{days}$, and $f=0.01$ \citep{park2020reconciling}.
+Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\ \mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\ \mathrm{days}$, and $f=0.01$ \cite{park2020reconciling}.
 Under this parameterization, we have symptomatic and asymptomatic reproduction numbers of $\RR_s = 4$ and $\RR_a = 3$.
@@
     \caption{Paramter descriptions and values for the generalized asymptomatic model.}
-    \label{tab:table1}
+    \label{tab:table2}
   \end{center}
@@
     \caption{Paramter descriptions and values for the asymptomatic model with immunity.}
-    \label{tab:table1}
+    \label{tab:table3}
   \end{center}

### 2022-09-07 main.tex
title again
edits

@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission can lead to the highest levels of fatalities
+		Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities
 	}
@@
 
-\section*{Short title}
-
-Asymptomatic transmission and fatalities
-
-\section*{One-sentence summary}
-
-Asymptomatic infections improve individual-level outcomes but increase overall transmission, thereby causing more deaths.
-
 \section*{Abstract}
@@
 Although asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals---unaware they are infected---transmit more than symptomatic individuals.
-Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the increase in asymptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
+Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the decrease in symptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
 We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission lead to the highest levels of fatalities.
 Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
+In particular, when immunity provides protection against symptoms, but not against infections or deaths, epidemic trajectories can have faster infection growth rates and higher peak, which in turn can lead to more deaths.
+Nonetheless, even a modest protection against infection can mitigate the population-level effects of asymptomatic spread .
+
+\section*{Significance}
+
+Asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, who can then develop symptoms and experience severe outcomes.
+Here, we explore the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals from behavioral and intervention efforts.
+In this case, the number of infections increases with asymptomatic prevalence, whereas epidemic fatalities peak at intermediate levels of asymptomatic prevalence.
+We integrate our findings to understanding the effects of interplay between behavior and immunity on disease spread and shed light on the dynamics of future variants.
 
 \pagebreak
+
+\section*{Introduction}
 
@@
 asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \cite{xie2022long}.
-Altogether, the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
 
-To explore this idea, we propose an epidemic model,
-in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
+In this manuscript, we explore the effects of asymptomatic infection and transmission on disease severity at the population level.
+In doing so, we assume that symptomatic individuals can reduce their transmission, reflecting changes in behavior (e.g., self-isolation after symptom onset) or nonpharmaceutical intervention measures.
+Under this assumption, we show that the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
+We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants.
+
+\section*{Results and Discussion}
+
+We propose an epidemic model, in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
 Asymptomatic individuals always recover, whereas a fraction $f$ of symptomatic individuals die.
@@
 For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal removal rates ($\gamma_a = \gamma_s$).
-We evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
+We then evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
 \begin{figure}[!ht]
@@
 
-\fref{base}B--C shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2, without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
+\fref{base}B--C shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2 (Table S1), without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
 For this model, the basic reproduction number is given by:
@@
 We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).
-Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1).
+Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1; Table S2).
 One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
@@
 In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \cite{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
-We assume a relatively strong behavioral effect $\delta=0.8$ for illustration.
+We assume a relatively strong behavioral effect $\delta=0.8$ for illustration (Table S3).
 
@@
 
-\section*{Acknowledgements}
-
-JSW acknowledges support from the Chaire Blaise Pascal Program of the Île-de-France region.
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
-\section*{Data availability}
-
-All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/asymptomaticvariant}).
-
-\pagebreak
-
 \section*{Materials and Methods}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
 
@@
 where $y = (1/\phi - 1)/\eta + (1/\eta - 1)/\rho$.
-In order to keep the mean infectious period fixed, we assume $1/\sigma=2\ \mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\ \mathrm{days}$.
-All other parameters are same as before.
+In order to keep the mean infectious period fixed, we assume $1/\sigma=2\ \mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\ \mathrm{days}$. All other parameters are same as before.
 
@@
 \end{align}
-where $0\leq \epsilon \leq 1$ represents the degree of protection against infection, symptoms and death. 
+where $0\leq \epsilon_i, \epsilon_s, \epsilon_d \leq 1$ represents the degree of protection against infection, symptoms and death, respectively. 
 The force of infection $\lambda(t)$ is given by:
@@
 Here, subscripts $p$ denote individuals who are immune and therefore are protected.
+
+\section*{Acknowledgements}
+
+JSW acknowledges support from the Chaire Blaise Pascal Program of the Île-de-France region.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
+\section*{Data availability}
+
+All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/asymptomaticvariant}).
+
+\pagebreak
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 

### 2022-09-08 main.tex
done

@@
 
-Asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, who can then develop symptoms and experience severe outcomes.
-Here, we explore the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals from behavioral and intervention efforts.
-In this case, the number of infections increases with asymptomatic prevalence, whereas epidemic fatalities peak at intermediate levels of asymptomatic prevalence.
-We integrate our findings to understanding the effects of interplay between behavior and immunity on disease spread and shed light on the dynamics of future variants.
+During an epidemic, asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, potentially leading to severe outcomes and fatalities.
+This manuscript proposes and analyzes the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals due to behavioral change and interventions.
+Theory and simulations reveal that although the number of infections increases with asymptomatic prevalence, epidemic fatalities peak at intermediate levels of asymptomatic prevalence.
+The same framework also shows how milder variants at the individual level can potentially lead to worse outcomes for the population---of relevance to ongoing efforts to explore the interplay between behaviour, immunity, and viral variants.
 
@@
 In this manuscript, we explore the effects of asymptomatic infection and transmission on disease severity at the population level.
-In doing so, we assume that symptomatic individuals can reduce their transmission, reflecting changes in behavior (e.g., self-isolation after symptom onset) or nonpharmaceutical intervention measures.
+In doing so, we assume that symptomatic individuals reduce their transmission, reflecting changes in behavior (e.g., self-isolation after symptom onset) and/or nonpharmaceutical intervention measures.
 Under this assumption, we show that the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
-We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants.
+We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants and explore mechanisms by which milder variants at the individual-level can nonetheless lead to similar or worse population-level outcomes.
 
@@
 
-In summary, using a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword leading to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+Via theory and simulation analysis of a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword leading to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants;
@@
 
-SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms. 
+As is increasingly evident, SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms. 
 % Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
@@
 
-First, we consider a compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
-The model dynamics are as follows:
+We consider a compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
+The basic model dynamics are as follows:
 \begin{align}
@@
 
-We then extend the first model to include both pre-symptomatic and asymptomatic transmission:
+We then extend this model to include both pre-symptomatic and asymptomatic transmission:
 \begin{align}
@@
 \end{align}
-For this model, the pre-symptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
+For this generalized model, the pre-symptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
 these reproduction numbers represent the average number of secondary cases caused by an infected individual in each compartment.
@@
 
-We then model the spread of infection in a partially immune population.
-The model dynamics are as follows:
+We model the spread of infection in a partially immune population as follows:
 \begin{align}
@@
 \end{equation}
-Here, subscripts $p$ denote individuals who are immune and therefore are protected.
+Here, subscripts $p$ denote individuals who are immune and therefore are protected, and as before the subscripts $a$ and $s$ denote asymptomatic and symptomatic infections.
 

### 2022-09-08 main.tex
Homing (xiangshan)

@@
 Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the decrease in symptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
-We generalize this result to include presymptomatic transmission, showing how intermediate levels of non-symptomatic transmission lead to the highest levels of fatalities.
+We generalize this result to include presymptomatic transmission, showing that intermediate levels of non-symptomatic transmission lead to the highest levels of fatalities.
 Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
-In particular, when immunity provides protection against symptoms, but not against infections or deaths, epidemic trajectories can have faster infection growth rates and higher peak, which in turn can lead to more deaths.
-Nonetheless, even a modest protection against infection can mitigate the population-level effects of asymptomatic spread .
+In particular, when immunity provides protection against symptoms, but not against infections or deaths, epidemic trajectories can have faster growth rates and higher peaks, leading to more total deaths.
+Conversely, even modest levels of protection against infection can mitigate the population-level effects of asymptomatic spread.
 
@@
 
-During an epidemic, asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, potentially leading to severe outcomes and fatalities.
-This manuscript proposes and analyzes the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals due to behavioral change and interventions.
-Theory and simulations reveal that although the number of infections increases with asymptomatic prevalence, epidemic fatalities peak at intermediate levels of asymptomatic prevalence.
+During an epidemic, asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, potentially leading to severe outcomes, including fatalities.
+This manuscript analyzes the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals due to behavioral change and interventions.
+Theory and simulations reveal that when reduction is strong the number of infections increases with asymptomatic frequency, while fatalities peak at intermediate levels.
 The same framework also shows how milder variants at the individual level can potentially lead to worse outcomes for the population---of relevance to ongoing efforts to explore the interplay between behaviour, immunity, and viral variants.
@@
 However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \cite{nogrady2020data}.
-The ratio of fatalities to documented cases (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \cite{rajgor2020many,VERITY2020669,yang2020early}.
+The ratio of fatalities to documented \emph{cases} (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \cite{rajgor2020many,VERITY2020669,yang2020early}.
 But many infections are never documented;
-the ratio of fatalities to total infections (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \cite{levin2020assessing}.
+the ratio of fatalities to total \emph{infections} (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \cite{levin2020assessing}.
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
@@
 the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \cite{mizumoto2020estimating}.
-It has been estimated that 75\% (95\% C: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
-The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
+It has been estimated that 75\% (95\% CI: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
+The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained by the analysis, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
+
 Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \cite{ferretti2020quantifying,lavezzo2020}.
 Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatic individuals (\fref{evidence}C, \cite{Kissler2020});
-however, differences between inferred total viral load from Ct values and infectious viral load add uncertainties to how well asymptomatic individuals can transmit relative to that of symptomatic individuals \cite{romero2021ct}.
-We note also that asymptomaticity is expected to be more heterogeneous in a diversity of outbreak settings \cite{alene2021magnitude}.
+however, differences between inferred \emph{total} viral load from Ct values and \emph{infectious} viral load add uncertainty \cite{romero2021ct}.
+We note also that asymptomaticity can change across outbreak settings \cite{alene2021magnitude}.
 For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals were less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.

### 2022-09-08 main.tex
Through intro material

@@
 This manuscript analyzes the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals due to behavioral change and interventions.
-Theory and simulations reveal that when reduction is strong the number of infections increases with asymptomatic frequency, while fatalities peak at intermediate levels.
+Theory and simulations reveal that when reduction is strong the number of infections increases with asymptomatic proportion, while fatalities peak at intermediate levels.
 The same framework also shows how milder variants at the individual level can potentially lead to worse outcomes for the population---of relevance to ongoing efforts to explore the interplay between behaviour, immunity, and viral variants.
@@
 (B) Posterior estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship \cite{emery2020}.
-Bar charts represent the posterior distributions.
+%% Bar charts represent the posterior distributions.
 Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a pre-symptomatic stage with an average of 2.1 days. 
@@
 Points represent each Ct measurement.
-Lines and shaded areas represent the smooth trajectories estimated via LOESS and the associated 95\% confidence intervals.
+Lines and shaded areas show LOESS fits.
+\jd{I don't really like calling them CIs. Maybe say default loess fits from R or something.}
 Data are publicly available in \cite{Kissler2020}.
@@
 In doing so, we assume that symptomatic individuals reduce their transmission, reflecting changes in behavior (e.g., self-isolation after symptom onset) and/or nonpharmaceutical intervention measures.
-Under this assumption, we show that the prevalence of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
-We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants and explore mechanisms by which milder variants at the individual-level can nonetheless lead to similar or worse population-level outcomes.
+Under this assumption, we show that a high proportion of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
+We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants and explore mechanisms by which milder variants at the individual level can nonetheless lead to similar or worse population-level outcomes.
 

### 2022-09-08 main.tex
Through IRD

@@
 
-We propose an epidemic model, in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
+We propose an epidemic model in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
 Asymptomatic individuals always recover, whereas a fraction $f$ of symptomatic individuals die.
 Asymptomatic and symptomatic individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and removal rates ($\gamma_a$ and $\gamma_s$).
-Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., via reducing contacts or increased mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
+Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., reducing contacts or increasing mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
 the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
 We note that intervention measures that target asymptomatic infections would reduce the effective value of $\delta$---for example, frequent testing and isolation may effectively increase the removal rate $\gamma_a$ of asymptomatic individuals.
-For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is modeled by assuming lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal removal rates ($\gamma_a = \gamma_s$).
-We then evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate for \emph{symptomatic} cases, $f$, constant (the IFR $(1-p)f$ thus decreases as $p$ increases).
+For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is implemented via lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal removal rates ($\gamma_a = \gamma_s$).
+We then evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate \emph{for symptomatic cases}, $f$, constant (the population-level IFR $(1-p)f$ thus decreases as $p$ increases).
 \begin{figure}[!ht]
@@
 therefore, in the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number.
+
 This relationship changes as $\delta$ increases.
 In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic reproductive number (including behavioral response) is less than that the asymptomatic reproductive number.
-For high values of $\delta$, we can find a critical level of asymptomatic proportion, $p_c$:
+When $\delta > 1-1/\RR_s$( and $\RR_a>1$), we can find a critical asymptomatic proportion, $p_c$:
 \begin{equation}
@@
 \end{equation}
-such that $p>p_c$ is required for an outbreak (see threshold effects for large values of $\delta$ in \fref{base}B).
+such that an outbreak will occur exactly when $p>p_c$ (see threshold effects for large values of $\delta$ in \fref{base}B).
 
@@
 For high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
-although fewer individuals die per infection for higher values of $p$, the increase in total infections also leads to an increase in total fatalities.
-In contrast, when $\delta$ is small enough such that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), then total fatalities decrease with $p$ because because both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
+although fewer individuals die per infection for higher values of $p$, the increase in total infections still leads to an increase in total fatalities.
+In contrast, when $\delta$ is small enough that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
+\jd{Got rid of a few words that might imply that this is necessary as well as sufficient for a decreasing curve.}
 
-High values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths may seem unrealistic.
+We can ask whether the high values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths are realistic.
 For this particular model, it does not make biological sense for $\delta$ to be greater than the amount of post-symptomatic transmission, because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
@@
 Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission could have been as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
-There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission suggests that high $\delta$ values are feasible (although not necessarily likely) during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of pre-symptomatic transmission from 80\% to as low as 40\%.
+There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission makes high $\delta$ values at least somewhat more likely during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of pre-symptomatic transmission from 80\% to as low as 40\%.
 
-We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission, to ask the following question:
-does an intermediate amount of non-symptomatic transmission lead to a peak in fatalities?
+We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission.
 For this model, we assume that $\delta$ decreases transmission only after symptom onset.
 We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).
+
 Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1; Table S2).
@@
 We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
-We assume that $10^{-4}$ proportion of individuals are initially infected.
+\jd{Please change all fractions to decimals in above s.}
+The initial exposed proportion is $10^{-4}$.
 See Materials and Methods for model details and Supplementary Table S3 for parameter descriptions and values.
@@
 Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatic individuals, who can readily transmit infections to other individuals.
-This also means that the critical level of asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
+This also means that the critical asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
 We note that the equivalence between protection against symptoms $\epsilon_s$ and fraction asymptomatic $p$ relies on our assumption that immunity does not provide protection against transmission.
@@
 Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using parameters from \fref{base} with $\delta=0.8$.
-Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for two different values of $p$ as shown.
-(B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
+Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for \emph{either} a milder ($p=0.8$) or a not-milder ($p=0.2$) variant.
+\jd{not-milder is not an actual sugggestion\ldots}
+(B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and milder (purple) and similar (orange) invading variants (dashed lines).
 (C, F) Dynamics of daily deaths for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
@@
 Finally, we use our framework to understand the impact of behavioral effects on invading variants (\fref{variant}).
-In doing so, we first simulate the dynamics of a wildtype variant for 1 year using our base model using identical parameters as in \fref{base}.
+We first simulate the dynamics of a wildtype variant for 1 year using our base model with parameters as in \fref{base}.
 We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}A), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
 We consider two types of variants (which are simulated separately): one with the same severity $p$ (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
+\jd{The more severe variant has the same severity. There is no conceptual problem, but I'm worried about a reader-comprehension problem.}
 
@@
 Although the milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
-The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic.
+The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic; the relatively severe second-wave variant causes more deaths than the wildtype first wave, despite causing many fewer cases.
 In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread more effectively and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms but not against infection or transmission.
@@
 Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity,
-but we neglected the effects of immunity on transmission, which also has important effects on disease dynamics \cite{saad2020immune,gallagher2021indirect}.
+but we neglected additional specific effects of immunity on transmission, which can also have important effects on disease dynamics \cite{saad2020immune,gallagher2021indirect}.
 In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts.
@@
 
-Even though we assumed a homogeneous population throughout, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity (as shown in \fref{evidence}D).
+Even though we assumed a homogeneous population here, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity (as shown in \fref{evidence}D).
 For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and improve individual-level outcomes.
@@
 
-Via theory and simulation analysis of a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can represent a double-edged sword leading to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
+Via theory and simulation analysis of a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can under some conditions lead to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
 Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants;
 these results extend previous work on post-pandemic trajectories that focus primarily on cross-immunity \cite{kissler2020projecting,lavine2021immunological}.
-For example, while protection against symptoms unaccompanied by protection against transmission protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
+For example, while protection against symptoms protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
 A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \cite{koelle2022changing}.
@@
 % Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
-Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \cite{Bavel2020,chande2020real,sinclair2021pairing}, asymptomatic testing programs \cite{mina2021covid,gibson2022surv,Ranoa2022}, mask-wearing indoors and in crowded environments \cite{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \cite{allen2021indoor,wang2021airborne}) while improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
+Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \cite{Bavel2020,chande2020real,sinclair2021pairing}, asymptomatic testing programs \cite{mina2021covid,gibson2022surv,Ranoa2022}, mask-wearing indoors and in crowded environments \cite{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \cite{allen2021indoor,wang2021airborne}) while also improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
 Given the link between age and asymptomatic infections \cite{davies2020}, interventions may consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).

### 2022-09-08 main.tex
new file:   outputs/main_olddiff.pdf

@@
     \end{tabular}
-    \caption{Paramter descriptions and values for the basic asymptomatic model.}
+    \caption{Parameter descriptions and values for the basic asymptomatic model.}
     \label{tab:table1}
@@
     \end{tabular}
-    \caption{Paramter descriptions and values for the generalized asymptomatic model.}
+    \caption{Parameter descriptions and values for the generalized asymptomatic model.}
     \label{tab:table2}
@@
     \end{tabular}
-    \caption{Paramter descriptions and values for the asymptomatic model with immunity.}
+    \caption{Parameter descriptions and values for the asymptomatic model with immunity.}
     \label{tab:table3}

### 2022-09-11 main.tex
i think im done

@@
 Points represent each Ct measurement.
-Lines and shaded areas show LOESS fits.
-\jd{I don't really like calling them CIs. Maybe say default loess fits from R or something.}
+Lines and shaded areas show default LOESS fits from \texttt{ggplot2} \cite{ggplot2}.
 Data are publicly available in \cite{Kissler2020}.
@@
 In contrast, when $\delta$ is small enough that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
-\jd{Got rid of a few words that might imply that this is necessary as well as sufficient for a decreasing curve.}
 
@@
 (B--D) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$ (B), symptoms $\epsilon_s$ (C), and deaths $\epsilon_d$ (D).
-We simulate the model for 365 days, assuming $\beta_s = 4/5/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=1/2/\mathrm{day}$, $\gamma_s=\gamma_a=1/5/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
-\jd{Please change all fractions to decimals in above s.}
-The initial exposed proportion is $10^{-4}$.
+We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
+The initial infected proportion is $10^{-4}$.
 See Materials and Methods for model details and Supplementary Table S3 for parameter descriptions and values.
@@
 Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using parameters from \fref{base} with $\delta=0.8$.
-Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for \emph{either} a milder ($p=0.8$) or a not-milder ($p=0.2$) variant.
-\jd{not-milder is not an actual sugggestion\ldots}
+Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for \emph{either} a milder ($p=0.8$) or an equally severe ($p=0.2$) variant.
 (B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and milder (purple) and similar (orange) invading variants (dashed lines).
@@
 We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}A), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
-We consider two types of variants (which are simulated separately): one with the same severity $p$ (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
-\jd{The more severe variant has the same severity. There is no conceptual problem, but I'm worried about a reader-comprehension problem.}
+We consider two types of variants (which are simulated separately): one with the same severity $p$ as before (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
 

### 2022-12-22 response.tex
done

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

We are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
We have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
In doing so, we have tried to give directions for future areas of research.
We also perform sensitivity analyses to show that our qualitative conclusions are robust to asymptomatic transmissibility (Supplementary Figure S1--2).

Below please find our detailed responses to reviewers.

\rev{Reviewer \#1}

\revtext{Park et al. analyzed how the level of asymptomatic transmission relates to the total fatality at the population level in an infectious disease outbreak. At an individual level, the milder the disease (e.g. asymptomatic infection) the better; however, at the population level, milder disease (everything else being equal) could lead to a higher level of transmission and thus potentially a higher fatality. The authors developed mathematical models to quantitatively understand the conditions when this tradeoff occurs, and the implications of this tradeoff. They showed that when symptomatic individuals change behavior to reduce the number of contacts and when this reduction is sufficiently large, intermediate level of asymptomatic transmission (again, everything else being equal) could lead to the highest level of fatality.}

\revtext{Under this framework, the authors further explored the impact of vaccination on fatality assuming vaccinations protect against infection, symptoms or deaths. Broadly speaking, vaccination reduces fatality as expected under most scenarios, except when vaccination protects against symptoms only and the proportion of asymptomatic infection is low. (My assessment is that these exceptions are not directly relevant to the COVID-19 pandemic, because most (or all) vaccines protect against infection, fatality and transmission to some extent and the proportion of asymptomatic infection is high). Finally, the authors used the framework to analyze the dynamics of an invading variant, and showed that when immunity induced by an earlier strain provides protection against symptoms only (without protection for infection), a future variant can cause more rapid outbreaks and a high level of fatality.}

\revtext{Overall, I enjoy reading the manuscript and the careful mathematical analyses. The work proposes a novel and interesting theoretical idea on how reducing symptomatic infection only may lead to worse outcomes at the population level. The work is very well written, and the analyses are rigorous and very thoughtful. I do not have any specific suggestions to strengthen the manuscript.}

Thank you for your comments.

\revtext{However, while this is a nicely designed and conducted piece of work and the resulting theory is interesting, the work remains theoretical in nature.}

We agree that our work remains theoretical. We feel that applying this work in a broader context would be better suited as a separate paper. We have thus decided to transfer our paper to PNAS Nexus while trying our best to address both reviewers' comments within the scope of this paper.

\revtext{Because of a lack of direct quantification of the key values in the model (e.g. $\delta$ and epsilons) from actual data, it is unclear which parameter regime a particular phase of the COVID-19 pandemic lies.}

\revtext{Therefore, despite some general principles derived from the theory, it is difficult to apply the results of the work to identify drivers of specific epidemiological patterns of SARS-CoV-2 or to make specific recommendations for intervention strategies. Therefore, I do not feel the scope of the work is broad enough for PNAS.}

\revtext{Here below I list a few reasons for my assessment.}

\revtext{1. The work first makes the assumption that symptomatic individuals change their behavior (e.g., reducing contacts or increasing mask wearing) after symptom onset, and as a result, their transmission rate is decreased by a factor (modeled using parameter $\delta$). The authors show that when $\delta$ is large (i.e. >0.6), intermediate levels of asymptomatic transmission would lead to the highest levels of fatality at the population level. This observation is the basis for a lot of nonintuitive results in this study. However, the value of $\delta$ is unknown for SARS-CoV-2 unfortunately. It is almost impossible to precisely estimate this parameter (due to the difficulties identifying all transmission events). And this parameter is likely to be very variable over time and across different countries. The authors discussed lines of indirect evidence.}

\revtext{For example, the fraction of presymptomatic transmission sets the upper bound for $\delta$. One study estimated the fraction of presymptomatic transmission to be around 30\%-60\%. Another study by some of the authors of this study suggests the pre-symptomatic transmission may only account for 20\% of transmission. This suggests that the fraction of symptomatic transmission may be as high as 80\%. The key question is whether the change of behavior change or intervention would allow symptomatic individuals to reduce transmission by 60\% (given that the upper bound is 70\% or 80\%). To reach 60\%, that would mean that most of the potential transmission events are prevented after symptom onset for all the individuals in the population.}

Thank you for pointing this out. We realize that our explanations were unclear. We previously explained that the fraction of presymptomatic transmission sets the upper bound for $\delta$ for the main model:

``For this particular model, it does not make biological sense for $\delta$ to be greater than the amount of post-symptomatic transmission, because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.''

However, this is not necessarily the case for the generalized model, which includes both pre-symptomatic and asymptomatic transmission. 
To avoid confusion, we now write $\delta_s$ for the generalized model to capture the decrease in transmission only after symptom onset---the value of $\delta_s$ is independent of the fraction of presymptomatic transmission.
Throughout, we have tried to clarify our explanations and provide a concrete example:

``To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
In doing so, we reparameterize the model \textit{non-symptomatic} transmission, rather than the proportion of cases.
We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).

Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta_s > 0.6$ (Supplementary Figure S1; Table S2).
For example, when 40\% of non-symptomatic transmission is caused by pre-symptomatic transmission, a 60\% reduction in transmission after symptom onset is sufficient to drive the nonlinear effect of non-symptomatic transmission on epidemic fatality (peaking at around 10\% non-symptomatic transmission).
One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely more feasible than reaching $\delta = 0.6$.
For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
Given that the majority of isolation and tracing measures likely target symptomatic transmission, the amount of reduction in symptomatic transmission would be similar to these values, providing indirect support for the feasibility of high $\delta_s$ values. 
Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
We return to the discussion of values of $\delta$ later in the Discussion section.''

\revtext{This seems to be unlikely given other indirect data, e.g. the relatively high household secondary attack rate (SAR) for symptomatic individuals (e.g. see Madewell et al. JAMA Netw Open. 2022;5(4):e229317.). Of course, the SAR data does not completely rule out the possibility that $\delta>0.6$; however, the different indirect datasets highlight the difficulty in estimating $\delta$ and that the evidence for large $\delta$ is unclear.}

We have further tried to highlight the uncertainty in the values of $\delta$:

``The main conclusion of our analysis relies on having high values of $\delta$, the degree of transmission reduction among symptomatic individuals throughout the course of their infection.
For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values infeasible.
Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high values of $\delta_s$, the degree of reduction in transmission after symptom onset, epidemic fatalities peak at intermediate values of non-symptomatic transmission.
The general framework allows us to overcome the limitation of the simple model because $\delta_s$ is no longer bounded by the amount of presymptomatic transmission.
We have argued for the feasibility of high $\delta_s$ values based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness}, but it is uncertain what are the realistic ranges for $\delta_s$.
Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
During the exponential growth phase, $\delta$ is likely low because there is limited awareness for the outbreak and intervention measures in place.
As the number of cases increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings.''

\revtext{2. The results in Fig. 3 and 4 are very interesting. They showed under certain parameter regimes, vaccination (that prevents symptoms only) or a milder variant may not lead to lower fatality at the population level. Again, this result is under the assumption that $\delta>0.6$. The results could be useful for vaccine design and understanding variant dynamics.}

Thank you.

\revtext{However, as mentioned earlier, most (or all) vaccines most (or all) vaccines protect against infection, fatality and transmission to some extent. In addition, without thorough analyses of experimental or epidemiological data to estimate the relevant parameters for SARS-CoV-2 variants (e.g. Delta vs. Omicron as mentioned by the authors), it is unclear how the theory relates to the observed impacts of vaccination and the selection of variants on SARS-CoV-2 dynamics.}

We have tried to highlight the uncertainty throughout the paper. 
We have also added the following paragraph in the discussion section to provide guidance for future work:

``Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving vaccination and evolutionary dynamics of SARS-CoV-2 is yet unclear.
Our study provides rationale for exploring these questions in detail.''

\revtext{Overall, the work is a very interesting and well-conducted theoretical work; however, its utility in understanding SARS-CoV-2 epidemiology or design of intervention strategies is limited.}

\rev{Reviewer \#2}

\revtext{Park et al present a very well developed epidemiological model focused on the interplay between asymptomatic transmission and symptom-responsive transmission reduction due to behavioral change or mitigation. The model is well developed and the analysis rigorous, in as far as it goes and overall everything is very well described and written. I found the key predictions - that intermediate asymmetric (and pre-symptomatic) transmission has the greatest population level impact when there is mitigation and behavioral change for symptomatic transmission and all of the outcomes in the abstract to be pretty intuitive. The detailed analysis in the main paper are very useful and the paper is at its most interesting when it is making more specific predictions.}

\revtext{I would focus on the level of transmission blocking that is important to mitigate the population level effects of asymptomatic spread, rather than simply stating that modest effects can do this.}

\revtext{My sense is that it is not at all surprising that anti-disease immunity CAN lead to greater population level effects, what is more interesting is when it does. In particular, this could be related to the specific challenges of the Sars-Cov-2 pandemic. The authors have the models to easily get these results and this refocusing will improve the impact of the paper. The discussion of the implications of the model to our understanding of new variants and in particular Omicron were very interesting and informative. Overall, I think that the manuscript would have a greater impact with a rewrite that focusses less on the general predictions, since these are not surprising and more on the detailed drivers of the results and how these are illustrated by SARS-Cov-2. I have some specific criticisms.}

We feel that making detailed predictions are beyond the scope of this paper. 
While immunity profiles have been discussed in other contexts (such as the Gandon et al paper as suggested by the same reviewer), there has been a limited discussion around the topic from epidemiological and public health perspectives.
Therefore, we wanted to focus on qualitative effects of asymptomatic transmission on epidemic fatalities in this paper, while leaving quantitative predictions for the future work.
Throughout the paper, we have tried to highlight the theoretical nature of our work and give guidance for future work.
In particular, we have added the following paragraph in the discussion section:

``Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving vaccination and evolutionary dynamics of SARS-CoV-2 is yet unclear.
Our study provides rationale for exploring these questions in detail.''

\revtext{My main question about the results comes from the assumption on Line 117 - how important is the 25\% reduction in asymptomatic transmission assumption? - I was surprised that given the uncertainty that they acknowledge, the authors didn't explore this. My sense is that this is a high assumption for asymptomatic transmission, but some exploration of the importance of this assumption to the outcomes is important. Clearly, this will not impact the general predictions, but if they focus the paper on more specific scenarios, it may have an impact and to my mind needs to be explored.}

We now perform sensitivity analyses to account for this uncertainty:

``We note that these results are robust to uncertainties in asymptomatic transmission rate.
In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
In this case, qualitative predictions are robust except for when $\RR_a = 0.25 \RR_s$.
In this case, high values of $\delta$ leads to $\RR_0 < 1$ and the epidemic does not take off.
In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
In this case, we find that intermediate values of asymptomaticity leads to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.''

\revtext{As a general criticism, the focus on immunity profiles and whether immunity is symptom blocking is interesting and novel in this context, but it is not surprising for reviewers who are familiar with the evolution of tolerance literature. In particular, there should be a better link to the Gandon et al paper on the impact of anti-disease vaccines on epidemiological as well as evolutionary outcomes. Clearly the focus of the Gandon paper is different, but they have comparable insights of the impact on epidemiology of anti-diseases (symptoms) immunity compared to transmission blocking ones. I think this literature should be acknowledged, not least because it makes many of the same predictions.}

We now discuss similarities and differences between our work and Gandon et al paper:

``Our conclusions echo earlier findings by \cite{gandon2001imperfect}, who showed that vaccine-derived immunity against diseases can promote the evolution of more virulent strains in unvaccinated individuals.
By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to the same effect even in the absence of evolution.''

\revtext{Line 58 - "subclinical or even asymptomatic' is imprecise and confusing}

Thank you for pointing this out. We have removed ``or even asymptomatic''

\revtext{Paragraph at line 59 - is the diamond princess data still the best data we have? It would be useful to emphasize that the asymptomatic transmission is very well established.}

Since we are focusing on the asyptomatic transmission of primary infection, we feel that the diamond princess data is one of the best data we have. We have tried to clarify this point while addressing other existing data on asymptomatic transmission:

``It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility of immunologically naive asymptomatic individuals.''

\revtext{Paragraph line 69 - It would be useful to discuss how important the symptoms may be for transmission here - coughing, sneezing etc may of course have a direct impact on transmission and it can be this rather than differences in viral loads are likely to be important.}

Thank you for pointing this out. We added the following:

``Notwithstanding similarities in the viral load dynamics, the presence (or the absence) of symptoms further contributes to the uncertainty in the differences in transmissibility of symptomatic and asymptomatic individuals.
In general, coughing and sneezing can help deliver virus containing droplets, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests speech droplets can also be an important mode of transmission.''

\bibliography{main}

\end{document}

### 2022-12-22 main.tex
change bib style
done

@@
 
-% \usepackage{natbib}
-\bibliographystyle{Science}
+\usepackage{natbib}
+\bibliographystyle{unsrt}
 \date{\today}
@@
 This means that more than 99\% of individuals infected with COVID-19 will survive. 
-Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical or even asymptomatic.
+Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical.
 
@@
 The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained by the analysis, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
+It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
+In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility of immunologically naive asymptomatic individuals.
 
@@
 however, differences between inferred \emph{total} viral load from Ct values and \emph{infectious} viral load add uncertainty \cite{romero2021ct}.
+Notwithstanding similarities in the viral load dynamics, the presence (or the absence) of symptoms further contributes to the uncertainty in the differences in transmissibility of symptomatic and asymptomatic individuals.
+In general, coughing and sneezing can help deliver virus containing droplets, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests speech droplets can also be an important mode of transmission \citep{stadnytskyi2021breathing}.
+
 We note also that asymptomaticity can change across outbreak settings \cite{alene2021magnitude}.
@@
 In contrast, when $\delta$ is small enough that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
+We note that these results are robust to uncertainties in asymptomatic transmission rate.
+In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
+In this case, qualitative predictions are robust except for when $\RR_a = 0.25 \RR_s$.
+In this case, high values of $\delta$ leads to $\RR_0 < 1$ and the epidemic does not take off.
+In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
+In this case, we find that intermediate values of asymptomaticity leads to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
 
@@
 
-We therefore extend our model to consider the effects of generalized \textit{non-symptomatic} transmission, which includes both pre-symptomatic and asymptomatic transmission.
-For this model, we assume that $\delta$ decreases transmission only after symptom onset.
+To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
+In doing so, we reparameterize the model \textit{non-symptomatic} transmission, rather than the proportion of cases.
+We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
 We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).
 
-Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta > 0.6$ (Supplementary Figure S1; Table S2).
+Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta_s > 0.6$ (Supplementary Figure S3; Table S2).
+For example, when 40\% of non-symptomatic transmission is caused by pre-symptomatic transmission, a 60\% reduction in transmission after symptom onset is sufficient to drive the nonlinear effect of non-symptomatic transmission on epidemic fatality (peaking at around 10\% non-symptomatic transmission).
 One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
+While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely more feasible than reaching $\delta = 0.6$.
+For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
+Given that the majority of isolation and tracing measures likely target symptomatic transmission, the amount of reduction in symptomatic transmission would be similar to these values, providing indirect support for the feasibility of high $\delta_s$ values.
 Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
+We return to the discussion of values of $\delta$ later in the Discussion section.
 
@@
 
+The main conclusion of our analysis relies on having high values of $\delta$, the degree of transmission reduction among symptomatic individuals throughout the course of their infection.
+For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values infeasible.
+Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high values of $\delta_s$, the degree of reduction in transmission after symptom onset, epidemic fatalities peak at intermediate values of non-symptomatic transmission.
+The general framework allows us to overcome the limitation of the simple model because $\delta_s$ is no longer bounded by the amount of presymptomatic transmission.
+We have argued for the feasibility of high $\delta_s$ values based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness}, but it is uncertain what are the realistic ranges for $\delta_s$.
+Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
+During the exponential growth phase, $\delta$ is likely low because there is limited awareness for the outbreak and intervention measures in place.
+As the number of cases increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
+A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings.
+
 Via theory and simulation analysis of a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can under some conditions lead to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
@@
 A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \cite{koelle2022changing}.
+Our conclusions echo earlier findings by \cite{gandon2001imperfect}, who showed that vaccine-derived immunity against diseases can promote the evolution of more virulent strains in unvaccinated individuals.
+By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to the same effect even in the absence of evolution.
+
+Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
+Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
+Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
+We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving vaccination and evolutionary dynamics of SARS-CoV-2 is yet unclear.
+Our study provides rationale for exploring these questions in detail.
 
@@
 \dot{S} &= -\beta_p I_p -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \nu E\\
+\dot{E} &= \beta_a S I_a + (1-\delta_s) \beta_s S I_s - \nu E\\
 \dot{I}_p &= \nu E - \sigma I_p\\
@@
       \hline
-      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
+      $\delta_s$ & Reduction in symptomatic transmission rate & 0--1\\
     \end{tabular}
@@
 \begin{center}
+\includegraphics[width=\textwidth]{figure_base_sens.ggp.pdf}
+\caption{
+\textbf{Simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction for a wide range of asymptomatic transmissibility.}
+Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
+We allow the ratios between the asymptomatic transmission rate $\beta_a$ and symptomatic transmission rate $\beta_s$ to vary between 0.25 and 1.
+See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
+}
+\end{center}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\textwidth]{figure_base_sens_fixR0.ggp.pdf}
+\caption{
+\textbf{Simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction for a wide range of asymptomatic transmissibility and a fixed $\mathcal{R}_0$ value at intermediate asymptomaticity.}
+Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
+We simulate the model for 365 days, assuming $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
+We allow the ratios between the asymptomatic transmission rate $\beta_a$ and symptomatic transmission rate $\beta_s$ to vary between 0.25 and 1.
+We also fix the basic reproduction number $\mathcal{R}_0=4$ when there are intermediate levels of asymptomaticity $p=0.5$ and no reduction in symptomatic transmission rate $\delta=0$.
+See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
+}
+\end{center}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!ht]
+\begin{center}
 \includegraphics[width=0.8\textwidth]{diagram_sub.pdf}
@@
 See Methods for model details.
-(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
+(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta_s$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
 See Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values.

### 2022-12-22 figure_base_sens.R

@@ -1,0 +1,38 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
+library(egg)
+library(shellpipes)
+
+simulate_base_sens <- rdsRead()
+
+simulate_base_max <- simulate_base_sens %>%
+  group_by(sim) %>%
+  filter(
+    time==max(time)
+  ) %>%
+  mutate(
+    beta_a=factor(beta_a, 
+                  levels=c(0.2, 0.4, 0.6, 0.8),
+                  labels=c(expression(beta[a]==0.25~beta[s],
+                                      beta[a]==0.5~beta[s],
+                                      beta[a]==0.75~beta[s],
+                                      beta[a]==beta[s])))
+  )
+
+g1 <- ggplot(simulate_base_max) +
+  geom_line(aes(p, D_n, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, 0.01),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  facet_wrap(~beta_a, labeller = label_parsed) +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "right"
+  )
+
+saveGG(g1, width=6, height=4)

### 2022-12-22 figure_base_sens_fixR0.R

@@ -1,0 +1,38 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_family="Times", base_size = 14))
+library(egg)
+library(shellpipes)
+
+simulate_base_sens_fixR0 <- rdsRead()
+
+simulate_base_max <- simulate_base_sens_fixR0 %>%
+  group_by(sim) %>%
+  filter(
+    time==max(time)
+  ) %>%
+  mutate(
+    beta_a_ratio=factor(beta_a_ratio, 
+                  levels=c(0.25, 0.5, 0.75, 1),
+                  labels=c(expression(beta[a]==0.25~beta[s],
+                                      beta[a]==0.5~beta[s],
+                                      beta[a]==0.75~beta[s],
+                                      beta[a]==beta[s])))
+  )
+
+g1 <- ggplot(simulate_base_max) +
+  geom_line(aes(p, D_n, col=factor(delta), group=factor(delta))) +
+  scale_x_continuous("Proportion asymptomatic", 
+                     breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1),
+                     labels=c("0", "0.2", "0.4", "0.6", "0.8", "1"),
+                     expand=c(0, 0)) +
+  scale_y_continuous("Proportion deceased",
+                     limits=c(0, 0.01),
+                     expand=c(0, 0)) +
+  scale_color_viridis_d(expression(delta)) +
+  facet_wrap(~beta_a_ratio, labeller = label_parsed) +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "right"
+  )
+
+saveGG(g1, width=6, height=4)

### 2022-12-22 figure_sub.R

@@ -29,7 +29,7 @@ g1 <- ggplot(simulate_sub_max) +
   scale_y_continuous("Proportion deceased",
                      limits=c(0, 0.0103),
                      expand=c(0, 0)) +
-  scale_color_viridis_d(expression(delta)) +
+  scale_color_viridis_d(expression(delta[s])) +
   facet_wrap(~p_pre_ratio, labeller = label_parsed) +
   ggtitle("B") +
   theme(

### 2022-12-22 simulate_base_sens.R

@@ -1,0 +1,21 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+p <- seq(0, 1, length.out=101)
+delta <- 0:5 * 0.2
+beta_a <- c(0.25, 0.5, 0.75, 1) * 4/5
+
+simparam <- expand.grid(p, delta, beta_a)
+names(simparam) <- c("p", "delta", "beta_a")
+
+simulation_base_sens <- apply(simparam, 1, function(x) {
+  out <- do.call(simulate, c(as.list(x), S=1, D0=0))
+  
+  as.data.frame(append(out, x))
+}) %>% bind_rows(.id="sim")
+
+rdsSave(simulation_base_sens)

### 2022-12-22 simulate_base_sens_fixR0.R

@@ -1,0 +1,34 @@
+library(deSolve)
+library(dplyr)
+
+library(shellpipes)
+
+loadEnvironments()
+
+R0 <- 4
+pbase <- 0.5
+p <- seq(0, 1, length.out=101)
+delta <- 0:5 * 0.2
+beta_a_ratio <- c(0.25, 0.5, 0.75, 1)
+
+simparam <- expand.grid(p, delta, beta_a_ratio)
+names(simparam) <- c("p", "delta", "beta_a_ratio")
+
+simulation_base_sens_fixR0 <- apply(simparam, 1, function(x) {
+  baratio <- x[[3]]
+  
+  beta_s <- R0/(1 + baratio)/5*2
+  
+  param <- list(
+    p=x[[1]],
+    delta=x[[2]],
+    beta_s=beta_s,
+    beta_a=beta_s*x[[3]]
+  )
+  
+  out <- do.call(simulate, c(param, S=1, D0=0))
+  
+  as.data.frame(append(out, x))
+}) %>% bind_rows(.id="sim")
+
+rdsSave(simulation_base_sens_fixR0)

### 2022-12-26 response.tex
Once through

@@
 
-We are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
+Following an editorial suggesting from PNAS, we are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
 We have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
@@
 In doing so, we reparameterize the model \textit{non-symptomatic} transmission, rather than the proportion of cases.
+\jd{Something garbled here.}
 We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
@@
 ``Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
-Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+Extending our work to real outbreak scenarios will require \jd{Omit?: developing finely tuned models and} narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
 For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.

### 2022-12-26 main.tex
Once through

@@
 however, differences between inferred \emph{total} viral load from Ct values and \emph{infectious} viral load add uncertainty \cite{romero2021ct}.
-Notwithstanding similarities in the viral load dynamics, the presence (or the absence) of symptoms further contributes to the uncertainty in the differences in transmissibility of symptomatic and asymptomatic individuals.
-In general, coughing and sneezing can help deliver virus containing droplets, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests speech droplets can also be an important mode of transmission \citep{stadnytskyi2021breathing}.
+The likely role of symptoms in transmission further contributes to this uncertainty.
+Coughing and sneezing can help deliver virus-containing droplets.
+However, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests that speech droplets can also be an important mode of transmission \citep{stadnytskyi2021breathing}.
 
@@
 In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
-In this case, qualitative predictions are robust except for when $\RR_a = 0.25 \RR_s$.
-In this case, high values of $\delta$ leads to $\RR_0 < 1$ and the epidemic does not take off.
+In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$.
+-- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
 In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
-In this case, we find that intermediate values of asymptomaticity leads to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
+In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
 
@@
 Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission could have been as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
-There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission makes high $\delta$ values at least somewhat more likely during the initial pandemic phase; and second, intermediate levels of behavioral effects ($\delta > 0$) would have been already present early in the pandemic to reduce the proportion of pre-symptomatic transmission from 80\% to as low as 40\%.
+There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission makes high $\delta$ values at least somewhat more likely during the initial pandemic phase; and second, substantial levels of behavioral effects ($\delta > 0$) may have been present early in the pandemic (to reduce the proportion of symptomatic transmission from 80\% to as low as 40\%).
 
 To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
-In doing so, we reparameterize the model \textit{non-symptomatic} transmission, rather than the proportion of cases.
+To do so, we reparameterize the model based on relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of of asymptomatic cases.
 We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
@@
 For example, when 40\% of non-symptomatic transmission is caused by pre-symptomatic transmission, a 60\% reduction in transmission after symptom onset is sufficient to drive the nonlinear effect of non-symptomatic transmission on epidemic fatality (peaking at around 10\% non-symptomatic transmission).
-One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
+One exception is the extreme case in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there is no asymptomatic transmission);
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely more feasible than reaching $\delta = 0.6$.
+While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely \jd{in fact, tautologically} more feasible than reaching $\delta = 0.6$.
+\jd{I think we can just say something like “it does not seem unrealistic”.}
 For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
@@
 
-The main conclusion of our analysis relies on having high values of $\delta$, the degree of transmission reduction among symptomatic individuals throughout the course of their infection.
-For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values infeasible.
-Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high values of $\delta_s$, the degree of reduction in transmission after symptom onset, epidemic fatalities peak at intermediate values of non-symptomatic transmission.
-The general framework allows us to overcome the limitation of the simple model because $\delta_s$ is no longer bounded by the amount of presymptomatic transmission.
-We have argued for the feasibility of high $\delta_s$ values based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness}, but it is uncertain what are the realistic ranges for $\delta_s$.
+The main conclusion of our analysis relies on having high levels of transmission reduction among symptomatic individuals ($\delta$). 
+For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values unrealistic for COVID-19.
+Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
+Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} we believe that high levels are $\delta_s$ are plausible, but a great deal of uncertainty remains. 
 Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
@@
 
-Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
-Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+Our work further illustrates the potential for asymptomatic infection and transmission to make population-level outcomes worse, but remains primarily theoretical.
+Applying these ideas to real outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
 For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
 Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
-We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving vaccination and evolutionary dynamics of SARS-CoV-2 is yet unclear.
-Our study provides rationale for exploring these questions in detail.
+We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving \jd{What?: vaccination and} evolutionary dynamics of SARS-CoV-2 is yet unclear.
+Our study provides a starting point for exploring these questions in more detail.
 

### 2022-12-29 response.tex
updating

--- 2022-12-22 response.tex
@@
 ``To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
-In doing so, we reparameterize the model \textit{non-symptomatic} transmission, rather than the proportion of cases.
+To do so, we reparameterize the model based on relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of of asymptomatic cases.
 We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
@@
 
-Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta_s > 0.6$ (Supplementary Figure S1; Table S2).
+Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta_s > 0.6$ (Supplementary Figure S3; Table S2).
 For example, when 40\% of non-symptomatic transmission is caused by pre-symptomatic transmission, a 60\% reduction in transmission after symptom onset is sufficient to drive the nonlinear effect of non-symptomatic transmission on epidemic fatality (peaking at around 10\% non-symptomatic transmission).
-One exception is the extreme (and unrealistic) case, in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there are no asymptomatic cases);
+One exception is the extreme case in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there is no asymptomatic transmission);
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely more feasible than reaching $\delta = 0.6$.
+While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it does not seem unrealistic.
 For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
@@
 
-``The main conclusion of our analysis relies on having high values of $\delta$, the degree of transmission reduction among symptomatic individuals throughout the course of their infection.
-For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values infeasible.
-Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high values of $\delta_s$, the degree of reduction in transmission after symptom onset, epidemic fatalities peak at intermediate values of non-symptomatic transmission.
-The general framework allows us to overcome the limitation of the simple model because $\delta_s$ is no longer bounded by the amount of presymptomatic transmission.
-We have argued for the feasibility of high $\delta_s$ values based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness}, but it is uncertain what are the realistic ranges for $\delta_s$.
+``The main conclusion of our analysis relies on having high levels of transmission reduction among symptomatic individuals ($\delta$). 
+For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values unrealistic for COVID-19.
+Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
+Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} we believe that high levels are $\delta_s$ are plausible, but a great deal of uncertainty remains. 
 Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
@@
 As the number of cases increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
-A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings.''
+A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings. ''
 
@@
 
-``Our work remains largely theoretical in illustrating the potential for asymptomatic infection and transmission to make population-level outcomes worse.
-Extending our work to real outbreak scenarios will require developing finely tuned models and narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+``Our work further illustrates the potential for asymptomatic infection and transmission to make population-level outcomes worse, but remains primarily theoretical.
+Applying these ideas to real outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
 For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
 Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
-We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving vaccination and evolutionary dynamics of SARS-CoV-2 is yet unclear.
-Our study provides rationale for exploring these questions in detail.''
+We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
+Our study provides a starting point for exploring these questions in more detail.''
 
@@
 In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
-In this case, qualitative predictions are robust except for when $\RR_a = 0.25 \RR_s$.
-In this case, high values of $\delta$ leads to $\RR_0 < 1$ and the epidemic does not take off.
+In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$.
+--- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
 In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
-In this case, we find that intermediate values of asymptomaticity leads to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.''
+In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.''
 
@@
 
-``Notwithstanding similarities in the viral load dynamics, the presence (or the absence) of symptoms further contributes to the uncertainty in the differences in transmissibility of symptomatic and asymptomatic individuals.
-In general, coughing and sneezing can help deliver virus containing droplets, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests speech droplets can also be an important mode of transmission.''
+``The likely role of symptoms in transmission further contributes to this uncertainty.
+Coughing and sneezing can help deliver virus-containing droplets.
+However, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests that speech droplets can also be an important mode of transmission \citep{stadnytskyi2021breathing}.''
 

### 2022-12-29 main.tex
updating

@@
 In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$.
--- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
+--- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
 In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
@@
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is likely \jd{in fact, tautologically} more feasible than reaching $\delta = 0.6$.
-\jd{I think we can just say something like “it does not seem unrealistic”.}
+While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it does not seem unrealistic.
 For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
@@
 Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
-We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving \jd{What?: vaccination and} evolutionary dynamics of SARS-CoV-2 is yet unclear.
+We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
 Our study provides a starting point for exploring these questions in more detail.

### 2022-12-29 response.tex
very minor

@@
 
-Following an editorial suggesting from PNAS, we are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
+Following an editorial suggestion from PNAS, we are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
 We have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
@@
 For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values unrealistic for COVID-19.
-Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
+Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($\delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
 Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} we believe that high levels are $\delta_s$ are plausible, but a great deal of uncertainty remains. 
@@
 
-Since we are focusing on the asyptomatic transmission of primary infection, we feel that the diamond princess data is one of the best data we have.
+Since we are focusing on the asymptomatic transmission of primary infection, we do feel that the diamond princess data is one of the best data we have.
 We have tried to clarify this point while addressing other existing data on asymptomatic transmission:
@@
 ``It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
-In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility of immunologically naive asymptomatic individuals.''
+In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility from immunologically naive asymptomatic individuals.''
 

### 2022-12-29 main.tex
very minor

@@
 It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
-In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility of immunologically naive asymptomatic individuals.
+In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility from immunologically naive asymptomatic individuals.
 

### 2022-12-29 main.tex
Update main.tex by JSW
Minor updates, re-organized the Diamond Princess justification and tried to match response letter to edits.

@@
 
-Early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2;
-the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \cite{mizumoto2020estimating}.
+It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but asymptomatic cases are increasingly shaped by prior immunity (whether through infection, vaccination, or both).
+In contrast, early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2 from and to immunologically naive individuals; the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \cite{mizumoto2020estimating}.
 It has been estimated that 75\% (95\% CI: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
 The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained by the analysis, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
-It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
-In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility from immunologically naive asymptomatic individuals.
 
@@
 In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
-In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$.
---- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
-In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
+In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$
+--- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large then $\RR_0 < 1$ and the epidemic does not take off.
+In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing x $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
 In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
@@
 To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
-To do so, we reparameterize the model based on relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of of asymptomatic cases.
+To do so, we reparameterize the model based on the relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of asymptomatic cases.
 We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
@@
 in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it does not seem unrealistic.
-For example, \cite{kucharski2020effectiveness} estimated that up to 64\% reduction in transmission would be possible under self-isolation and -quarantine as well as manual contact tracing of all contacts (and as low as 47\% reduction using digital contact tracing without manual tracing).
+While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is plausible given behavioural and policy responses.
+For example, \cite{kucharski2020effectiveness} estimated that up to $\sim$64\% reduction in transmission would be possible under self-isolation and self-quarantine as well as manual contact tracing (and as low as $\sim$47\% reduction using digital contact tracing without manual tracing).
 Given that the majority of isolation and tracing measures likely target symptomatic transmission, the amount of reduction in symptomatic transmission would be similar to these values, providing indirect support for the feasibility of high $\delta_s$ values.
@@
 Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($\delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
-Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} we believe that high levels are $\delta_s$ are plausible, but a great deal of uncertainty remains. 
-Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
+Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} high levels are $\delta_s$ are plausible, but uncertainty remains. 
+Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of an epidemic.
 During the exponential growth phase, $\delta$ is likely low because there is limited awareness for the outbreak and intervention measures in place.
-As the number of cases increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
-A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings.
+As the number of cases, hospitalizations, and fatalities increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
+Further analysis is needed to constrain the uncertainty in $\delta$ and to apply the model framework in distinct disease, policy, and socioeconomic contexts.
 
@@
 Our conclusions echo earlier findings by \cite{gandon2001imperfect}, who showed that vaccine-derived immunity against diseases can promote the evolution of more virulent strains in unvaccinated individuals.
-By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to the same effect even in the absence of evolution.
+By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to a similar effect even in the absence of evolution.
 
-Our work further illustrates the potential for asymptomatic infection and transmission to make population-level outcomes worse, but remains primarily theoretical.
-Applying these ideas to real outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
-For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
-Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
+Our work uses theoretical model to illustrate the potential for asymptomatic infection and transmission to make population-level outcomes worse.
+Applying these ideas to specific outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+For example, we simulated unmitigated outbreaks with fixed parameters but epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
+Calibrating the model to outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
 We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
@@
 Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
-As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \cite{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of the pandemic.
+As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \cite{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of this and future pandemics.
 

### 2022-12-30 response.tex
done

@@
 Following an editorial suggestion from PNAS, we are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
-We have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
-In doing so, we have tried to give directions for future areas of research.
-We also perform sensitivity analyses to show that our qualitative conclusions are robust to asymptomatic transmissibility (Supplementary Figure S1--2).
+Our submission builds on multiple years of empirical evidence that asymptomatic transmission of SARS-CoV-2 impacts pandemic controllability and severity. 
+Here, we provide and apply a significant refinement of the underlying theory, linking asymptomatic transmission at individual scales to population-level epidemic outcomes. 
+In doing so, our theory reveals how asymptomatic infection and transmission presents a double-edged sword.
+
+While asymptomatic infections generally improve individual-level outcomes (by preventing severe cases), they can also make control more difficult.
+Foundational analyses of asymptomatic infections have primarily focused on the latter effect – neglecting the effects of asymptomatic spread on population-level outcomes, such as overall fatalities (Fraser et al., 2004).
+
+Here, we use a series of epidemic models to assess the role of asymptomatic spread on epidemic fatalities in the presence of transmission-reducing effects among symptomatic individuals (e.g., via behavioural change or symptom-based interventions).
+We find that intermediate levels of asymptomaticity lead to the highest levels of epidemic fatalities over a wide range of parameters. 
+This counterintuitive result has not yet been demonstrated previously, nonetheless it emerges as a result of the countervailing forces of increased cases and decreasing severity with increasing asymptomaticity.
+We then extend our framework to assess the ways in which immunity profiles can lead to potentially paradoxical outcomes: with less virulent strains spreading precisely because asymptomatic individuals can transmit while unaware they are infected.
+Moreover, our findings show that emergence of milder variants at the individual scale can nonetheless cause similar or even higher levels of fatalities at population scales – mirroring many features of the Omicron variant.
+
+Following the advice of the two PNAS reviewers, we have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
+We also perform sensitivity analyses to show that our qualitative conclusions are robust to parameter variation associated with asymptomatic transmissibility (Supplementary Figure S1--2).
+Finally, we provide directions for future areas of research that can assess both the impact of asymptomatic transmission and the extent to which control can improve population-level outcomes.
 
 Below please find our detailed responses to reviewers.
+
+Sincerely,
+Joshua S. Weitz
+
+\pagebreak
 
@@
 
-We agree that our work remains theoretical.
+We agree that our work remains theoretical --- and that PNAS \& PNAS Nexus are both forums for compelling theoretical work.
 We feel that applying this work in a broader context would be better suited as a separate paper.
-We have thus decided to transfer our paper to PNAS Nexus while trying our best to address both reviewers' comments within the scope of this paper.
+We have thus decided to transfer our paper to PNAS Nexus while addressing both reviewers' comments within the scope of this paper.
 
@@
 
-Thank you for pointing this out.
-We realize that our explanations were unclear.
+Thank you for pointing out a potential misunderstanding with respect to the value of $\delta$.
 We previously explained that the fraction of presymptomatic transmission sets the upper bound for $\delta$ for the main model:
@@
 ``To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
-To do so, we reparameterize the model based on relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of of asymptomatic cases.
+To do so, we reparameterize the model based on the relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of asymptomatic cases.
 We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
@@
 Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($\delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
-Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} we believe that high levels are $\delta_s$ are plausible, but a great deal of uncertainty remains. 
-Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of a real epidemic.
+Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} high levels are $\delta_s$ are plausible, but uncertainty remains. 
+Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of an epidemic.
 During the exponential growth phase, $\delta$ is likely low because there is limited awareness for the outbreak and intervention measures in place.
-As the number of cases increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
-A detailed analysis is needed to better constrain the uncertainties around the values of $\delta$ and to understand the relevance of our work in realistic settings. ''
+As the number of cases, hospitalizations, and fatalities increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
+Further analysis is needed to constrain the uncertainty in $\delta$ and to apply the model framework in distinct disease, policy, and socioeconomic contexts. ''
 
@@
 
-We have tried to highlight the uncertainty throughout the paper. 
+We have tried to highlight the uncertainty in parameters and robustness of results to parameter uncertainty throughout the paper. 
 We have also added the following paragraph in the discussion section to provide guidance for future work:
 
-``Our work further illustrates the potential for asymptomatic infection and transmission to make population-level outcomes worse, but remains primarily theoretical.
-Applying these ideas to real outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
-For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
-Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
+``Our work uses theoretical model to illustrate the potential for asymptomatic infection and transmission to make population-level outcomes worse.
+Applying these ideas to specific outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+For example, we simulated unmitigated outbreaks with fixed parameters but epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
+Calibrating the model to outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
 We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
@@
 \revtext{Overall, the work is a very interesting and well-conducted theoretical work; however, its utility in understanding SARS-CoV-2 epidemiology or design of intervention strategies is limited.}
+
+We hope that our revisions clarify the generalized nature of our findings and also point the way towards practical steps to link individual and population-level outcomes associated with asymptomatic transmission in realistic contexts.
+This process will require moving beyond epidemic models to include behaviour and policy responses.  
+Overall, we believe that the revised work fits the PNAS Nexus mission of welcoming rigorous research with broad, interdisciplinary appeal.
 
@@
 
-We feel that making detailed predictions are beyond the scope of this paper. 
+We appreciate the suggestion of a rewrite but note that we are unaware of another paper that makes the explicit link between intermediate levels of asymptomatic transmission and worst-case population-level outcomes.
+The fact that our models are simple to understand and 'easily' get these results seems to us to be a positive reflection of our choice of a minimal set of features that nonetheless lead to a robust gap between individual and population level outcomes.
+We hope that this model embodies the spirit of 'Everything is obvious: once you know the answer' - the book on discovery and common sense by Duncan Watts from >10 years ago. 
+Turning back to early 2020, it was certainly not clear to (most) policy makers and (most) models that asymptomatic transmission and control could be a significant driver of improving population level outcomes.
+Hence, we have decided to leave the core structure in place. 
+In addition, we feel that making detailed predictions are beyond the scope of this paper.
+
 While immunity profiles have been discussed in other contexts (such as the Gandon et al paper as suggested by the same reviewer), there has been a limited discussion around the topic from epidemiological and public health perspectives.
@@
 
-``Our work further illustrates the potential for asymptomatic infection and transmission to make population-level outcomes worse, but remains primarily theoretical.
-Applying these ideas to real outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
-For example, we simulated unmitigated outbreaks with assumed parameters, but real epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
-Calibrating the model to real outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
+``Our work uses theoretical model to illustrate the potential for asymptomatic infection and transmission to make population-level outcomes worse.
+Applying these ideas to specific outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
+For example, we simulated unmitigated outbreaks with fixed parameters but epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
+Calibrating the model to outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
 We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
@@
 ``We note that these results are robust to uncertainties in asymptomatic transmission rate.
-In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
-In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$.
---- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large, $\RR_0 < 1$ and the epidemic does not take off.
-In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 but now we fix $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
+In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_a$ and $\RR_s$ between 0.25 and 1.
+In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$
+--- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large then $\RR_0 < 1$ and the epidemic does not take off.
+In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
 In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.''
@@
 ``Our conclusions echo earlier findings by \cite{gandon2001imperfect}, who showed that vaccine-derived immunity against diseases can promote the evolution of more virulent strains in unvaccinated individuals.
-By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to the same effect even in the absence of evolution.''
+By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to a similar effect even in the absence of evolution.
+''
 
@@
 
-``It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but the majority of asymptomatic cases are now driven by prior immunity.
-In contrast, the Diamond Princess data provide a unique perspective on the relative transmissibility from immunologically naive asymptomatic individuals.''
+``It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but asymptomatic cases are increasingly shaped by prior immunity (whether through infection, vaccination, or both).
+In contrast, early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2 from and to immunologically naive individuals ...''
 

### 2022-12-30 main.tex
done

@@
 We note that these results are robust to uncertainties in asymptomatic transmission rate.
-In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_s$ and $\RR_a$ between 0.25 and 1.
+In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_a$ and $\RR_s$ between 0.25 and 1.
 In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$
 --- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large then $\RR_0 < 1$ and the epidemic does not take off.
-In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing x $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
+In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
 In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.

### 2022-12-30 response.tex
Update response.tex
Trying to simplify and make the point that we made the changes reviewers wanted (not sure how to do spellcheck on Github)

@@
 Following an editorial suggestion from PNAS, we are transferring our manuscript ``Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities'' for consideration for publication in PNAS Nexus.
-Our submission builds on multiple years of empirical evidence that asymptomatic transmission of SARS-CoV-2 impacts pandemic controllability and severity. 
-Here, we provide and apply a significant refinement of the underlying theory, linking asymptomatic transmission at individual scales to population-level epidemic outcomes. 
-In doing so, our theory reveals how asymptomatic infection and transmission presents a double-edged sword.
+Our revised submission builds on multiple years of empirical evidence that asymptomatic transmission of SARS-CoV-2 impacts pandemic controllability and severity. 
+Here, we provide and apply a significant refinement of the associated theory, linking asymptomatic transmission at individual scales to population-level epidemic outcomes.
 
-While asymptomatic infections generally improve individual-level outcomes (by preventing severe cases), they can also make control more difficult.
-Foundational analyses of asymptomatic infections have primarily focused on the latter effect – neglecting the effects of asymptomatic spread on population-level outcomes, such as overall fatalities (Fraser et al., 2004).
+Consistent with PNAS reviewer suggestions, we have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions.
+Notably, we perform sensitivity analyses to show that our central conclusions are robust to parameter variation associated with asymptomatic transmissibility (Supplementary Figure S1--2).
+We also provide directions for future areas of research that can assess both the impact of asymptomatic transmission and the extent to which control can improve population-level outcomes.
+Our original conclusions remain: we find that intermediate levels of asymptomaticity lead to the highest levels of epidemic fatalities over a wide range of parameters.
+Hence, asymptomatic infection and transmission presents a double-edged sword.
 
+As context, foundational analyses of asymptomatic infections have primarily focused on the impact of asymptomatic infections on disease controllability (e.g., following the seminal work of Fraser et al., PNAS 2004) – neglecting the effects of asymptomatic spread on population-level outcomes, such as overall fatalities.
 Here, we use a series of epidemic models to assess the role of asymptomatic spread on epidemic fatalities in the presence of transmission-reducing effects among symptomatic individuals (e.g., via behavioural change or symptom-based interventions).
-We find that intermediate levels of asymptomaticity lead to the highest levels of epidemic fatalities over a wide range of parameters. 
-This counterintuitive result has not yet been demonstrated previously, nonetheless it emerges as a result of the countervailing forces of increased cases and decreasing severity with increasing asymptomaticity.
 We then extend our framework to assess the ways in which immunity profiles can lead to potentially paradoxical outcomes: with less virulent strains spreading precisely because asymptomatic individuals can transmit while unaware they are infected.
 Moreover, our findings show that emergence of milder variants at the individual scale can nonetheless cause similar or even higher levels of fatalities at population scales – mirroring many features of the Omicron variant.
-
-Following the advice of the two PNAS reviewers, we have revised our manuscript to better highlight the theoretical nature of our work and the associated uncertainties in parametric assumptions we make.
-We also perform sensitivity analyses to show that our qualitative conclusions are robust to parameter variation associated with asymptomatic transmissibility (Supplementary Figure S1--2).
-Finally, we provide directions for future areas of research that can assess both the impact of asymptomatic transmission and the extent to which control can improve population-level outcomes.
+Overall, we believe that the revised work is timely, novel, and fits the PNAS Nexus mission of welcoming rigorous research with broad, interdisciplinary appeal.
 
@@
 We hope that our revisions clarify the generalized nature of our findings and also point the way towards practical steps to link individual and population-level outcomes associated with asymptomatic transmission in realistic contexts.
-This process will require moving beyond epidemic models to include behaviour and policy responses.  
-Overall, we believe that the revised work fits the PNAS Nexus mission of welcoming rigorous research with broad, interdisciplinary appeal.
 

### 2022-12-30 response.tex
Update response.tex
Fixed ref to Fraser et al.

@@
 
-As context, foundational analyses of asymptomatic infections have primarily focused on the impact of asymptomatic infections on disease controllability (e.g., following the seminal work of Fraser et al., PNAS 2004) – neglecting the effects of asymptomatic spread on population-level outcomes, such as overall fatalities.
+As context, foundational analyses of asymptomatic infections have primarily focused on the impact of asymptomatic infections on disease controllability (e.g., Fraser et al., 2004) – neglecting the effects of asymptomatic spread on population-level outcomes, such as overall fatalities.
 Here, we use a series of epidemic models to assess the role of asymptomatic spread on epidemic fatalities in the presence of transmission-reducing effects among symptomatic individuals (e.g., via behavioural change or symptom-based interventions).

### 2023-01-05 main.tex
separate out a supp file

@@
 
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-
-\pagebreak
-
-\section*{Supplementary Tables}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
-      \hline
-      $\beta_s$ & Symptomatic transmission rate & $0.8/\mathrm{days}$\\
-      \hline
-      $\beta_a$ & Asymptomatic transmission rate & $0.75 \beta_s$ \\
-      \hline
-      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_s$ & Mean symptomatic infectious period & $5\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_a$ & Mean asymptomatic infectious period & $5\ \mathrm{days}$\\
-      \hline
-      $p$ & Proportion asymptomatic & 0--1\\
-      \hline
-      $f$ & Fatality rate for symptomatic case & 0.01\\
-      \hline
-      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
-    \end{tabular}
-    \caption{Parameter descriptions and values for the basic asymptomatic model.}
-    \label{tab:table1}
-  \end{center}
-\end{table}
-
-\pagebreak
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
-      \hline
-      $\beta_s$ & Symptomatic transmission rate & See Materials and Methods\\
-      \hline
-      $\beta_a$ & Asymptomatic transmission rate & See Materials and Methods \\
-      \hline
-      $\beta_p$ & Presymptomatic transmission rate & See Materials and Methods \\
-      \hline
-      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
-      \hline
-      $1/\sigma$ & Mean presymptomatic infectious period & $2\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_s$ & Mean symptomatic infectious period & $3\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_a$ & Mean asymptomatic infectious period & $3\ \mathrm{days}$\\
-      \hline
-      $p$ & Proportion asymptomatic & 0--1\\
-      \hline
-      $f$ & Fatality rate for symptomatic case & 0.01\\
-      \hline
-      $\delta_s$ & Reduction in symptomatic transmission rate & 0--1\\
-    \end{tabular}
-    \caption{Parameter descriptions and values for the generalized asymptomatic model.}
-    \label{tab:table2}
-  \end{center}
-\end{table}
-
-\pagebreak
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{c|l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
-      \textbf{Parameter} & \textbf{Description} & \textbf{Assumed values}\\
-      \hline
-      $\beta_s$ & Symptomatic transmission rate & $0.8/\mathrm{days}$\\
-      \hline
-      $\beta_a$ & Asymptomatic transmission rate & $0.75 \beta_s$ \\
-      \hline
-      $1/\nu$ & Mean latent period & $2\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_s$ & Mean symptomatic infectious period & $5\ \mathrm{days}$\\
-      \hline
-      $1/\gamma_a$ & Mean asymptomatic infectious period & $5\ \mathrm{days}$\\
-      \hline
-      $p$ & Proportion asymptomatic & 0--1\\
-      \hline
-      $f$ & Fatality rate for symptomatic case & 0.01\\
-      \hline
-      $\delta$ & Reduction in symptomatic transmission rate & 0--1\\
-      \hline
-      $\epsilon_i$ & Protection against infection & 0--0.8\\
-      \hline
-      $\epsilon_s$ & Protection against symptoms & 0--0.8\\
-      \hline
-      $\epsilon_d$ & Protection against deaths & 0--0.8\\
-    \end{tabular}
-    \caption{Parameter descriptions and values for the asymptomatic model with immunity.}
-    \label{tab:table3}
-  \end{center}
-\end{table}
-
-\pagebreak
-
-\section*{Supplementary Figures}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\textwidth]{figure_base_sens.ggp.pdf}
-\caption{
-\textbf{Simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction for a wide range of asymptomatic transmissibility.}
-Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
-We allow the ratios between the asymptomatic transmission rate $\beta_a$ and symptomatic transmission rate $\beta_s$ to vary between 0.25 and 1.
-See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
-}
-\end{center}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\textwidth]{figure_base_sens_fixR0.ggp.pdf}
-\caption{
-\textbf{Simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction for a wide range of asymptomatic transmissibility and a fixed $\mathcal{R}_0$ value at intermediate asymptomaticity.}
-Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
-We allow the ratios between the asymptomatic transmission rate $\beta_a$ and symptomatic transmission rate $\beta_s$ to vary between 0.25 and 1.
-We also fix the basic reproduction number $\mathcal{R}_0=4$ when there are intermediate levels of asymptomaticity $p=0.5$ and no reduction in symptomatic transmission rate $\delta=0$.
-See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
-}
-\end{center}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{diagram_sub.pdf}
-\caption{
-\textbf{Schematic diagram and simulations of a model with pre-symptomatic and asymptomatic transmission and symptom-responsive transmission reduction.}
-(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_p$ represents pre-symptomatic individuals; $I_a$ represents symptomatic individuals; $I_s$ represents symptomatic individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
-See Methods for model details.
-(B) Total deaths as a function of the proportion of non-symptomatic transmission $\phi$ across a wide range scenarios for $\delta_s$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta$ (between 0\% and 100\%).
-See Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values.
-}
-\end{center}
-\end{figure}
-
-\pagebreak
-
 \bibliography{main}

### 2023-01-05 supp.tex
separate out a supp file

--- 2022-12-30 main.tex
@@
 	\textbf\newline{
-		Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities
+		Supplementary Materials for Intermediate levels of asymptomatic transmission can lead to the highest epidemic fatalities
 	}
@@
 \end{flushleft}
-
-\section*{Abstract}
-
-Asymptomatic infections have hampered the ability to characterize and prevent the transmission of SARS-CoV-2 throughout the pandemic.
-Although asymptomatic infections reduce severity at the individual level, they can make population-level outcomes worse if asymptomatic individuals---unaware they are infected---transmit more than symptomatic individuals.
-Using an epidemic model, we show that intermediate levels of asymptomatic infection lead to the highest levels of epidemic fatalities when the decrease in symptomatic transmission, due either to individual behavior or mitigation efforts, is strong.
-We generalize this result to include presymptomatic transmission, showing that intermediate levels of non-symptomatic transmission lead to the highest levels of fatalities.
-Finally, we extend our framework to illustrate how the intersection of asymptomatic spread and immunity profiles determine epidemic trajectories, including population-level severity, of future variants.
-In particular, when immunity provides protection against symptoms, but not against infections or deaths, epidemic trajectories can have faster growth rates and higher peaks, leading to more total deaths.
-Conversely, even modest levels of protection against infection can mitigate the population-level effects of asymptomatic spread.
-
-\section*{Significance}
-
-During an epidemic, asymptomatically infected individuals may avoid severe outcomes but can still transmit to others, potentially leading to severe outcomes, including fatalities.
-This manuscript analyzes the population-level effects of asymptomatic spread in the presence of strong transmission reduction among symptomatic individuals due to behavioral change and interventions.
-Theory and simulations reveal that when reduction is strong the number of infections increases with asymptomatic proportion, while fatalities peak at intermediate levels.
-The same framework also shows how milder variants at the individual level can potentially lead to worse outcomes for the population---of relevance to ongoing efforts to explore the interplay between behaviour, immunity, and viral variants.
-
-\pagebreak
-
-\section*{Introduction}
-
-SARS-CoV-2 has had devastating effects at the population level.
-However, many individuals experienced mild cases, making it harder to estimate the magnitude of spread and fatality rate \cite{nogrady2020data}.
-The ratio of fatalities to documented \emph{cases} (the case-fatality rate, CFR) is typically between 1\%--4\%, varying across population because of testing patterns, treatment practice, case definitions, and other factors \cite{rajgor2020many,VERITY2020669,yang2020early}.
-But many infections are never documented;
-the ratio of fatalities to total \emph{infections} (the infection fatality rate, IFR) has been estimated to be closer to 0.5\%--1\% for pre-vaccinated populations whose demographics are similar to those of the United States \cite{levin2020assessing}.
-This means that more than 99\% of individuals infected with COVID-19 will survive. 
-Moreover, at least half of the infections are sufficiently mild that they could be classified as subclinical.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figure_evidence.ggp.pdf}
-\caption{
-\textbf{Asymptomatic transmissibility of SARS-CoV-2.}
-(A) Posterior estimates of the proportion of asymptomatic infections from the Diamond Princess cruise ship \cite{emery2020}.
-(B) Posterior estimates of the ratio $\theta_a$ of the transmission rates between asymptomatic and symptomatic individuals from the Diamond Princess Cruise Ship \cite{emery2020}.
-%% Bar charts represent the posterior distributions.
-Symptomatic individuals were assumed to transmit at rate $\beta(t)$ for an average of 2.9 days, followed by a pre-symptomatic stage with an average of 2.1 days. 
-Asymptomatic individuals were assumed to transmit at rate $\theta_a \beta(t)$ for an average of 5 days.
-Both estimates are publicly available with further details in \cite{emery2020}.
-(C) Viral load trajectory data from players, staff, and vendors of the National Basketball Association (NBA).
-Points represent each Ct measurement.
-Lines and shaded areas show default LOESS fits from \texttt{ggplot2} \cite{ggplot2}.
-Data are publicly available in \cite{Kissler2020}.
-(D) Inferred proportion of infections that are subclinical for each age group using surveillance data from six countries (China, Italy, Japan, Singapore, South Korea, and Canada) \cite{davies2020}.
-}
-\label{fig:evidence}
-\end{figure}
-
-It is now well established that asymptomatic individuals can transmit SARS-CoV-2 infections \citep{gao2021role,johansson2021sars,subramanian2021quantifying,koelle2022changing,lizewski2022navy}, but asymptomatic cases are increasingly shaped by prior immunity (whether through infection, vaccination, or both).
-In contrast, early in the pandemic, a COVID-19 outbreak on the Diamond Princess cruise ship played a critical role in understanding the role of asymptomatic infections in the spread of SARS-CoV-2 from and to immunologically naive individuals; the outbreak occurred among 3711 passengers and crew, of whom 634 individuals tested positive by 20 February 2020 \cite{mizumoto2020estimating}.
-It has been estimated that 75\% (95\% CI: 70\%-78\%) of all infections on the cruise ship were asymptomatic (\fref{evidence}A) with about half of total infections undetected \cite{emery2020}.
-The relative transmission rate of asymptomatic individuals aboard the Diamond Princess was not well constrained by the analysis, but low relative transmission rate (below 25\%) by asymptomatic individuals was ruled out because it required unrealistically high transmissibility for symptomatic individuals (\fref{evidence}B).
-
-Modeling studies have typically assumed that transmissibility is lower for asymptomatic than for symptomatic individuals; assumptions have ranged from 10\%--100\% \cite{ferretti2020quantifying,lavezzo2020}.
-Similarities in viral load trajectories of asymptomatic and symptomatic individuals provide indirect support for the transmissibility of asymptomatic individuals (\fref{evidence}C, \cite{Kissler2020});
-however, differences between inferred \emph{total} viral load from Ct values and \emph{infectious} viral load add uncertainty \cite{romero2021ct}.
-The likely role of symptoms in transmission further contributes to this uncertainty.
-Coughing and sneezing can help deliver virus-containing droplets.
-However, the prevalence of presymptomatic and asymptomatic transmission of SARS-CoV-2 suggests that speech droplets can also be an important mode of transmission \citep{stadnytskyi2021breathing}.
-
-We note also that asymptomaticity can change across outbreak settings \cite{alene2021magnitude}.
-For example, during the early pandemic, Davies \textit{et al.}'s analyses of surveillance data across six countries revealed that older individuals were less likely to have subclinical infections (\fref{evidence}D), providing indirect evidence for heterogeneity in asymptomaticity \cite{davies2020}.
-Differences in contact rates between age classes further contribute to the heterogeneity in asymptomatic transmissibility.
-For now, we primarily focus on a homogeneous population and return to the age effect in discussing our model-based findings.
-
-Despite quantitative uncertainties in asymptomatic transmissibility, individuals infected asymptomatically with SARS-CoV-2 can still transmit to others. 
-This means that the presence of asymptomatic infections may have countervailing effects at the population level. 
-On one hand, an asymptomatic infection means that the individual infected avoids hospitalization and death. 
-On the other hand, asymptomatic infections are less likely to be detected \cite{fraser2004factors,moghadas2020implications}, meaning that asymptomatic individuals are less likely to take precautions and relatively more likely to infect others;
-asymptomatic SARS-CoV-2 infections present additional challenges to managing overall disease burden due to the possibility of long COVID \cite{xie2022long}.
-
-In this manuscript, we explore the effects of asymptomatic infection and transmission on disease severity at the population level.
-In doing so, we assume that symptomatic individuals reduce their transmission, reflecting changes in behavior (e.g., self-isolation after symptom onset) and/or nonpharmaceutical intervention measures.
-Under this assumption, we show that a high proportion of asymptomatic infections could paradoxically make population-level outcomes worse than if SARS-CoV-2 was more dangerous at the individual level.
-We further extend our framework to understand the interaction between immunity against symptomatic infections on the dynamics of emerging variants and explore mechanisms by which milder variants at the individual level can nonetheless lead to similar or worse population-level outcomes.
-
-\section*{Results and Discussion}
-
-We propose an epidemic model in which infected individuals can be asymptomatic or symptomatic, with probabilities $p$ and $1-p$, respectively (\fref{base}A).  
-Asymptomatic individuals always recover, whereas a fraction $f$ of symptomatic individuals die.
-Asymptomatic and symptomatic individuals can also have different infection characteristics, including their transmission rates ($\beta_a$ and $\beta_s$) and removal rates ($\gamma_a$ and $\gamma_s$).
-Our key assumption is that symptomatic individuals take greater precautions than do asymptomatic individuals (e.g., reducing contacts or increasing mask-wearing) and therefore reduce their transmission rate by a fraction $\delta$;
-the parameter $\delta$ may also capture intervention measures that target symptomatic individuals, such as symptom-based isolation. 
-We note that intervention measures that target asymptomatic infections would reduce the effective value of $\delta$---for example, frequent testing and isolation may effectively increase the removal rate $\gamma_a$ of asymptomatic individuals.
-For our main simulations, we assume that asymptomatic individuals have a lower reproduction number---this is implemented via lower transmission rates for asymptomatic individuals ($\beta_a = 0.75 \beta_s$) and equal removal rates ($\gamma_a = \gamma_s$).
-We then evaluate the effects on population-level mortality of changing the asymptomatic proportion $p$ while holding the fatality rate \emph{for symptomatic cases}, $f$, constant (the population-level IFR $(1-p)f$ thus decreases as $p$ increases).
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{diagram_base.pdf}
-\caption{
-\textbf{Schematic diagram and simulations of a model with asymptomatic transmission and symptom-responsive transmission reduction.}
-(A) $S$ represents susceptible individuals; $E$ represents exposed individuals; $I_a$ represents asymptomatic individuals; $I_s$ represents symptomatic individuals; $R$ represents recovered individuals; and $D$ represents deceased individuals.
-See Methods for model details.
-(B) Total infections as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-(C) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for $\delta$.
-We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, and $f=0.01$, and an initial exposed proportion of $10^{-4}$.
-See Materials and Methods for model details and Supplementary Table S1 for parameter descriptions and values.
-}
-\label{fig:base}
-\end{figure}
-
-\fref{base}B--C shows simulated epidemic outcomes using parameters similar to those of the originating strain of SARS-CoV-2 (Table S1), without any mitigation other than that individuals who are symptomatic reduce their transmission rate by $\delta$.
-For this model, the basic reproduction number is given by:
-\begin{equation}
-\RR_0 = (1-p) (1-\delta) \RR_s + p \RR_a,
-\end{equation}
-where $\RR_s = \beta_s/\gamma_s$ and $\RR_a = \beta_a/\gamma_a$ represent the reproduction numbers of asymptomatic and symptomatic individuals (i.e., the average number of secondary infections caused by asymptomatic and symptomatic individuals);
-therefore, in the absence of the behavioral effect ($\delta=0$), the final size decreases with the asymptomatic proportion $p$ because more symptomatic infections leads to a higher basic reproduction number.
-
-This relationship changes as $\delta$ increases.
-In particular, when $\delta > 1-\RR_a/\RR_s$ (in this case, $\delta > 0.25)$, the basic reproduction number (and thus epidemic size) increases with $p$ because the effective symptomatic reproductive number (including behavioral response) is less than that the asymptomatic reproductive number.
-When $\delta > 1-1/\RR_s$( and $\RR_a>1$), we can find a critical asymptomatic proportion, $p_c$:
-\begin{equation}
-    p_c = \frac{1 - (1-\delta) \RR_s}{\RR_a - (1-\delta) \RR_s}
-\end{equation}
-such that an outbreak will occur exactly when $p>p_c$ (see threshold effects for large values of $\delta$ in \fref{base}B).
-
-When behavioral protection is high, the effect of asymptomatic proportion on fatalities shows countervailing effects of individual-level protection and population-level risk (\fref{base}C).
-For high values of $\delta$, the peak fatality occurs at intermediate levels of asymptomatic spread:
-although fewer individuals die per infection for higher values of $p$, the increase in total infections still leads to an increase in total fatalities.
-In contrast, when $\delta$ is small enough that $(1-\delta) \RR_s \geq \RR_a$ (in this case, $\delta < 0.25$), both the number of infections and the IFR ($(1-p)f$) decrease with increasing $p$.
-We note that these results are robust to uncertainties in asymptomatic transmission rate.
-In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_a$ and $\RR_s$ between 0.25 and 1.
-In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$
---- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large then $\RR_0 < 1$ and the epidemic does not take off.
-In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
-In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
-
-We can ask whether the high values of $\delta$ required for the nonlinear effects of asymptomaticity on deaths are realistic.
-For this particular model, it does not make biological sense for $\delta$ to be greater than the amount of post-symptomatic transmission, because pre-symptomatic transmission is implicitly included in the $I_s$ compartment.
-While several studies have estimated the proportion of pre-symptomatic transmission to be around 30\%--60\% for the SARS-CoV-2 wildtype strain, many of these were likely affected by intervention and behavioral effects, as they were conducted after SARS-CoV-2 awareness became widespread \cite{he2020temporal}.
-Instead, \cite{sender2021unmitigated} recently estimated that the proportion of pre-symptomatic transmission could have been as low as 20\% (95\%CI: 6\%--32\%) during the first few weeks of the pandemic when the pandemic-awareness and intervention measures were minimal.
-There are two implications of this updated estimate---first, a low proportion of pre-symptomatic transmission makes high $\delta$ values at least somewhat more likely during the initial pandemic phase; and second, substantial levels of behavioral effects ($\delta > 0$) may have been present early in the pandemic (to reduce the proportion of symptomatic transmission from 80\% to as low as 40\%).
-
-To further address the uncertainty in the values of $\delta$, we extend our model to include both pre-symptomatic and asymptomatic transmission.
-To do so, we reparameterize the model based on the relative importance of \textit{non-symptomatic} transmission, rather than on the proportion of asymptomatic cases.
-We then use $\delta_s$ to capture the decrease in transmission only after symptom onset---this means that $\delta_s$ is now independent of the amount of presymptomatic transmission.
-We then fix the reproduction number of symptomatic individuals and calculate fatalities at the population level as a function of the proportion of total non-symptomatic transmission and the proportion of non-symptomatic transmission that is caused by pre-symptomatic transmission (see Materials and Methods for model details and Supplementary Table S2 for parameter descriptions and values).
-
-Using the generalized non-symptomatic transmission model, we find a wide variety of scenarios for which peak fatalities occur at intermediate levels of non-symptomatic transmission in the presence of moderate to strong behavioral effects, $\delta_s > 0.6$ (Supplementary Figure S3; Table S2).
-For example, when 40\% of non-symptomatic transmission is caused by pre-symptomatic transmission, a 60\% reduction in transmission after symptom onset is sufficient to drive the nonlinear effect of non-symptomatic transmission on epidemic fatality (peaking at around 10\% non-symptomatic transmission).
-One exception is the extreme case in which all non-symptomatic transmission is caused by pre-symptomatic transmission (i.e., there is no asymptomatic transmission);
-in this case, total infections and fatalities are maximized when all transmission is caused by pre-symptomatic transmission. 
-While 60\% reduction in transmission after symptom onset ($\delta_s = 0.6$) is still high, it is plausible given behavioural and policy responses.
-For example, \cite{kucharski2020effectiveness} estimated that up to $\sim$64\% reduction in transmission would be possible under self-isolation and self-quarantine as well as manual contact tracing (and as low as $\sim$47\% reduction using digital contact tracing without manual tracing).
-Given that the majority of isolation and tracing measures likely target symptomatic transmission, the amount of reduction in symptomatic transmission would be similar to these values, providing indirect support for the feasibility of high $\delta_s$ values.
-Hereafter, we focus on asymptomatic infections for simplicity, but our conclusions have implications for the more general case of non-symptomatic transmission.
-We return to the discussion of values of $\delta$ later in the Discussion section.
-
-We now apply our framework to understand the impact of immunity on total fatalities at the population scale by dividing the population into two groups: immunologically naive and protected.
-For simplicity, we do not distinguish whether the immunity is derived from natural infections or vaccines.
-The dynamics of immunologically naive individuals are equivalent to our original model (\fref{base}).
-The dynamics of protected individuals include three additional parameters, which characterize the amount of protection against infection $\epsilon_i$, symptoms (given infection) $\epsilon_s$, and deaths (given symptoms) $\epsilon_d$ (\fref{immune}).
-For simplicity, we assume that the population is split in half (50\% naive and 50\% protected) and mixes homogeneously.
-We also do not consider the separate effect of immunity on transmission (beyond the effect on infection).
-In other words, we assume that asymptomatic infections in protected and unprotected people have the same reproduction numbers (and likewise for the symptomatic infections).
-In practice, both asymptomatic and symptomatic infections in protected people are less likely to transmit than their unprotected counterparts \cite{lipsitch2020understanding}: asymptomatic infections in protected people may indicate limited viral replication or even immune boosting, in which case an exposed individual may successfully fight off the pathogen early in infection before it can be transmitted; and symptomatic infections in protected people may reflect a strong immune response (rather than high viral load), in which case symptomaticity can be a poor proxy for transmission.
-We assume a relatively strong behavioral effect $\delta=0.8$ for illustration (Table S3).
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{diagram_immune.pdf}
-\caption{
-\textbf{Schematic diagram and simulations of a model with symptom-responsive transmission reduction and immunity.}
-(A) The subscript $p$ represents protected individuals. 
-Immunity may provide protection against infection, symptoms, or deaths.
-The dynamics of immunologically naive individuals are described in \fref{base}.
-(B--D) Total deaths as a function of the proportion of asymptomatic infections $p$ across a wide range scenarios for protection against infection $\epsilon_i$ (B), symptoms $\epsilon_s$ (C), and deaths $\epsilon_d$ (D).
-We simulate the model for 365 days, assuming $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $\nu=0.5/\mathrm{day}$, $\gamma_s=\gamma_a=0.2/\mathrm{day}$, $f=0.01$, and $\delta=0.8$.
-The initial infected proportion is $10^{-4}$.
-See Materials and Methods for model details and Supplementary Table S3 for parameter descriptions and values.
-}
-\label{fig:immune}
-\end{figure}
-
-We consider each protection effect---$\epsilon_i$, $\epsilon_s$, and $\epsilon_d$---separately and consider joint effects later on.
-The impact of protection against infection $\epsilon_i$ is analogous to changing $\RR_0$ in the original model: as immunity provides stronger protection against infection (higher $\epsilon_i$), the number of deaths decreases and a higher asymptomatic fraction $p$ is required for the infection to spread (\fref{immune}B).
-We note that protection against infection scales the fatality curve nonlinearly, reflecting the nonlinear relationship between $\RR_0$ and the final size of the outbreak.
-The impact of protection against symptoms $\epsilon_s$ is equivalent to changing the asymptomatic fraction $p$ for the protected population because protected individuals are less likely to develop symptoms:
-the peaks of the fatality curves move to lower values of $p$ as we increase the degree of protection $\epsilon_s$ (\fref{immune}C).
-Therefore, for low values of $p$, protection against symptoms can increase the total number of fatalities at the population level by increasing the proportion (and number) of asymptomatic individuals, who can readily transmit infections to other individuals.
-This also means that the critical asymptomatic proportion decreases, allowing more dangerous infections (with lower $p$) to invade, which would not have been able to spread in an otherwise immunologically naive population.
-We note that the equivalence between protection against symptoms $\epsilon_s$ and fraction asymptomatic $p$ relies on our assumption that immunity does not provide protection against transmission.
-Protection against deaths $\epsilon_d$ directly modulates the fatality rate for symptomatic cases and therefore linearly scales the fatality curves (\fref{immune}D).
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figure_variant.ggp.pdf}
-\caption{
-\textbf{Dynamics of invading variants under symptom-responsive transmission reduction and immunity.}
-(A, D) Asymptomaticity--fatality curves for the first (solid lines) and second waves (dashed lines).
-Points represent specific scenarios we assume for the first and second waves.
-Fatality curves for the first wave are calculated by simulating an epidemic for 1 year using parameters from \fref{base} with $\delta=0.8$.
-Fatality curves for the second wave are calculated by first simulating the first wave assuming $p=0.2$ for 1 year to calculate the proportion immune and then simulating the extended model presented in \fref{immune} for \emph{either} a milder ($p=0.8$) or an equally severe ($p=0.2$) variant.
-(B, E) Dynamics of infection prevalence for the wildtype variant (black, solid line) and milder (purple) and similar (orange) invading variants (dashed lines).
-(C, F) Dynamics of daily deaths for the wildtype variant (black, solid line) and two possible invading variants (colored, dashed line).
-}
-\label{fig:variant}
-\end{figure}
-
-Finally, we use our framework to understand the impact of behavioral effects on invading variants (\fref{variant}).
-We first simulate the dynamics of a wildtype variant for 1 year using our base model with parameters as in \fref{base}.
-We then simulate a new variant invading a partially immune population using our extended model (\fref{immune}A), where the immunity is solely derived from natural infections caused by the wildtype variant in the first year.
-We consider two types of variants (which are simulated separately): one with the same severity $p$ as before (variant 1, orange) and a milder one with higher $p$ (variant 2, purple).
-
-First, we consider a scenario in which immunity only provides protection against symptoms, $\epsilon_s = 0.4$ (\fref{variant}A--C).
-In this case, protection against symptoms allows new variants to spread faster by increasing the amount of asymptomatic infections, resulting in larger outbreaks (\fref{variant}B).
-Although the milder (purple) variant exhibits a faster epidemic growth rate and reaches a higher peak (\fref{variant}B), it reaches similar peak fatality as the more severe (orange) variant (\fref{variant}C).
-The asymptomaticity--fatality curve provides additional insight (\fref{variant}A): even though a milder, invading variant (purple square) gives higher peak fatality than the original, wildtype variant (black circle), it leads to lower fatalities overall because deaths are concentrated over a shorter period of time in the epidemic; the relatively severe second-wave variant causes more deaths than the wildtype first wave, despite causing many fewer cases.
-In general, when $\delta$ is large, invading variants with similar asymptomaticity $p$ will spread more effectively and result in worse population-level outcomes if immunity (either from vaccination or natural infection) provides protection against symptoms but not against infection or transmission.
-
-Next, we consider a more realistic scenario in which immunity provides protection against both symptoms, $\epsilon_s = 0.4$, and infection, $\epsilon_i = 0.4$ (\fref{variant}D--F).
-In this case, cross-protection against infection has a large effect on the more severe (orange) variant, causing its peak infection prevalence (\fref{variant}E) and fatality (\fref{variant}F) to be lower than that of the original, wildtype variant.
-Across a wide range of asymptomatic proportion $p$, we find that this immunity profile is sufficient to prevent worse outcomes at the population level;
-we note that the second wave of deaths is still high (and having higher peaks in some cases) even if the overall deaths are lower.
-
-The outcomes in our simulations of invading variants resemble the dynamics of the SARS-CoV-2 Omicron variant.
-Despite moderate levels of vaccine effectiveness against symptomatic and reduced levels of severe cases caused by the Omicron variant, especially after booster shots \cite{andres2022omicron}, both vaccine- and infection-derived immunity provided limited protection against infections \cite{pearson2021omicron}.
-This immune evasion helped the Omicron variant to cause more infections in South Africa than previous variants \cite{sun2022omicron}.
-Moreover, even though the Omicron variant is probably milder than the Delta variant \cite{MENNI20221618,ulloa2022estimates}, the number of hospitalizations and deaths caused by the Omicron variant was higher than those caused by the Delta variant in many locations \cite{Iacobuccio254,faust2022omicron,sigal2022estimating}.
-
-There are several limitations to our analysis.
-First, while we are able to generalize the model to include both pre-symptomatic and asymptomatic transmission, behavioral and intervention effects must be relatively large in order for the fatality to peak at intermediate levels of asymptomaticity (typically requiring a reduction in transmission rate of 60\% or more for most of our chosen parameter sets).
-Second, the model framework is able to incorporate the impacts of immunity of infection, symptoms, and severity,
-but we neglected additional specific effects of immunity on transmission, which can also have important effects on disease dynamics \cite{saad2020immune,gallagher2021indirect}.
-In particular, if immunity provides stronger protection against transmission among immune individuals, population-level outcomes will be better than what our model predicts.
-Estimating protection against different endpoints (e.g., infection, symptom, death, and transmission) can help narrow this uncertainty.
-Finally, we assumed that asymptomatic and symptomatic individuals are infected for the same amount of time.
-Analysis of viral load trajectories suggests that asymptomatic individuals may clear infections faster \cite{Kissler2020};
-however, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
-The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \cite{park2020time,harris2022time};
-nonetheless, we expect that predictions on the final size of the epidemic and total fatalities will be robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
-
-Even though we assumed a homogeneous population here, our analysis also has important implications for age-dependent heterogeneity in asymptomaticity (as shown in \fref{evidence}D).
-For example, vaccinations and intervention measures primarily targeting older individuals can prevent severe infections and improve individual-level outcomes.
-However, asymptomatic individuals, especially younger individuals with high contact rates, can still transmit to other, older individuals, potentially making population-level outcomes worse than they would be if intervention measures were distributed differently.
-We note that other factors, such as the efficacy of a vaccine and types of immunity provided by the vaccine, also play critical roles in making these decisions---in many cases, protecting the most vulnerable will be the optimal decision to minimize deaths \cite{moore2021modelling}.
-
-The main conclusion of our analysis relies on having high levels of transmission reduction among symptomatic individuals ($\delta$). 
-For the simple model, the value of $\delta$ is limited by the amount of presymptomatic transmission, which makes high $\delta$ values unrealistic for COVID-19.
-Therefore, we extended our model to include both presymptomatic and asymptomatic transmission to show that our results hold for a more general case: for high levels of transmission reduction after symptom onset ($\delta_s$), epidemic fatalities peak at intermediate values of non-symptomatic transmission.
-Based on early estimates for the contact tracing effectiveness \citep{kucharski2020effectiveness} high levels are $\delta_s$ are plausible, but uncertainty remains. 
-Furthermore, values of $\delta$ (and likewise, $\delta_s$) likely change over the course of an epidemic.
-During the exponential growth phase, $\delta$ is likely low because there is limited awareness for the outbreak and intervention measures in place.
-As the number of cases, hospitalizations, and fatalities increase, $\delta$ will also increase, reflecting changes in awareness-driven behavior and intensity of non-pharmaceutical interventions \citep{weitz2020awareness}.
-Further analysis is needed to constrain the uncertainty in $\delta$ and to apply the model framework in distinct disease, policy, and socioeconomic contexts.
-
-Via theory and simulation analysis of a series of simplified models, we have shown that asymptomatic infections (or, more generally, non-symptomatic transmission) can under some conditions lead to a better outcome for many individuals while facilitating onward transmission that leads to a worse outcome for the population as a whole.
-Extending our framework further shows that immunity profile (i.e., reduction of infection, symptoms, and/or severity due to immunity) plays a critical role in determining the dynamics of future variants;
-these results extend previous work on post-pandemic trajectories that focus primarily on cross-immunity \cite{kissler2020projecting,lavine2021immunological}.
-For example, while protection against symptoms protects health at the individual level, it can lead to more infections, and potentially more deaths, at the population level.
-A similar concern was raised in prioritizing vaccine choices that could reduce severe outcomes vs. others that could reduce transmission \cite{koelle2022changing}.
-Our conclusions echo earlier findings by \cite{gandon2001imperfect}, who showed that vaccine-derived immunity against diseases can promote the evolution of more virulent strains in unvaccinated individuals.
-By allowing explicit parasite evolution, they also showed that total malaria mortality peaks at intermediate levels of vaccination; our analysis reveals that the differences in symptomatic and asymptomatic transmission behavior can give rise to a similar effect even in the absence of evolution.
-
-Our work uses theoretical model to illustrate the potential for asymptomatic infection and transmission to make population-level outcomes worse.
-Applying these ideas to specific outbreak scenarios will require narrowing down uncertainties in key parameters, such as the degree of symptom-responsive transmission reduction and immunity profiles.
-For example, we simulated unmitigated outbreaks with fixed parameters but epidemic dynamics, especially those of SARS-CoV-2, are more complex, reflecting changes in intervention efforts and the emergence of new variants.
-Calibrating the model to outbreak data is therefore critical to understanding the role of asymptomatic transmission across different epidemic phases.
-We also showed that asymptomatic infections can have important implications for evolutionary dynamics, but their contributions in driving evolutionary dynamics of SARS-CoV-2 is yet unclear.
-Our study provides a starting point for exploring these questions in more detail.
-
-As is increasingly evident, SARS-CoV-2 has proven hard to control in large part because transmission is often decoupled from symptoms. 
-% Mitigation efforts have often prioritized responding to symptoms---including symptom-based testing, fever checks, mask-wearing for infectious individuals---a different approach that strives to reduce the chance of asymptomatic transmission while increasing treatment of symptomatically infected individuals could both reduce infection risk at the source and the overall rates of severe outcomes for the population as a whole.
-Our model reinforces the need for dual approaches---prioritizing the reduction of asymptomatic spread (e.g., via risk awareness campaigns \cite{Bavel2020,chande2020real,sinclair2021pairing}, asymptomatic testing programs \cite{mina2021covid,gibson2022surv,Ranoa2022}, mask-wearing indoors and in crowded environments \cite{jones2020two,prather2020reducing,howard2021ev}, and through improvements in ventilation \cite{allen2021indoor,wang2021airborne}) while also improving the treatment of symptomatic cases, particularly amongst older individuals at highest risk for severe outcomes.
-Given the link between age and asymptomatic infections \cite{davies2020}, interventions may consider different approaches in strongly age-structured populations (e.g., schools or long-term care facilities).
-Mass vaccination is also expected to be important especially if future vaccines induce more transmission blocking.
-As more variants continue to emerge, monitoring the impacts of preexisting immunity (whether through vaccination and/or infections  \cite{lopman2021framework}) on preventing infections, and not just disease, will be critical to controlling the course of this and future pandemics.
-
-\section*{Materials and Methods}
-
-\subsection*{Epidemic models with asymptomatic infection and transmission in the absence of immunity}
-
-We consider a compartmental model with asymptomatic and symptomatic infections in a homogeneously mixing population.
-The basic model dynamics are as follows:
-\begin{align}
-\dot{S} &= -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta) \beta_s S I_s - \nu E\\
-\dot{I}_a &= p \nu E - \gamma_a I_a\\
-\dot{I}_s &= (1-p) \nu E -\gamma_s I_s\\
-\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
-\dot{D} &= f \gamma_s I_s
-\end{align}
-where the transmission rate $\beta$ and removal rate $\gamma$ can be potentially differ between asymptomatic and symptomatic individuals.  
-Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatic individuals.
-Throughout the paper, we use parameters that are broadly consistent with the dynamics of the originating strain of SARS-CoV-2: $\beta_s = 0.8/\mathrm{day}$, $\beta_a = 0.75 \beta_s$, $1/\nu=2\ \mathrm{days}$, $1/\gamma_s=1/\gamma_a=5\ \mathrm{days}$, and $f=0.01$ \cite{park2020reconciling}.
-Under this parameterization, we have symptomatic and asymptomatic reproduction numbers of $\RR_s = 4$ and $\RR_a = 3$.
-
-We then extend this model to include both pre-symptomatic and asymptomatic transmission:
-\begin{align}
-\dot{S} &= -\beta_p I_p -\beta_a S I_a -(1-\delta) \beta_s S I_s \\
-\dot{E} &= \beta_a S I_a + (1-\delta_s) \beta_s S I_s - \nu E\\
-\dot{I}_p &= \nu E - \sigma I_p\\
-\dot{I}_a &= p \sigma I_p - \gamma_a I_a\\
-\dot{I}_s &= (1-p) \sigma I_p -\gamma_s I_s\\
-\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
-\dot{D} &= f \gamma_s I_s
-\end{align}
-For this generalized model, the pre-symptomatic $\RR_p$, symptomatic $\RR_s$, and asymptomatic $\RR_a$ reproduction numbers are given by $\RR_p = \beta_p/\sigma$, $\RR_s=\beta_s/\gamma_s$, and $\RR_a=\beta_a/\gamma_a$ in the absence of the behavioral effect;
-these reproduction numbers represent the average number of secondary cases caused by an infected individual in each compartment.
-Then, the reproduction number of individuals who will eventually develop symptoms is equal to: $\RR_p + \RR_s$;
-similarly, the reproduction number of individuals who remain asymptomatic is equal to: $\RR_p + \RR_a$.
-Since the proportion $p$ of all infections is asymptomatic, the basic reproduction number is given by the weighted average of these two reproduction numbers: 
-\begin{equation}
-\RR_0 = p(\RR_p + \RR_a) + (1-p) (\RR_p + \RR_s) = \RR_p + p \RR_a + (1-p) \RR_s.
-\end{equation}
-Then, the proportion of non-symptomatic transmission $\phi$ is given by:
-\begin{equation}
-\phi = \frac{\RR_p + p \RR_a}{\RR_0}.
-\end{equation}
-
-For simulations of the combined model, we start by fixing the reproduction number of individuals who will eventually develop symptoms: $\RR_{\textrm{symp}} = \RR_p + \RR_s = 4$. 
-Consistent with previous assumptions, we also assume that asymptomatic reproduction number is lower than that of the symptomatic reproduction number: $\RR_a = \rho \RR_s$ where $\rho = 0.75$.
-Then, for a given value of the proportion of non-symptomatic transmission $\phi$ and proportion of non-symptomatic transmission caused by the pre-symptomatic transmission, $\eta = \RR_p/(\RR_p + p \RR_a)$, we can solve for the transmission rate for each compartment $\beta$ and the proportion asymptomatic $p$.
-More specifically:
-\begin{align}
-\RR_p &= \frac{\RR_{\textrm{symp}}}{1 + y}\\
-\RR_s &= \RR_{\textrm{symp}} - \RR_p\\
-\RR_a &=  \rho \RR_s\\
-p &= \left(\frac{1}{\eta} - 1 \right) \frac{\RR_p}{\RR_a},
-\end{align}
-where $y = (1/\phi - 1)/\eta + (1/\eta - 1)/\rho$.
-In order to keep the mean infectious period fixed, we assume $1/\sigma=2\ \mathrm{days}$ and $1/\gamma_s=1/\gamma_a=3\ \mathrm{days}$. All other parameters are same as before.
-
-\subsection*{Epidemic models with asymptomatic infection and transmission in the presence of immunity}
-
-We model the spread of infection in a partially immune population as follows:
-\begin{align}
-\dot{S} &= -\lambda (t) S \\
-\dot{E} &= \lambda (t) S - \nu E\\
-\dot{I}_a &= p \nu E - \gamma_a I_a\\
-\dot{I}_s &= (1-p) \nu E -\gamma_s I_s\\
-\dot{R} &= \gamma_a I_a + (1-f) \gamma_s I_s \\
-\dot{D} &= f \gamma_s I_s\\
-\dot{S}_p &= - (1-\epsilon_i) \lambda (t) S_p \\
-\dot{E}_p &= (1-\epsilon_i) \lambda (t) S_p - \nu E_p\\
-\dot{I}_{p, a} &= (1 - (1-\epsilon_s) (1-p)) \nu E_p - \gamma_a I_{p,a}\\
-\dot{I}_{p, s} &= (1-\epsilon_s) (1-p) \nu E_p -\gamma_s I_{p,s}\\
-\dot{R}_p &= \gamma_a I_{p,a} + (1-(1-\epsilon_d) f) \gamma_s I_{p,s} \\
-\dot{D}_p &= (1-\epsilon_d) f \gamma_s I_{p,s}
-\end{align}
-where $0\leq \epsilon_i, \epsilon_s, \epsilon_d \leq 1$ represents the degree of protection against infection, symptoms and death, respectively. 
-The force of infection $\lambda(t)$ is given by:
-\begin{equation}
-\lambda(t) = \beta_a (I_a + I_{p,a}) + (1-\delta) \beta_s (I_s + I_{p,s}).
-\end{equation}
-Here, subscripts $p$ denote individuals who are immune and therefore are protected, and as before the subscripts $a$ and $s$ denote asymptomatic and symptomatic infections.
-
-\section*{Acknowledgements}
-
-JSW acknowledges support from the Chaire Blaise Pascal Program of the Île-de-France region.
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
-\section*{Data availability}
-
-All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/asymptomaticvariant}).
-
-\pagebreak
 

### 2023-03-21 main.tex
updating

@@
 however, asymptomatic individuals may still transmit for a longer period of time if symptomatic individuals self-isolate quickly after symptom onset.
-The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \cite{park2020time,harris2022time};
+The individual-level differences in the asymptomatic and symptomatic transmission time scale can have important implications for the inferences and predictions of pathogen dynamics \cite{park2020time,harris2023time};
 nonetheless, we expect that predictions on the final size of the epidemic and total fatalities will be robust to small differences in the transmission time scale between asymptomatic and symptomatic individuals.
@@
 where the transmission rate $\beta$ and removal rate $\gamma$ can be potentially differ between asymptomatic and symptomatic individuals.  
+Similar models have been previously used to study the dynamics of SARS-CoV-2 \citep{fox2020impact,lavezzo2020,harris2023time}.
 Here, $\delta$ denotes the reduction in transmissibility due to responsive measures taken by symptomatic individuals.
@@
 
-We then extend this model to include both pre-symptomatic and asymptomatic transmission:
+We then extend this model to include both pre-symptomatic and asymptomatic transmission.
+This is done by adding a presymptomatic compartment before entering symptomatic or asymptomatic compartments \citep{lavezzo2020}.
 \begin{align}
@@
 
-We model the spread of infection in a partially immune population as follows:
+We model the spread of infection in a partially immune population by assuming a leaky protection.
+The leaky assumption has been widely used throughout the SARS-CoV-2 pandemic \citep{saad2020immune}.
+Another option is to rely on the all-or-nothing assumption \citep{smith1984assessment,buckner2021dynamic,bubar2021model};
+for simplicity, we do not explore this option.
+Then, the model equations are given by:
 \begin{align}

### 2023-03-21 final_files/main.tex
done?

--- 2023-03-21 main.tex
@@
 We note that these results are robust to uncertainties in asymptomatic transmission rate.
-In Supplementary Figure S2, we perform the same analysis while varying the ratio between $\RR_a$ and $\RR_s$ between 0.25 and 1.
+In Supplementary Figure S1, we perform the same analysis while varying the ratio between $\RR_a$ and $\RR_s$ between 0.25 and 1.
 In this case, qualitative predictions are robust when $\RR_a \geq 0.5 \RR_s$
 --- when $\RR_a = 0.25 \RR_s$ and $\delta$ is large then $\RR_0 < 1$ and the epidemic does not take off.
-In Supplementary Figure S3, we perform the same sensitivity analysis as Supplementary Figure S2 while fixing $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
+In Supplementary Figure S2, we perform the same sensitivity analysis as Supplementary Figure S1 while fixing $\RR_0 = 4$ when $p=0.5$ and $\delta=0$ to prevent the epidemic from dying out.
 In this case, we find that intermediate values of asymptomaticity lead to the highest epidemic fatalities at high $\delta$ values across all ranges of $\RR_a/\RR_s$ we consider.
@@
 
-\bibliography{main}
+\begin{thebibliography}{10}
+
+\bibitem{nogrady2020data}
+Bianca Nogrady.
+\newblock What the data say about asymptomatic {COVID} infections.
+\newblock {\em Nature}, 587(7835):534--536, 2020.
+
+\bibitem{rajgor2020many}
+Dimple~D Rajgor, Meng~Har Lee, Sophia Archuleta, Natasha Bagdasarian, and
+  Swee~Chye Quek.
+\newblock {The many estimates of the COVID-19 case fatality rate}.
+\newblock {\em The Lancet Infectious Diseases}, 20(7):776--777, 2020.
+
+\bibitem{VERITY2020669}
+Robert Verity, Lucy~C Okell, Ilaria Dorigatti, Peter Winskill, Charles
+  Whittaker, Natsuko Imai, Gina Cuomo-Dannenburg, Hayley Thompson, Patrick G~T
+  Walker, Han Fu, Amy Dighe, Jamie~T Griffin, Marc Baguelin, Sangeeta Bhatia,
+  Adhiratha Boonyasiri, Anne Cori, Zulma Cucunubá, Rich FitzJohn, Katy
+  Gaythorpe, Will Green, Arran Hamlet, Wes Hinsley, Daniel Laydon, Gemma
+  Nedjati-Gilani, Steven Riley, Sabine {van Elsland}, Erik Volz, Haowei Wang,
+  Yuanrong Wang, Xiaoyue Xi, Christl~A Donnelly, Azra~C Ghani, and Neil~M
+  Ferguson.
+\newblock Estimates of the severity of coronavirus disease 2019: a model-based
+  analysis.
+\newblock {\em The Lancet Infectious Diseases}, 20(6):669--677, 2020.
+
+\bibitem{yang2020early}
+Shu Yang, Peihua Cao, Peipei Du, Ziting Wu, Zian Zhuang, Lin Yang, Xuan Yu,
+  Qi~Zhou, Xixi Feng, Xiaohui Wang, Weiguo Li, Enmei Liu, Ju~Chen, Yaolong
+  Chen, and {Daihai He on behalf of COVID-19 evidence and recommendations
+  working group}.
+\newblock {Early estimation of the case fatality rate of COVID-19 in mainland
+  China: a data-driven analysis}.
+\newblock {\em Annals of translational medicine}, 8(4), 2020.
+
+\bibitem{levin2020assessing}
+Andrew~T Levin, William~P Hanage, Nana Owusu-Boaitey, Kensington~B Cochran,
+  Seamus~P Walsh, and Gideon Meyerowitz-Katz.
+\newblock {Assessing the age specificity of infection fatality rates for
+  COVID-19: systematic review, meta-analysis, and public policy implications}.
+\newblock {\em European journal of epidemiology}, 35(12):1123--1138, 2020.
+
+\bibitem{emery2020}
+Jon~C Emery, Timothy~W Russell, Yang Liu, Joel Hellewell, Carl~AB Pearson,
+  CMMID COVID-19~Working Group, Gwenan~M Knight, Rosalind~M Eggo, Adam~J
+  Kucharski, Sebastian Funk, Stefan Flasche, and Rein~MGJ Houben.
+\newblock {The contribution of asymptomatic SARS-CoV-2 infections to
+  transmission on the Diamond Princess cruise ship}.
+\newblock {\em eLife}, 9:e58699, aug 2020.
+
+\bibitem{ggplot2}
+Hadley Wickham.
+\newblock {\em {ggplot2: Elegant Graphics for Data Analysis}}.
+\newblock Springer-Verlag New York, 2016.
+
+\bibitem{Kissler2020}
+Stephen~M. Kissler, Joseph~R. Fauver, Christina Mack, Scott~W. Olesen, Caroline
+  Tai, Kristin~Y. Shiue, Chaney~C. Kalinich, Sarah Jednak, Isabel~M. Ott,
+  Chantal~B.F. Vogels, Jay Wohlgemuth, James Weisberger, John DiFiori,
+  Deverick~J. Anderson, Jimmie Mancell, David~D. Ho, Nathan~D. Grubaugh, and
+  Yonatan~H. Grad.
+\newblock {Viral dynamics of acute SARS-CoV-2 infection}.
+\newblock {\em medRxiv}, 2021.
+
+\bibitem{davies2020}
+Nicholas~G. Davies, Petra Klepac, Yang Liu, Kiesha Prem, Mark Jit, Rosalind~M.
+  Eggo, and {CMMID COVID-19 working group}.
+\newblock {Age-dependent effects in the transmission and control of COVID-19
+  epidemics}.
+\newblock {\em Nature Medicine}, 26(8):1205--1211, 2020.
+
+\bibitem{gao2021role}
+Wenjing Gao, Jun Lv, Yuanjie Pang, and Li-Ming Li.
+\newblock {Role of asymptomatic and pre-symptomatic infections in COVID-19
+  pandemic}.
+\newblock {\em bmj}, 375, 2021.
+
+\bibitem{johansson2021sars}
+Michael~A Johansson, Talia~M Quandelacy, Sarah Kada, Pragati~Venkata Prasad,
+  Molly Steele, John~T Brooks, Rachel~B Slayton, Matthew Biggerstaff, and Jay~C
+  Butler.
+\newblock {SARS-CoV-2 transmission from people without COVID-19 symptoms}.
+\newblock {\em JAMA network open}, 4(1):e2035057--e2035057, 2021.
+
+\bibitem{subramanian2021quantifying}
+Rahul Subramanian, Qixin He, and Mercedes Pascual.
+\newblock {Quantifying asymptomatic infection and transmission of COVID-19 in
+  New York City using observed cases, serology, and testing capacity}.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  118(9):e2019716118, 2021.
+
+\bibitem{koelle2022changing}
+Katia Koelle, Michael~A Martin, Rustom Antia, Ben Lopman, and Natalie~E Dean.
+\newblock {The changing epidemiology of SARS-CoV-2}.
+\newblock {\em Science}, 375(6585):1116--1121, 2022.
+
+\bibitem{lizewski2022navy}
+Rhonda~A. Lizewski, Rachel S.~G. Sealfon, Sang~Woo Park, Gregory~R. Smith,
+  Chad~K. Porter, Ana~S. Gonzalez-Reiche, Yongchao Ge, Clare~M. Miller, Carl~W.
+  Goforth, Hanna Pincas, Michael~S. Termini, Irene Ramos, Venugopalan~D. Nair,
+  Stephen~E. Lizewski, Hala Alshammary, Regina~Z. Cer, Hua~Wei Chen,
+  Mary-Catherine George, Catherine~E. Arnold, Lindsay~A. Glang, Kyle~A. Long,
+  Francisco Malagon, Jan~J. Marayag, Edgar Nunez, Gregory~K. Rice, Ernesto
+  Santa~Ana, Megan~A. Schilling, Darci~R. Smith, Victor~A. Sugiharto, Peifang
+  Sun, Adriana van~de Guchte, Zenab Khan, Jayeeta Dutta, Sindhu Vangeti,
+  Logan~J. Voegtly, Dawn~L. Weir, C.~Jessica~E. Metcalf, Olga~G. Troyanskaya,
+  Kimberly~A. Bishop-Lilly, Bryan~T. Grenfell, Harm van Bakel, Andrew~G.
+  Letizia, and Stuart~C. Sealfon.
+\newblock {SARS-CoV-2} outbreak dynamics in an isolated {US} military recruit
+  training center with rigorous prevention measures.
+\newblock {\em Epidemiology}, 33(6), 2022.
+
+\bibitem{mizumoto2020estimating}
+Kenji Mizumoto, Katsushi Kagaya, Alexander Zarebski, and Gerardo Chowell.
+\newblock {Estimating the asymptomatic proportion of coronavirus disease 2019
+  (COVID-19) cases on board the Diamond Princess cruise ship, Yokohama, Japan,
+  2020}.
+\newblock {\em Eurosurveillance}, 25(10):2000180, 2020.
+
+\bibitem{ferretti2020quantifying}
+Luca Ferretti, Chris Wymant, Michelle Kendall, Lele Zhao, Anel Nurtay, Lucie
+  Abeler-D{\"o}rner, Michael Parker, David Bonsall, and Christophe Fraser.
+\newblock {Quantifying SARS-CoV-2 transmission suggests epidemic control with
+  digital contact tracing}.
+\newblock {\em Science}, 368(6491):eabb6936, 2020.
+
+\bibitem{lavezzo2020}
+Enrico Lavezzo, Elisa Franchin, Constanze Ciavarella, Gina Cuomo-Dannenburg,
+  Luisa Barzon, Claudia Del~Vecchio, Lucia Rossi, Riccardo Manganelli, Arianna
+  Loregian, Nicol{\`o} Navarin, Davide Abate, Manuela Sciro, Stefano
+  Merigliano, Ettore De~Canale, Maria~Cristina Vanuzzo, Valeria Besutti,
+  Francesca Saluzzo, Francesco Onelia, Monia Pacenti, Saverio~G. Parisi,
+  Giovanni Carretta, Daniele Donato, Luciano Flor, Silvia Cocchio, Giulia Masi,
+  Alessandro Sperduti, Lorenzo Cattarino, Renato Salvador, Michele Nicoletti,
+  Federico Caldart, Gioele Castelli, Eleonora Nieddu, Beatrice Labella,
+  Ludovico Fava, Matteo Drigo, Katy A.~M. Gaythorpe, Alessandra~R. Brazzale,
+  Stefano Toppo, Marta Trevisan, Vincenzo Baldo, Christl~A. Donnelly, Neil~M.
+  Ferguson, Ilaria Dorigatti, Andrea Crisanti, Kylie E.~C. Ainslie, Marc
+  Baguelin, Samir Bhatt, Adhiratha Boonyasiri, Olivia Boyd, Helen~L. Coupland,
+  Zulma Cucunub{\'a}, Bimandra~A. Djafaara, Christl~A. Donnelly, Sabine~L. van
+  Elsland, Rich FitzJohn, Seth Flaxman, Katy A.~M. Gaythorpe, Will~D. Green,
+  Timothy Hallett, Arran Hamlet, David Haw, Natsuko Imai, Benjamin Jeffrey,
+  Edward Knock, Daniel~J. Laydon, Thomas Mellan, Swapnil Mishra, Gemma
+  Nedjati-Gilani, Pierre Nouvellet, Lucy~C. Okell, Kris~V. Parag, Steven Riley,
+  Hayley~A. Thompson, H.~Juliette~T. Unwin, Robert Verity, Michaela A.~C.
+  Vollmer, Patrick G.~T. Walker, Caroline~E. Walters, Haowei Wang, Yuanrong
+  Wang, Oliver~J. Watson, Charles Whittaker, Lilith~K. Whittles, Xiaoyue Xi,
+  Neil~M. Ferguson, and {Imperial College COVID-19 Response Team}.
+\newblock {Suppression of a SARS-CoV-2 outbreak in the Italian municipality of
+  Vo'}.
+\newblock {\em Nature}, 584(7821):425--429, 2020.
+
+\bibitem{romero2021ct}
+Mar{\'\i}a~Pilar Romero-G{\'o}mez, Silvia G{\'o}mez-Sebastian, Emilio
+  Cendejas-Bueno, Mar{\'\i}a~Dolores Montero-Vega, Jes{\'u}s Mingorance, and
+  Julio Garc{\'\i}a-Rodr{\'\i}guez.
+\newblock Ct value is not enough to discriminate patients harbouring infective
+  virus.
+\newblock {\em Journal of Infection}, 82(3):e35--e37, 2021.
+
+\bibitem{stadnytskyi2021breathing}
+V~Stadnytskyi, P~Anfinrud, and A~Bax.
+\newblock {Breathing, speaking, coughing or sneezing: What drives transmission
+  of SARS-CoV-2?}
+\newblock {\em Journal of Internal Medicine}, 290(5):1010--1027, 2021.
+
+\bibitem{alene2021magnitude}
+Muluneh Alene, Leltework Yismaw, Moges~Agazhe Assemie, Daniel~Bekele Ketema,
+  Belayneh Mengist, Bekalu Kassie, and Tilahun~Yemanu Birhan.
+\newblock {Magnitude of asymptomatic COVID-19 cases throughout the course of
+  infection: A systematic review and meta-analysis}.
+\newblock {\em PloS one}, 16(3):e0249090, 2021.
+
+\bibitem{fraser2004factors}
+Christophe Fraser, Steven Riley, Roy~M Anderson, and Neil~M Ferguson.
+\newblock Factors that make an infectious disease outbreak controllable.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  101(16):6146--6151, 2004.
+
+\bibitem{moghadas2020implications}
+Seyed~M Moghadas, Meagan~C Fitzpatrick, Pratha Sah, Abhishek Pandey, Affan
+  Shoukat, Burton~H Singer, and Alison~P Galvani.
+\newblock {The implications of silent transmission for the control of COVID-19
+  outbreaks}.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  117(30):17513--17515, 2020.
+
+\bibitem{xie2022long}
+Yan Xie, Evan Xu, Benjamin Bowe, and Ziyad Al-Aly.
+\newblock {Long-term cardiovascular outcomes of COVID-19}.
+\newblock {\em Nature medicine}, 28(3):583--590, 2022.
+
+\bibitem{he2020temporal}
+Xi~He, Eric H.~Y. Lau, Peng Wu, Xilong Deng, Jian Wang, Xinxin Hao, Yiu~Chung
+  Lau, Jessica~Y. Wong, Yujuan Guan, Xinghua Tan, Xiaoneng Mo, Yanqing Chen,
+  Baolin Liao, Weilie Chen, Fengyu Hu, Qing Zhang, Mingqiu Zhong, Yanrong Wu,
+  Lingzhai Zhao, Fuchun Zhang, Benjamin~J. Cowling, Fang Li, and Gabriel~M.
+  Leung.
+\newblock {Temporal dynamics in viral shedding and transmissibility of
+  COVID-19}.
+\newblock {\em Nature Medicine}, 26(5):672--675, 2020.
+
+\bibitem{sender2021unmitigated}
+Ron Sender, Yinon~M Bar-On, Sang~Woo Park, Elad Noor, Jonathan Dushoffd, and
+  Ron Milo.
+\newblock {The unmitigated profile of COVID-19 infectiousness}.
+\newblock {\em medRxiv}, 2021.
+
+\bibitem{kucharski2020effectiveness}
+Adam~J Kucharski, Petra Klepac, Andrew~JK Conlan, Stephen~M Kissler, Maria~L
+  Tang, Hannah Fry, Julia~R Gog, W~John Edmunds, and {CMMID COVID-19 working
+  group}.
+\newblock Effectiveness of isolation, testing, contact tracing, and physical
+  distancing on reducing transmission of sars-cov-2 in different settings: a
+  mathematical modelling study.
+\newblock {\em The Lancet Infectious Diseases}, 20(10):1151--1160, 2020.
+
+\bibitem{lipsitch2020understanding}
+Marc Lipsitch and Natalie~E Dean.
+\newblock Understanding {COVID-19} vaccine efficacy.
+\newblock {\em Science}, 370(6518):763--765, 2020.
+
+\bibitem{andres2022omicron}
+Nick Andrews, Julia Stowe, Freja Kirsebom, Samuel Toffa, Tim Rickeard, Eileen
+  Gallagher, Charlotte Gower, Meaghan Kall, Natalie Groves, Anne-Marie
+  O’Connell, David Simons, Paula~B. Blomquist, Asad Zaidi, Sophie Nash, Nurin
+  Iwani Binti Abdul~Aziz, Simon Thelwall, Gavin Dabrera, Richard Myers, Gayatri
+  Amirthalingam, Saheer Gharbia, Jeffrey~C. Barrett, Richard Elson, Shamez~N.
+  Ladhani, Neil Ferguson, Maria Zambon, Colin~N.J. Campbell, Kevin Brown, Susan
+  Hopkins, Meera Chand, Mary Ramsay, and Jamie Lopez~Bernal.
+\newblock Covid-19 vaccine effectiveness against the {Omicron (B.1.1.529)}
+  variant.
+\newblock {\em New England Journal of Medicine}, 386(16):1532--1546, 2022.
+
+\bibitem{pearson2021omicron}
+Carl A.~B. Pearson, Sheetal~P. Silal, Michael~W.Z. Li, Jonathan Dushoff,
+  Benjamin~M. Bolker, Sam Abbott, Cari van Schalkwyk, Nicholas~G. Davies,
+  Rosanna~C. Barnard, W.~John Edmunds, Jeremy Bingham, Gesine Meyer-Rath, Lise
+  Jamieson, Allison Glass, Nicole Wolter, Nevashan Govender, Wendy~S. Stevens,
+  Lesley Scott, Koleka Mlisana, Harry Moultrie, and Juliet R.~C. Pulliam.
+\newblock {Bounding the levels of transmissibility \& immune evasion of the
+  Omicron variant in South Africa}.
+\newblock {\em medRxiv}, 2021.
+
+\bibitem{sun2022omicron}
+Kaiyuan Sun, Stefano Tempia, Jackie Kleynhans, Anne~von Gottberg, Meredith~L
+  McMorrow, Nicole Wolter, Jinal~N. Bhiman, Jocelyn Moyes, Mignon~du Plessis,
+  Maimuna Carrim, Amelia Buys, Neil~A Martinson, Kathleen Kahn, Stephen
+  Tollman, Limakatso Lebina, Floidy Wafawanaka, Jacques~D. du~Toit,
+  Francesc~Xavier G{\'o}mez-Oliv{\'e}, Thulisa Mkhencele, C{\'e}cile Viboud,
+  Cheryl Cohen, and the PHIRST~group.
+\newblock {Persistence of SARS-CoV-2 immunity, Omicron{\textquoteright}s
+  footprints, and projections of epidemic resurgences in South African
+  population cohorts}.
+\newblock {\em medRxiv}, 2022.
+
+\bibitem{MENNI20221618}
+Cristina Menni, Ana~M Valdes, Lorenzo Polidori, Michela Antonelli, Satya
+  Penamakuri, Ana Nogal, Panayiotis Louca, Anna May, Jane~C Figueiredo,
+  Christina Hu, Erika Molteni, Liane Canas, Marc~F Österdahl, Marc Modat,
+  Carole~H Sudre, Ben Fox, Alexander Hammers, Jonathan Wolf, Joan Capdevila,
+  Andrew~T Chan, Sean~P David, Claire~J Steves, Sebastien Ourselin, and Tim~D
+  Spector.
+\newblock {Symptom prevalence, duration, and risk of hospital admission in
+  individuals infected with SARS-CoV-2 during periods of omicron and delta
+  variant dominance: a prospective observational study from the ZOE COVID
+  Study}.
+\newblock {\em The Lancet}, 399(10335):1618--1624, 2022.
+
+\bibitem{ulloa2022estimates}
+Ana~Cecilia Ulloa, Sarah~A Buchan, Nick Daneman, and Kevin~A Brown.
+\newblock {Estimates of SARS-CoV-2 omicron variant severity in Ontario,
+  Canada}.
+\newblock {\em Jama}, 327(13):1286--1288, 2022.
+
+\bibitem{Iacobuccio254}
+Gareth Iacobucci.
+\newblock Covid-19: unravelling the conundrum of {Omicron} and deaths.
+\newblock {\em BMJ}, 376, 2022.
+
+\bibitem{faust2022omicron}
+Jeremy~Samuel Faust, Chengan Du, Chenxue Liang, Katherine~Dickerson Mayes,
+  Benjamin Renton, Kristen Panthagani, and Harlan~M. Krumholz.
+\newblock {Excess Mortality in {Massachusetts} During the {Delta} and {Omicron}
+  Waves of {COVID-19}}.
+\newblock {\em JAMA}, 05 2022.
+
+\bibitem{sigal2022estimating}
+Alex Sigal, Ron Milo, and Waasila Jassat.
+\newblock {Estimating disease severity of Omicron and Delta SARS-CoV-2
+  infections}.
+\newblock {\em Nature Reviews Immunology}, pages 1--3, 2022.
+
+\bibitem{saad2020immune}
+Chadi~M Saad-Roy, Caroline~E Wagner, Rachel~E Baker, Sinead~E Morris, Jeremy
+  Farrar, Andrea~L Graham, Simon~A Levin, Michael~J Mina, C~Jessica~E Metcalf,
+  and Bryan~T Grenfell.
+\newblock {Immune life history, vaccination, and the dynamics of SARS-CoV-2
+  over the next 5 years}.
+\newblock {\em Science}, 370(6518):811--818, 2020.
+
+\bibitem{gallagher2021indirect}
+Molly~E Gallagher, Andrew~J Sieben, Kristin~N Nelson, Alicia~NM Kraay, Walter~A
+  Orenstein, Ben Lopman, Andreas Handel, and Katia Koelle.
+\newblock {Indirect benefits are a crucial consideration when evaluating
+  SARS-CoV-2 vaccine candidates}.
+\newblock {\em Nature medicine}, 27(1):4--5, 2021.
+
+\bibitem{park2020time}
+Sang~Woo Park, Daniel~M Cornforth, Jonathan Dushoff, and Joshua~S Weitz.
+\newblock {The time scale of asymptomatic transmission affects estimates of
+  epidemic potential in the COVID-19 outbreak}.
+\newblock {\em Epidemics}, 31:100392, 2020.
+
+\bibitem{harris2023time}
+Jeremy~D Harris, Sang~Woo Park, Jonathan Dushoff, and Joshua~S Weitz.
+\newblock How time-scale differences in asymptomatic and symptomatic
+  transmission shape sars-cov-2 outbreak dynamics.
+\newblock {\em Epidemics}, page 100664, 2023.
+
+\bibitem{moore2021modelling}
+Sam Moore, Edward~M Hill, Louise Dyson, Michael~J Tildesley, and Matt~J
+  Keeling.
+\newblock {Modelling optimal vaccination strategy for SARS-CoV-2 in the UK}.
+\newblock {\em PLoS computational biology}, 17(5):e1008849, 2021.
+
+\bibitem{weitz2020awareness}
+Joshua~S Weitz, Sang~Woo Park, Ceyhun Eksin, and Jonathan Dushoff.
+\newblock Awareness-driven behavior changes can shift the shape of epidemics
+  away from peaks and toward plateaus, shoulders, and oscillations.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  117(51):32764--32771, 2020.
+
+\bibitem{kissler2020projecting}
+Stephen~M Kissler, Christine Tedijanto, Edward Goldstein, Yonatan~H Grad, and
+  Marc Lipsitch.
+\newblock {Projecting the transmission dynamics of SARS-CoV-2 through the
+  postpandemic period}.
+\newblock {\em Science}, 368(6493):860--868, 2020.
+
+\bibitem{lavine2021immunological}
+Jennie~S Lavine, Ottar~N Bjornstad, and Rustom Antia.
+\newblock {Immunological characteristics govern the transition of COVID-19 to
+  endemicity}.
+\newblock {\em Science}, 371(6530):741--745, 2021.
+
+\bibitem{gandon2001imperfect}
+Sylvain Gandon, Margaret~J Mackinnon, Sean Nee, and Andrew~F Read.
+\newblock Imperfect vaccines and the evolution of pathogen virulence.
+\newblock {\em Nature}, 414(6865):751--756, 2001.
+
+\bibitem{Bavel2020}
+Jay J.~Van Bavel, Katherine Baicker, Paulo~S. Boggio, Valerio Capraro,
+  Aleksandra Cichocka, Mina Cikara, Molly~J. Crockett, Alia~J. Crum, Karen~M.
+  Douglas, James~N. Druckman, John Drury, Oeindrila Dube, Naomi Ellemers,
+  Eli~J. Finkel, James~H. Fowler, Michele Gelfand, Shihui Han, S.~Alexander
+  Haslam, Jolanda Jetten, Shinobu Kitayama, Dean Mobbs, Lucy~E. Napper,
+  Dominic~J. Packer, Gordon Pennycook, Ellen Peters, Richard~E. Petty, David~G.
+  Rand, Stephen~D. Reicher, Simone Schnall, Azim Shariff, Linda~J. Skitka,
+  Sandra~Susan Smith, Cass~R. Sunstein, Nassim Tabri, Joshua~A. Tucker, Sander
+  van~der Linden, Paul~van Lange, Kim~A. Weeden, Michael J.~A. Wohl, Jamil
+  Zaki, Sean~R. Zion, and Robb Willer.
+\newblock {Using social and behavioural science to support COVID-19 pandemic
+  response}.
+\newblock {\em Nature Human Behaviour}, 4(5):460--471, 2020.
+
+\bibitem{chande2020real}
+Aroon Chande, Seolha Lee, Mallory Harris, Quan Nguyen, Stephen~J Beckett, Troy
+  Hilley, Clio Andris, and Joshua~S Weitz.
+\newblock {Real-time, interactive website for US-county-level COVID-19 event
+  risk assessment}.
+\newblock {\em Nature Human Behaviour}, 4(12):1313--1319, 2020.
+
+\bibitem{sinclair2021pairing}
+Alyssa~H Sinclair, Shabnam Hakimi, Matthew~L Stanley, R~Alison Adcock, and
+  Gregory~R Samanez-Larkin.
+\newblock Pairing facts with imagined consequences improves pandemic-related
+  risk perception.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  118(32):e2100970118, 2021.
+
+\bibitem{mina2021covid}
+Michael~J Mina and Kristian~G Andersen.
+\newblock {COVID-19} testing: One size does not fit all.
+\newblock {\em Science}, 371(6525):126--127, 2021.
+
+\bibitem{gibson2022surv}
+Greg Gibson, Joshua~S. Weitz, Michael~P. Shannon, Benjamin Holton, Anton
+  Bryksin, Brian Liu, Madeline Sieglinger, Ashley~R. Coenen, Conan Zhao,
+  Stephen~J. Beckett, Sandra Bramblett, JulieAnne Williamson, Michael Farrell,
+  Alexander Ortiz, Chaouki~T. Abdallah, and Andr{\'e}s~J. Garc{\'\i}a.
+\newblock Surveillance-to-diagnostic testing program for asymptomatic
+  {SARS-CoV-2} infections on a large, urban campus in fall 2020.
+\newblock {\em Epidemiology}, 33(2), 2022.
+
+\bibitem{Ranoa2022}
+Diana Rose~E. Ranoa, Robin~L. Holland, Fadi~G. Alnaji, Kelsie~J. Green, Leyi
+  Wang, Richard~L. Fredrickson, Tong Wang, George~N. Wong, Johnny Uelmen,
+  Sergei Maslov, Zachary~J. Weiner, Alexei~V. Tkachenko, Hantao Zhang, Zhiru
+  Liu, Ahmed Ibrahim, Sanjay~J. Patel, John~M. Paul, Nickolas~P. Vance,
+  Joseph~G. Gulick, Sandeep~Puthanveetil Satheesan, Isaac~J. Galvan, Andrew
+  Miller, Joseph Grohens, Todd~J. Nelson, Mary~P. Stevens, P.~Mark Hennessy,
+  Robert~C. Parker, Edward Santos, Charles Brackett, Julie~D. Steinman,
+  Melvin~R. Fenner, Kirstin Dohrer, Michael DeLorenzo, Laura Wilhelm-Barr,
+  Brian~R. Brauer, Catherine Best-Popescu, Gary Durack, Nathan Wetter, David~M.
+  Kranz, Jessica Breitbarth, Charlie Simpson, Julie~A. Pryde, Robin~N. Kaler,
+  Chris Harris, Allison~C. Vance, Jodi~L. Silotto, Mark Johnson, Enrique~Andres
+  Valera, Patricia~K. Anton, Lowa Mwilambwe, Stephen~P. Bryan, Deborah~S.
+  Stone, Danita~B. Young, Wanda~E. Ward, John Lantz, John~A. Vozenilek, Rashid
+  Bashir, Jeffrey~S. Moore, Mayank Garg, Julian~C. Cooper, Gillian Snyder,
+  Michelle~H. Lore, Dustin~L. Yocum, Neal~J. Cohen, Jan~E. Novakofski,
+  Melanie~J. Loots, Randy~L. Ballard, Mark Band, Kayla~M. Banks, Joseph~D.
+  Barnes, Iuliana Bentea, Jessica Black, Jeremy Busch, Abigail Conte, Madison
+  Conte, Michael Curry, Jennifer Eardley, April Edwards, Therese Eggett, Judes
+  Fleurimont, Delaney Foster, Bruce~W. Fouke, Nicholas Gallagher, Nicole
+  Gastala, Scott~A. Genung, Declan Glueck, Brittani Gray, Andrew Greta,
+  Robert~M. Healy, Ashley Hetrick, Arianna~A. Holterman, Nahed Ismail, Ian
+  Jasenof, Patrick Kelly, Aaron Kielbasa, Teresa Kiesel, Lorenzo~M. Kindle,
+  Rhonda~L. Lipking, Yukari~C. Manabe, Jade Mayes, Reubin McGuffin, Kenton~G.
+  McHenry, Agha Mirza, Jada Moseley, Heba~H. Mostafa, Melody Mumford, Kathleen
+  Munoz, Arika~D. Murray, Moira Nolan, Nil~A. Parikh, Andrew Pekosz, Janna
+  Pflugmacher, Janise~M. Phillips, Collin Pitts, Mark~C. Potter, James
+  Quisenberry, Janelle Rear, Matthew~L. Robinson, Edith Rosillo, Leslie~N. Rye,
+  MaryEllen Sherwood, Anna Simon, Jamie~M. Singson, Carly Skadden, Tina~H.
+  Skelton, Charlie Smith, Mary Stech, Ryan Thomas, Matthew~A. Tomaszewski,
+  Erika~A. Tyburski, Scott Vanwingerden, Evette Vlach, Ronald~S. Watkins,
+  Karriem Watson, Karen~C. White, Timothy~L. Killeen, Robert~J. Jones,
+  Andreas~C. Cangellaris, Susan~A. Martinis, Awais Vaid, Christopher~B. Brooke,
+  Joseph~T. Walsh, Ahmed Elbanna, William~C. Sullivan, Rebecca~L. Smith, Nigel
+  Goldenfeld, Timothy~M. Fan, Paul~J. Hergenrother, and Martin~D. Burke.
+\newblock {Mitigation of SARS-CoV-2 transmission at a large public university}.
+\newblock {\em Nature Communications}, 13(1):3207, 2022.
+
+\bibitem{jones2020two}
+Nicholas~R Jones, Zeshan~U Qureshi, Robert~J Temple, Jessica~PJ Larwood, Trisha
+  Greenhalgh, and Lydia Bourouiba.
+\newblock Two metres or one: what is the evidence for physical distancing in
+  {COVID-19?}
+\newblock {\em bmj}, 370, 2020.
+
+\bibitem{prather2020reducing}
+Kimberly~A Prather, Chia~C Wang, and Robert~T Schooley.
+\newblock Reducing transmission of {SARS-CoV-2}.
+\newblock {\em Science}, 368(6498):1422--1424, 2020.
+
+\bibitem{howard2021ev}
+Jeremy Howard, Austin Huang, Zhiyuan Li, Zeynep Tufekci, Vladimir Zdimal,
+  Helene-Mari van~der Westhuizen, Arne von Delft, Amy Price, Lex Fridman,
+  Lei-Han Tang, Viola Tang, Gregory~L. Watson, Christina~E. Bax, Reshama
+  Shaikh, Frederik Questier, Danny Hernandez, Larry~F. Chu, Christina~M.
+  Ramirez, and Anne~W. Rimoin.
+\newblock {An evidence review of face masks against COVID-19}.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  118(4):e2014564118, 2021.
+
+\bibitem{allen2021indoor}
+Joseph~G Allen and Andrew~M Ibrahim.
+\newblock {Indoor air changes and potential implications for SARS-CoV-2
+  transmission}.
+\newblock {\em Jama}, 325(20):2112--2113, 2021.
+
+\bibitem{wang2021airborne}
+Chia~C Wang, Kimberly~A Prather, Josu{\'e} Sznitman, Jose~L Jimenez, Seema~S
+  Lakdawala, Zeynep Tufekci, and Linsey~C Marr.
+\newblock Airborne transmission of respiratory viruses.
+\newblock {\em Science}, 373(6558):eabd9149, 2021.
+
+\bibitem{lopman2021framework}
+Benjamin~A Lopman, Kayoko Shioda, Quan Nguyen, Stephen~J Beckett, Aaron~J
+  Siegler, Patrick~S Sullivan, and Joshua~S Weitz.
+\newblock {A framework for monitoring population immunity to SARS-CoV-2}.
+\newblock {\em Annals of Epidemiology}, 63:75--78, 2021.
+
+\bibitem{fox2020impact}
+Spencer~J Fox, Remy Pasco, Mauricio Tec, Zhanwei Du, Michael Lachmann, James
+  Scott, and Lauren~Ancel Meyers.
+\newblock {The impact of asymptomatic COVID-19 infections on future pandemic
+  waves}.
+\newblock {\em MedRxiv}, pages 2020--06, 2020.
+
+\bibitem{park2020reconciling}
+Sang~Woo Park, Benjamin~M Bolker, David Champredon, David~JD Earn, Michael Li,
+  Joshua~S Weitz, Bryan~T Grenfell, and Jonathan Dushoff.
+\newblock {Reconciling early-outbreak estimates of the basic reproductive
+  number and its uncertainty: framework and applications to the novel
+  coronavirus (SARS-CoV-2) outbreak}.
+\newblock {\em Journal of the Royal Society Interface}, 17(168):20200144, 2020.
+
+\bibitem{smith1984assessment}
+PG~Smith, LC~Rodrigues, and PEM Fine.
+\newblock Assessment of the protective efficacy of vaccines against common
+  diseases using case-control and cohort studies.
+\newblock {\em International journal of epidemiology}, 13(1):87--93, 1984.
+
+\bibitem{buckner2021dynamic}
+Jack~H Buckner, Gerardo Chowell, and Michael~R Springborn.
+\newblock {Dynamic prioritization of COVID-19 vaccines when social distancing
+  is limited for essential workers}.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  118(16):e2025786118, 2021.
+
+\bibitem{bubar2021model}
+Kate~M Bubar, Kyle Reinholt, Stephen~M Kissler, Marc Lipsitch, Sarah Cobey,
+  Yonatan~H Grad, and Daniel~B Larremore.
+\newblock {Model-informed COVID-19 vaccine prioritization strategies by age and
+  serostatus}.
+\newblock {\em Science}, 371(6532):916--921, 2021.
+
+\end{thebibliography}
 

### 2023-03-21 main.tex
minor edit

@@
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript. 
+This manuscript was posted on a preprint: \url{https://www.medrxiv.org/content/10.1101/2022.08.01.22278288v1.full.pdf}.
 

### 2023-03-21 final_files/main.tex
formatting

--- 2023-03-21 main.tex
@@
 \usepackage{natbib}
-\bibliographystyle{unsrt}
+\bibliographystyle{pnas}
 \date{\today}
@@
 
+\section*{Author contribution}
+
+SWP, JD, and JSW designed the project with input from BTG.
+SWP and JSW performed model analysis.
+SWP, JD,  and JSW analyzed model results.
+SWP performed data analysis with input from all authors.
+SWP, JD, BTG, and JSW wrote the paper.
+
 \pagebreak
 
-\bibliography{main}
+\begin{thebibliography}{10}
+
+\bibitem{nogrady2020data}
+B Nogrady, What the data say about asymptomatic {COVID} infections.
+\newblock {\em {Nature}} \textbf{587}, 534--536 (2020).
+
+\bibitem{rajgor2020many}
+DD Rajgor, MH Lee, S Archuleta, N Bagdasarian, SC Quek, {The many estimates of
+  the COVID-19 case fatality rate}.
+\newblock {\em {The Lancet Infectious Diseases}}
+  \textbf{20}, 776--777 (2020).
+
+\bibitem{VERITY2020669}
+R Verity, et~al., Estimates of the severity of coronavirus disease 2019: a
+  model-based analysis.
+\newblock {\em {The Lancet Infectious Diseases}}
+  \textbf{20}, 669--677 (2020).
+
+\bibitem{yang2020early}
+S Yang, et~al., {Early estimation of the case fatality rate of COVID-19 in
+  mainland China: a data-driven analysis}.
+\newblock {\em {Annals of translational medicine}}
+  \textbf{8} (2020).
+
+\bibitem{levin2020assessing}
+AT Levin, et~al., {Assessing the age specificity of infection fatality rates
+  for COVID-19: systematic review, meta-analysis, and public policy
+  implications}.
+\newblock {\em {European journal of epidemiology}}
+  \textbf{35}, 1123--1138 (2020).
+
+\bibitem{emery2020}
+JC Emery, et~al., {The contribution of asymptomatic SARS-CoV-2 infections to
+  transmission on the Diamond Princess cruise ship}.
+\newblock {\em {eLife}} \textbf{9}, e58699 (2020).
+
+\bibitem{ggplot2}
+H Wickham, {\em {ggplot2: Elegant Graphics for Data Analysis}}.
+\newblock (Springer-Verlag New York), (2016).
+
+\bibitem{Kissler2020}
+SM Kissler, et~al., {Viral dynamics of acute SARS-CoV-2 infection}.
+\newblock {\em {medRxiv}} (2021).
+
+\bibitem{davies2020}
+NG Davies, et~al., {Age-dependent effects in the transmission and control of
+  COVID-19 epidemics}.
+\newblock {\em {Nature Medicine}} \textbf{26}, 1205--1211
+  (2020).
+
+\bibitem{gao2021role}
+W Gao, J Lv, Y Pang, LM Li, {Role of asymptomatic and pre-symptomatic
+  infections in COVID-19 pandemic}.
+\newblock {\em {bmj}} \textbf{375} (2021).
+
+\bibitem{johansson2021sars}
+MA Johansson, et~al., {SARS-CoV-2 transmission from people without COVID-19
+  symptoms}.
+\newblock {\em {JAMA network open}} \textbf{4},
+  e2035057--e2035057 (2021).
+
+\bibitem{subramanian2021quantifying}
+R Subramanian, Q He, M Pascual, {Quantifying asymptomatic infection and
+  transmission of COVID-19 in New York City using observed cases, serology, and
+  testing capacity}.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{118}, e2019716118 (2021).
+
+\bibitem{koelle2022changing}
+K Koelle, MA Martin, R Antia, B Lopman, NE Dean, {The changing epidemiology of
+  SARS-CoV-2}.
+\newblock {\em {Science}} \textbf{375}, 1116--1121 (2022).
+
+\bibitem{lizewski2022navy}
+RA Lizewski, et~al., {SARS-CoV-2} outbreak dynamics in an isolated {US}
+  military recruit training center with rigorous prevention measures.
+\newblock {\em {Epidemiology}} \textbf{33} (2022).
+
+\bibitem{mizumoto2020estimating}
+K Mizumoto, K Kagaya, A Zarebski, G Chowell, {Estimating the asymptomatic
+  proportion of coronavirus disease 2019 (COVID-19) cases on board the Diamond
+  Princess cruise ship, Yokohama, Japan, 2020}.
+\newblock {\em {Eurosurveillance}} \textbf{25}, 2000180
+  (2020).
+
+\bibitem{ferretti2020quantifying}
+L Ferretti, et~al., {Quantifying SARS-CoV-2 transmission suggests epidemic
+  control with digital contact tracing}.
+\newblock {\em {Science}} \textbf{368}, eabb6936 (2020).
+
+\bibitem{lavezzo2020}
+E Lavezzo, et~al., {Suppression of a SARS-CoV-2 outbreak in the Italian
+  municipality of Vo'}.
+\newblock {\em {Nature}} \textbf{584}, 425--429 (2020).
+
+\bibitem{romero2021ct}
+MP Romero-G{\'o}mez, et~al., Ct value is not enough to discriminate patients
+  harbouring infective virus.
+\newblock {\em {Journal of Infection}} \textbf{82},
+  e35--e37 (2021).
+
+\bibitem{stadnytskyi2021breathing}
+V Stadnytskyi, P Anfinrud, A Bax, {Breathing, speaking, coughing or sneezing:
+  What drives transmission of SARS-CoV-2?}
+\newblock {\em {Journal of Internal Medicine}}
+  \textbf{290}, 1010--1027 (2021).
+
+\bibitem{alene2021magnitude}
+M Alene, et~al., {Magnitude of asymptomatic COVID-19 cases throughout the
+  course of infection: A systematic review and meta-analysis}.
+\newblock {\em {PloS one}} \textbf{16}, e0249090 (2021).
+
+\bibitem{fraser2004factors}
+C Fraser, S Riley, RM Anderson, NM Ferguson, Factors that make an infectious
+  disease outbreak controllable.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{101}, 6146--6151 (2004).
+
+\bibitem{moghadas2020implications}
+SM Moghadas, et~al., {The implications of silent transmission for the control
+  of COVID-19 outbreaks}.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{117}, 17513--17515 (2020).
+
+\bibitem{xie2022long}
+Y Xie, E Xu, B Bowe, Z Al-Aly, {Long-term cardiovascular outcomes of COVID-19}.
+\newblock {\em {Nature medicine}} \textbf{28}, 583--590
+  (2022).
+
+\bibitem{he2020temporal}
+X He, et~al., {Temporal dynamics in viral shedding and transmissibility of
+  COVID-19}.
+\newblock {\em {Nature Medicine}} \textbf{26}, 672--675
+  (2020).
+
+\bibitem{sender2021unmitigated}
+R Sender, et~al., {The unmitigated profile of COVID-19 infectiousness}.
+\newblock {\em {medRxiv}} (2021).
+
+\bibitem{kucharski2020effectiveness}
+AJ Kucharski, et~al., Effectiveness of isolation, testing, contact tracing, and
+  physical distancing on reducing transmission of sars-cov-2 in different
+  settings: a mathematical modelling study.
+\newblock {\em {The Lancet Infectious Diseases}}
+  \textbf{20}, 1151--1160 (2020).
+
+\bibitem{lipsitch2020understanding}
+M Lipsitch, NE Dean, Understanding {COVID-19} vaccine efficacy.
+\newblock {\em {Science}} \textbf{370}, 763--765 (2020).
+
+\bibitem{andres2022omicron}
+N Andrews, et~al., Covid-19 vaccine effectiveness against the {Omicron
+  (B.1.1.529)} variant.
+\newblock {\em {New England Journal of Medicine}}
+  \textbf{386}, 1532--1546 (2022).
+
+\bibitem{pearson2021omicron}
+CAB Pearson, et~al., {Bounding the levels of transmissibility \& immune evasion
+  of the Omicron variant in South Africa}.
+\newblock {\em {medRxiv}} (2021).
+
+\bibitem{sun2022omicron}
+K Sun, et~al., {Persistence of SARS-CoV-2 immunity, Omicron{\textquoteright}s
+  footprints, and projections of epidemic resurgences in South African
+  population cohorts}.
+\newblock {\em {medRxiv}} (2022).
+
+\bibitem{MENNI20221618}
+C Menni, et~al., {Symptom prevalence, duration, and risk of hospital admission
+  in individuals infected with SARS-CoV-2 during periods of omicron and delta
+  variant dominance: a prospective observational study from the ZOE COVID
+  Study}.
+\newblock {\em {The Lancet}} \textbf{399}, 1618--1624
+  (2022).
+
+\bibitem{ulloa2022estimates}
+AC Ulloa, SA Buchan, N Daneman, KA Brown, {Estimates of SARS-CoV-2 omicron
+  variant severity in Ontario, Canada}.
+\newblock {\em {Jama}} \textbf{327}, 1286--1288 (2022).
+
+\bibitem{Iacobuccio254}
+G Iacobucci, Covid-19: unravelling the conundrum of {Omicron} and deaths.
+\newblock {\em {BMJ}} \textbf{376} (2022).
+
+\bibitem{faust2022omicron}
+JS Faust, et~al., {Excess Mortality in {Massachusetts} During the {Delta} and
+  {Omicron} Waves of {COVID-19}}.
+\newblock {\em {JAMA}} (2022).
+
+\bibitem{sigal2022estimating}
+A Sigal, R Milo, W Jassat, {Estimating disease severity of Omicron and Delta
+  SARS-CoV-2 infections}.
+\newblock {\em {Nature Reviews Immunology}} pp. 1--3
+  (2022).
+
+\bibitem{saad2020immune}
+CM Saad-Roy, et~al., {Immune life history, vaccination, and the dynamics of
+  SARS-CoV-2 over the next 5 years}.
+\newblock {\em {Science}} \textbf{370}, 811--818 (2020).
+
+\bibitem{gallagher2021indirect}
+ME Gallagher, et~al., {Indirect benefits are a crucial consideration when
+  evaluating SARS-CoV-2 vaccine candidates}.
+\newblock {\em {Nature medicine}} \textbf{27}, 4--5 (2021).
+
+\bibitem{park2020time}
+SW Park, DM Cornforth, J Dushoff, JS Weitz, {The time scale of asymptomatic
+  transmission affects estimates of epidemic potential in the COVID-19
+  outbreak}.
+\newblock {\em {Epidemics}} \textbf{31}, 100392 (2020).
+
+\bibitem{harris2023time}
+JD Harris, SW Park, J Dushoff, JS Weitz, How time-scale differences in
+  asymptomatic and symptomatic transmission shape sars-cov-2 outbreak dynamics.
+\newblock {\em {Epidemics}} p. 100664 (2023).
+
+\bibitem{moore2021modelling}
+S Moore, EM Hill, L Dyson, MJ Tildesley, MJ Keeling, {Modelling optimal
+  vaccination strategy for SARS-CoV-2 in the UK}.
+\newblock {\em {PLoS computational biology}} \textbf{17},
+  e1008849 (2021).
+
+\bibitem{weitz2020awareness}
+JS Weitz, SW Park, C Eksin, J Dushoff, Awareness-driven behavior changes can
+  shift the shape of epidemics away from peaks and toward plateaus, shoulders,
+  and oscillations.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{117}, 32764--32771 (2020).
+
+\bibitem{kissler2020projecting}
+SM Kissler, C Tedijanto, E Goldstein, YH Grad, M Lipsitch, {Projecting the
+  transmission dynamics of SARS-CoV-2 through the postpandemic period}.
+\newblock {\em {Science}} \textbf{368}, 860--868 (2020).
+
+\bibitem{lavine2021immunological}
+JS Lavine, ON Bjornstad, R Antia, {Immunological characteristics govern the
+  transition of COVID-19 to endemicity}.
+\newblock {\em {Science}} \textbf{371}, 741--745 (2021).
+
+\bibitem{gandon2001imperfect}
+S Gandon, MJ Mackinnon, S Nee, AF Read, Imperfect vaccines and the evolution of
+  pathogen virulence.
+\newblock {\em {Nature}} \textbf{414}, 751--756 (2001).
+
+\bibitem{Bavel2020}
+JJV Bavel, et~al., {Using social and behavioural science to support COVID-19
+  pandemic response}.
+\newblock {\em {Nature Human Behaviour}} \textbf{4},
+  460--471 (2020).
+
+\bibitem{chande2020real}
+A Chande, et~al., {Real-time, interactive website for US-county-level COVID-19
+  event risk assessment}.
+\newblock {\em {Nature Human Behaviour}} \textbf{4},
+  1313--1319 (2020).
+
+\bibitem{sinclair2021pairing}
+AH Sinclair, S Hakimi, ML Stanley, RA Adcock, GR Samanez-Larkin, Pairing facts
+  with imagined consequences improves pandemic-related risk perception.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{118}, e2100970118 (2021).
+
+\bibitem{mina2021covid}
+MJ Mina, KG Andersen, {COVID-19} testing: One size does not fit all.
+\newblock {\em {Science}} \textbf{371}, 126--127 (2021).
+
+\bibitem{gibson2022surv}
+G Gibson, et~al., Surveillance-to-diagnostic testing program for asymptomatic
+  {SARS-CoV-2} infections on a large, urban campus in fall 2020.
+\newblock {\em {Epidemiology}} \textbf{33} (2022).
+
+\bibitem{Ranoa2022}
+DRE Ranoa, et~al., {Mitigation of SARS-CoV-2 transmission at a large public
+  university}.
+\newblock {\em {Nature Communications}} \textbf{13}, 3207
+  (2022).
+
+\bibitem{jones2020two}
+NR Jones, et~al., Two metres or one: what is the evidence for physical
+  distancing in {COVID-19?}
+\newblock {\em {bmj}} \textbf{370} (2020).
+
+\bibitem{prather2020reducing}
+KA Prather, CC Wang, RT Schooley, Reducing transmission of {SARS-CoV-2}.
+\newblock {\em {Science}} \textbf{368}, 1422--1424 (2020).
+
+\bibitem{howard2021ev}
+J Howard, et~al., {An evidence review of face masks against COVID-19}.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{118}, e2014564118 (2021).
+
+\bibitem{allen2021indoor}
+JG Allen, AM Ibrahim, {Indoor air changes and potential implications for
+  SARS-CoV-2 transmission}.
+\newblock {\em {Jama}} \textbf{325}, 2112--2113 (2021).
+
+\bibitem{wang2021airborne}
+CC Wang, et~al., Airborne transmission of respiratory viruses.
+\newblock {\em {Science}} \textbf{373}, eabd9149 (2021).
+
+\bibitem{lopman2021framework}
+BA Lopman, et~al., {A framework for monitoring population immunity to
+  SARS-CoV-2}.
+\newblock {\em {Annals of Epidemiology}} \textbf{63},
+  75--78 (2021).
+
+\bibitem{fox2020impact}
+SJ Fox, et~al., {The impact of asymptomatic COVID-19 infections on future
+  pandemic waves}.
+\newblock {\em {MedRxiv}} pp. 2020--06 (2020).
+
+\bibitem{park2020reconciling}
+SW Park, et~al., {Reconciling early-outbreak estimates of the basic
+  reproductive number and its uncertainty: framework and applications to the
+  novel coronavirus (SARS-CoV-2) outbreak}.
+\newblock {\em {Journal of the Royal Society Interface}}
+  \textbf{17}, 20200144 (2020).
+
+\bibitem{smith1984assessment}
+P Smith, L Rodrigues, P Fine, Assessment of the protective efficacy of vaccines
+  against common diseases using case-control and cohort studies.
+\newblock {\em {International journal of epidemiology}}
+  \textbf{13}, 87--93 (1984).
+
+\bibitem{buckner2021dynamic}
+JH Buckner, G Chowell, MR Springborn, {Dynamic prioritization of COVID-19
+  vaccines when social distancing is limited for essential workers}.
+\newblock {\em {Proceedings of the National Academy of
+  Sciences}} \textbf{118}, e2025786118 (2021).
+
+\bibitem{bubar2021model}
+KM Bubar, et~al., {Model-informed COVID-19 vaccine prioritization strategies by
+  age and serostatus}.
+\newblock {\em {Science}} \textbf{371}, 916--921 (2021).
+
+\end{thebibliography}
 
